# Automating Analyses {#automate}

## Questions {#automate-questions}

-   How can I make my analyses easier to reproduce?
-   How can I make it easier to repeat analyses when I get new data, or when my data or scripts change?

## Introduction {#automate-intro}

Zipf's Law states that the second most common word in a body of text
appears half as often as the most common,
the third most common appears a third as often,
and so on.
To test it,
we will use some novels from [Project Gutenberg][gutenberg]:

| Book                            | Length (words) |
| ------------------------------- | -------------: |
| anne_of_green_gables.txt        |         105642 |
| common_sense.txt                |          24999 |
| count_of_monte_cristo.txt       |         464226 |
| dracula.txt                     |         164424 |
| emma.txt                        |         160458 |
| ethan_frome.txt                 |          37732 |
| frankenstein.txt                |          78098 |
| jane_eyre.txt                   |         188455 |
| life_of_frederick_douglass.txt  |          43789 |
| moby_dick.txt                   |         215830 |
| mysterious_affair_at_styles.txt |          59604 |
| pride_and_prejudice.txt         |         124974 |
| sense_and_sensibility.txt       |         121590 |
| sherlock_holmes.txt             |         107533 |
| time_machine.txt                |          35524 |
| treasure_island.txt             |          71616 |

The most common words in this [corpus][corpus] appear this many times:

| Word | Count |
| ---- | ----: |
| the  | 97278 |
| and  | 59385 |
| to   | 56028 |
| of   | 55190 |
| I    | 45680 |
| a    | 40483 |
| in   | 30030 |
| was  | 24512 |
| that | 24386 |
| you  | 22123 |
| it   | 21420 |

The frequencies don't match Zipf's predictions exactly—for example,
we would expect about 48,600 occurrences of "and"—but
there certainly seems to be a decay curve of some kind.
We'll look more closely at this data as we go along.

The project we have inherited as a starting point contains the following:

1.  The books we are analyzing are in <code>data/<em>title</em>.txt</code>.
2.  A program called `bin/countwords.py` reads a text file
    and produces a [CSV][csv] file with two columns:
    the words in the text
    and the number of times that word occurs.
3.  `bin/countwords.py` can analyze several files at once if we provide many filenames on the command line
    or concatenate the files and send them to standard input in a pipeline
    using something like `cat data/*.txt | bin/countwords.py`.
4.  Another program, `bin/plotcounts.py`, will create a visualization for us
    that shows word rank on the X axis and word counts on the Y axis.
    (It doesn't show the actual words.)
5.  A third program, `bin/collate.py`,
    takes one or more CSV files as input and merges them
    by combining the counts for words they have in common.
6.  Finally,
    `bin/testfit.py` compares actual distributions against expectations
    and give a fitting score.

It's easy enough to run these programs by hand if we only want to analyze a handful of files,
but doing this becomes tedious and error-prone as the number of files grows.
Instead,
we can write a program to do multi-file analyses.
Doing this documents the pipeline so that our colleagues (and future selves) can understand it,
and enables us to re-do the entire analysis with a single command if we get new data
or change our methods, parameters, or libraries.
It also prevents us from making lots of little errors:
there's no guarantee we'll get the script right the first time,
but once we've fixed it, it will stay fixed.

However,
re-running the entire analysis every time we get a new file is inefficient:
we don't need to re-count the words in our first thousand books
when we add the thousand and first.
This isn't a problem when calculations can be done quickly,
but many can't,
and anyway,
the point of this chapter is to introduce a tool for handling cases in which
we really want to avoid doing unnecessary work.

What we want is a way to describe which files depend on which other files
and how to generate or update a file when necessary.
This is the job of a [build tool][build-tool].
As the name suggests,
a build tool's job is to build new files out of old ones.
The most widely used build tool,
[Make][make],
was written in 1976 to recompile programs
(which at the time was a slow process).
[GNU Make][gnu-make] is a free, fast, and well-documented version of Make;
we will use it throughout this book.

> **Alternatives**
>
> [Many better build tools][build-tool] have been developed since Make---so many,
> in fact,
> that none has been able to capture more than a small fraction of potential users.
> [Snakemake][snakemake] has a lot of fans,
> and a future version of this tutorial might well use it.

This introduction based on
the [Software Carpentry lesson on Make][swc-make] maintained by [Gerard Capes][capes-gerard]
and on [Jonathan Dursi][dursi-jonathan]'s [introduction to pattern rules][dursi-pattern-rules].

## How can I update a file when its prerequisites change? {#automate-first}

Make is based on three key ideas:

1.  The operating system automatically records a [timestamp](glossary.html#timestamp)
    every time a file is changed.
    By checking this,
    Make can tell whether files are newer or older than other files.
2.  A programmer writes a [Makefile](glossary.html#makefile)
    to tell Make how files depend on each other.
    For example,
    the Makefile could say that `results/moby_dick.csv` depends on `data/moby_dick.txt`,
    or that `plots/aggregate.svg` depends on all of the CSV files in the `results/` directory.
3.  The Makefile includes shell commands to create or update files that are out of date.
    For example,
    it could include a command to (re-)run `bin/countwords.py`
    to create `results/moby_dick.csv` from `data/moby_dick.txt`.
    (Make's use on shell commands is one reason for its longevity,
    since it allows programmers to write tools for updating files
    in whatever language they want.)

Let's start by creating a file called `Makefile` that contains the following three lines:

<!-- src="automate/single_rule.mk" -->
```
# Regenerate results for "Moby Dick"
results/moby_dick.csv : data/moby_dick.txt
        python bin/countwords.py data/moby_dick.txt > results/moby_dick.csv
```

The `#` character starts a comment,
which runs to the end of a line (just as it does in Python or R).
`results/moby_dick.csv` is the [target](glossary.html#make-target) of a [rule](glossary.html#make-rule),
i.e., something that may need to be created or updated.
Every rule in a Makefile has one or more targets,
and must be written flush against the left margin.

`data/moby_dick.txt` is a [prerequisite](glossary.html#make-prerequisite) in that rule,
i.e.,
something that the target of the rule depends on.
A single colon separates the target from its prerequisites,
and a rule can have any number of prerequisites---we'll see examples soon.

The indented line that uses Python to run `bin/countwords.py`
is the rule's [action](glossary.html#make-action).
It creates or updates the target when it is out of date.
A rule can have any number of actions,
but they *must* be indented by a single tab character.
Notice that the output of `bin/countwords.py` is [redirected](glossary.html#redirection) using `>`
to create the output file:
we will look at modifying the script in Chapter \@ref(configure)
so that it can take the name of an output file as an argument.

Together,
the three parts of this rule tell Make when and how to re-create `results/moby_dick.csv`.
To test that it works,
run this command in the shell:

```shell
$ make
```

Make automatically looks for a file called `Makefile` and checks the rules it contains.
In this case,
one of three things will happen:

1.  Make won't be able to find the file `data/moby_dick.csv`,
    so it will run the script to create it.
2.  Make will see that `data/moby_dick.txt` is newer than `results/moby_dick.csv`,
    in which case it will run the script to update the results file.
3.  Make will see that `results/moby_dick.csv` is newer than its prerequisite,
    so it won't do anything.

In the first two cases,
Make will show the command it runs,
along with anything the command prints to the screen
via [standard output](glossary.html#stdout) or [standard error](glossary.html#stderr).
In this case,
there is no screen output,
so we only see the command.

> **Indentation Errors**
>
> If `Makefile` contains spaces instead of tabs to indent the rule's action,
> we will see an error message like this:
> 
> ```text
> Makefile:3: *** missing separator.  Stop.
> ```
>
> The requirement to use tabs is a legacy of Make's origins as a student intern project,
> and no,
> I'm not kidding.

If we run `make` again right away it doesn't re-run our script
because we're in situation #3 from the list above:
the target is newer than its prerequisites,
so no action is required.
We can check this by listing the files with their timestamps,
ordered by how recently they have been updated:

```shell
$ ls -l -t data/moby_dick.txt results/moby_dick.csv
```
```text
-rw-r--r--  1 gvwilson  staff   219107 31 Dec 08:58 results/moby_dick.csv
-rw-r--r--  1 gvwilson  staff  1276201 31 Dec 08:58 data/moby_dick.txt
```

When Make sees that a target is newer than its prerequisites
it displays a message like this:

```text
make: `results/moby_dick.csv' is up to date.
```

To test that Make is actually doing the right thing, we can:

1.  Delete `results/moby_dick.csv` and type `make` again (situation #1).
2.  Run the command `touch data/moby_dick.txt` to update the timestamp on the source file,
    then run `make` (situation #2).

## How can I tell Make where to find rules? {#automate-naming}

We don't have to call our file of rules `Makefile`.
If we want,
we can rename the file `single_rule.mk`
and then run it with `make -f single_rule.mk`.
Most people don't do this in real projects,
but in a lesson like this,
which includes many example Makefiles,
it comes in handy.

Using `-f` doesn't change our [working directory](glossary.html#working-directory).
If, for example, we are in `/home/gvwilson/still-magic` and run `make -f src/automate/single_rule.mk`,
our working directory remains `/home/gvwilson/still-magic`.
This means that Make will look for the rule's prerequisite in `/home/gvwilson/still-magic/data/moby_dick.txt`,
not in `/home/gvwilson/still-magic/src/automate/data/moby_dick.txt`.

FIXME: figure

## How can I update multiple files when their prerequisites change? {#automate-extend}

Our Makefile isn't particularly exciting so far.
Let's add another rule to the end and save the result as `double_rule.mk`:

<!-- src="automate/double_rule.mk" -->
```
# Regenerate results for "Moby Dick"
results/moby_dick.csv : data/moby_dick.txt
        python bin/countwords.py data/moby_dick.txt > results/moby_dick.csv

# Regenerate results for "Jane Eyre"
results/jane_eyre.csv : data/jane_eyre.txt
        python bin/countwords.py data/jane_eyre.txt > results/jane_eyre.csv
```

When we ask Make to run this file:

```shell
$ make -f double_rule.mk
```

<!-- == noindent -->
we get this rather disappointing message:

```text
make: `results/moby_dick.csv' is up to date.
```

Nothing happens because by default Make only attempts to update
the first target it finds in the Makefile,
which is called the [default target](glossary.html#default-target).
To update something else,
we need to tell Make we want it:

```shell
$ make -f double_rule.mk results/jane_eyre.csv
```

This time Make runs:

```text
python bin/countwords.py data/jane_eyre.txt > results/jane_eyre.csv
```

## How can I get rid of temporary files that I don't need? {#automate-phony}

A [phony target](glossary.html#phony-target) in a Makefile
is one that doesn't correspond to any files
and doesn't have any prerequisites.
Phony targets are just a way to save useful commands in a Makefile,
but saying "just" is a bit misleading:
what we're actually doing is recording all of the steps in our workflow,
even if those steps don't create or update files.

For example,
let's add another target to our Makefile to delete all of the files we have generated.
By convention this target is called `clean`,
and ours looks like this:

<!-- src="automate/clean.mk" -->
```
# Remove all generated files.
clean :
        rm -f results/*.csv
```

(The `-f` flag to `rm` means "force removal".
When we use it, `rm` won't complain if the files it's trying to remove are already gone.)
Let's run Make:

```shell
$ make -f pipeline.mk clean
```

<!-- == noindent -->
and then use `ls` to list the contents of `results`.
Sure enough, it's empty.

Phony targets are useful as a way of documenting actions in a project,
but there's a catch.
Use `mkdir` to create a directory called `clean`,
then run `make -f clean.mk clean`.
Make will print:

```text
make: `clean' is up to date.
```

The problem is that Make finds something called `clean` and assumes that's what the rule is referring to.
Since the rule has no prerequisites,
it can't be out of date,
so no actions are executed.

There are two ways to solve this problem.
The first is to make sure we don't have phony targets with the same names as files or directories.
That works as long as our project is small and we're paying attention,
but as the project grows,
or we're rushing to meet a deadline
or have inherited the project from someone else and don't realize that this might be a problem,
it's bound to fail at exactly the worst time.

A much better solution is to tell Make that the target is phony
by putting this in the Makefile:

```make
.PHONY : clean
```

Most people declare all of their phony targets together near the top of the file,
though some put the `.PHONY` declarations right before the rules they refer to.
As with most other rules about programming style (Chapter \@ref(style)),
consistency matters more than exactly what you do.

## How can I make a target depend on several prerequisites? {#automate-multi}

Right now,
our Makefile says that each result file depends only on the corresponding data file.
That's not accurate:
in reality,
each result also depends on the script used to generate it.
If we change our script,
we ought to regenerate our results and then check to see if they've changed.
(We can rely on version control to tell us that.)

Here's a modified version of the Makefile
in which each result depends on both the data file and the script:

<!-- src="automate/depend_on_script.mk" -->
```
.PHONY: clean

# Regenerate results for "Moby Dick"
results/moby_dick.csv : data/moby_dick.txt bin/countwords.py
	python bin/countwords.py data/moby_dick.txt > results/moby_dick.csv

# Regenerate results for "Jane Eyre"
results/jane_eyre.csv : data/jane_eyre.txt bin/countwords.py
	python bin/countwords.py data/jane_eyre.txt > results/jane_eyre.csv

# ...clean target as before...
```

We can test this by touching the script and then making one or the other result:

```shell
$ touch bin/countwords.py
$ make -f depend_on_script.mk results/jane_eyre.csv
```
```text
python bin/countwords.py data/jane_eyre.txt > results/jane_eyre.csv
```

## How can I reduce the amount of typing I have to do? {#automate-variables}

The name of our script now appears four times in our Makefile,
which will make for a lot of typing if and when we decide to move it or rename it.
We can fix that by defining a [variable](glossary.html#make-variable) at the top of our file
to refer to the script,
then using that variable in our rules:

<!-- src="automate/define_variable.mk" -->
```
.PHONY: clean

COUNT=bin/countwords.py

# Regenerate results for "Moby Dick"
results/moby_dick.csv : data/moby_dick.txt ${COUNT}
	python ${COUNT} data/moby_dick.txt > results/moby_dick.csv

# Regenerate results for "Jane Eyre"
results/jane_eyre.csv : data/jane_eyre.txt ${COUNT}
	python ${COUNT} data/jane_eyre.txt > results/jane_eyre.csv

# ...clean target as before...
```

The definition takes the form `NAME=value`.
By convention,
variables are written in UPPER CASE
so that they'll stand out from filenames (which are usually in lower case),
but that's not required.
What *is* required is using `${NAME}` to refer to the variable:
if we write `$NAME`,
Make interprets that as "the variable called `N` followed by the three literal characters 'AME'."
If no variable called `N` exists,
`$NAME` becomes `AME`,
which is almost certainly not useful.

Using variables doesn't just cut down on typing.
They also make rules easier to understand,
since they signal to readers that several things are always and exactly the same.

## How can I make one update depend on another? {#automate-chain}

We can re-create all the results files with a single command
by listing multiple targets when we run Make:

```shell
$ make results/moby_dick.csv results/jane_eyre.csv
```

However,
users have to know what files they might want to create in order to do this,
and have to type their names exactly right.
A better approach is to create a phony target that depends on all of the output files
and make it the first rule in the file so that it is the default.
By convention,
this target is called `all`,
and while we don't have to list all our phony targets in alphabetical order,
it makes them a lot easier to find:

<!-- src="automate/multi_prereq.mk" -->
```
.PHONY: all clean

COUNT=bin/countwords.py

# Regenerate all results.
all : results/moby_dick.csv results/jane_eyre.csv

# ...rules for moby_dick, jane_eyre, and clean...
```

If we run Make now,
it sees that `all` is only "up to date" if the two CSV files are up to date,
so it looks for a rule for each and runs each of those rules.

We can draw the prerequisites defined in the Makefile as a [dependency graph](glossary.html#dependency-graph),
with arrows showing what each target depends on.

FIXME: figure

Note that the Makefile doesn't define the order
in which `results/moby_dick.csv` and `results/jane_eyre.csv` are updated,
so Make can rebuild them in whatever order it wants.
This is called [declarative programming](glossary.html#declarative-programming):
we declare what outcome we want,
and the program figures out how to achieve it.

## How can I abbreviate my update rules? {#automate-automatic}

We could add a third book to our Makefile,
then a fourth,
but any time we find ourselves duplicating code,
there's almost certainly a way to write a general rule.
In order to create these,
though,
we first need to learn about [automatic variables](glossary.html#automatic-variable).

The first step is to use the very cryptic expression `$@` in the rule's action
to mean "the target of the rule".
We start with this:

<!-- src="automate/single_rule.mk" -->
```
# Regenerate results for "Moby Dick"
results/moby_dick.csv : data/moby_dick.txt
        python bin/countwords.py data/moby_dick.txt > results/moby_dick.csv
```

<!-- == noindent -->
and turn it into this:

<!-- src="automate/automatic_variables_wrong.mk" -->
```
# Regenerate results for "Moby Dick"
results/moby_dick.csv : data/moby_dick.txt
        python bin/countwords.py data/moby_dick.txt > $@
```

<!-- == noindent -->
`$@` is an automatic variable:
Make defines its value for us separately in each rule.
And yes,
`$@` is an unfortunate name:
something like `$TARGET` would be easier to understand,
but we're stuck with it.

Step 2 is to replace the list of prerequisites in the action with `$^`,
which is another automatic variable meaning "all the prerequisites of the current rule":

<!-- src="automate/automatic_variables_wrong.mk" -->
```
# Regenerate results for "Jane Eyre"
results/jane_eyre.csv : data/jane_eyre.txt
        python bin/countwords.py $^ > $@
```

But wait:
our results files don't just have books as dependencies.
They also depend on `bin/countwords.py`.
What happens if we include that in the rule while using automatic variables?
(We'll do this for a third book to keep the three rules separate in the example Makefile.)

<!-- src="automate/automatic_variables_wrong.mk" -->
```
# Regenerate results for "The Time Machine" - WRONG
results/time_machine.csv : data/time_machine.txt ${COUNT}
        python bin/countwords.py $^ > $@
```

<!-- == noindent -->
This doesn't do the right thing because `$^` includes *all* of the prerequisites,
so the action tries to process the script as if it were a data file:

```shell
python bin/countwords.py data/time_machine.txt bin/countwords.py results/time_machine.csv
```

This situation comes up so often that
Make helpfully provides another automatic variable `$<` meaning "the first prerequisite",
which lets us rewrite our rules like this:

<!-- src="automate/automatic_variables.mk" -->
```
# Regenerate results for "Janey Eyre"
results/jane_eyre.csv : data/jane_eyre.txt ${COUNT}
        python bin/countwords.py $< > $@
```

And yes,
`$< > $@` is hard to read,
even with practice,
and `< $<` (reading the first prerequisite from [standard input](glossary.html#stdin))
is even harder.
Using an editor that does syntax highlighting helps (Chapter \@ref(tools)),
and if you are ever designing software for other people to use,
remember this case and don't do it.

## How can I write one general rule to update many files in the same way? {#automate-pattern}

We can now replace all the rules for generating results files
with one [pattern rule](glossary.html#pattern-rule)
that uses `%` as a [wildcard](glossary.html#wildcard).
Whatever part of a filename `%` matches in the target,
it must also match in the prerequisites,
so the single rule:

<!-- src="automate/pattern_rule.mk" -->
```
results/%.csv : data/%.txt ${COUNT}
        python bin/countwords.py $< > $@
```

<!-- == noindent -->
will handle *Jane Eyre*, *Moby Dick*, and *The Time Machine*.
(Unfortunately, `%` cannot be used in rules' actions,
which is why `$<` and `$@` are needed.)
With this rule in place,
our entire Makefile is reduced to:

<!-- src="automate/pattern_rule.mk" -->
```
.PHONY: all clean

COUNT=bin/countwords.py

# Regenerate all results.
all : results/moby_dick.csv results/jane_eyre.csv results/time_machine.csv

# Regenerate result for any book.
results/%.csv : data/%.txt ${COUNT}
	python ${COUNT} $< > $@

# Remove all generated files.
clean :
	rm -f results/*.csv
```

Let's delete all of the results files and re-create them all:

```shell
$ make -f pattern_rule.mk clean
```
```text
rm -f results/*.csv
```

```shell
$ make -f pattern_rule.mk all
```
```text
python bin/countwords.py data/moby_dick.txt > results/moby_dick.csv
python bin/countwords.py data/jane_eyre.txt > results/jane_eyre.csv
python bin/countwords.py data/time_machine.txt > results/time_machine.csv
```

We can still rebuild individual files:

```shell
$ touch data/jane_eyre.txt
$ make -f pattern_rule.mk results/jane_eyre.csv
```
```text
python bin/countwords.py data/jane_eyre.txt > results/jane_eyre.csv
```

## How can I define sets of files automatically? {#automate-functions}

Our "automated" analysis is still not fully automated:
If we add another book to `raw`,
we have to remember to add it to the `all` target in the Makefile as well.
Once again,
we will fix this in steps.

To start,
imagine that all the results files already exist,
and we just want to update them.
We can define a variable called `RESULTS` to be a list of all the results files
using the same notation we'd use in the shell to match all the CSV files in the `results/` directory:

<!-- src="automate/filename_wildcard.mk" -->
```
RESULTS=results/*.csv
```

<!-- == noindent -->
and then make `all` depend on that:

<!-- src="automate/filename_wildcard.mk" -->
```
# Regenerate all results.
all : ${RESULTS}
```

This works,
but only for re-creating files:
if a results file doesn't already exist when we run Make,
its name won't be included in `RESULTS`,
and Make won't realize that we want to generate it.

What we really want to do is generate the list of results files
from the list of books in the `data/` directory.
We can use a [function](glossary.html#make-function) to do this.
The syntax is a little odd,
because functions were added to Make long after it was first written,
but at least they have readable names.
Let's create a variable `DATA` that holds the names of all of our data files:

<!-- src="automate/function_wildcard.mk" -->
```
DATA = $(wildcard data/*.txt)
```

<!-- == noindent -->
This calls the function `wildcard` with the argument `data/*.txt`.
The result is a list of all the text files in the `raw` directory,
just as we'd get with `data/*.txt` in the shell.
(We could use a shell wildcard here as we did when defining `RESULTS`,
but we want to show how functions work.)

Did this do the right thing?
To check,
we can add another phony target to the end of the file called `settings`
that uses the shell command `echo` to print the name and value of a variable:

<!-- src="automate/function_wildcard.mk" -->
```
.PHONY: all clean settings

# ...everything else...

# Show variables' values.
settings :
        echo COUNT: ${COUNT}
        echo DATA: ${DATA}
```

Let's run this:

```shell
$ make -f function_wildcard.mk settings
```
```text
echo COUNT: bin/countwords.py
COUNT: bin/countwords.py
echo DATA: data/common_sense.txt data/jane_eyre.txt data/life_of_frederick_douglass.txt data/moby_dick.txt data/sense_and_sensibility.txt data/time_machine.txt
DATA: data/common_sense.txt data/jane_eyre.txt data/life_of_frederick_douglass.txt data/moby_dick.txt data/sense_and_sensibility.txt data/time_machine.txt
```

The output appears twice because Make shows us the command it's going to run before running it.
If we put `@` before the command,
Make doesn't show it before running it:

<!-- src="automate/function_wildcard_silent.mk" -->
```
settings :
	@echo COUNT: ${COUNT}
        @echo DATA: ${DATA}
```

```shell
$ make -f function_wildcard.mk settings
```
```text
COUNT: bin/countwords.py
DATA: data/common_sense.txt data/jane_eyre.txt data/life_of_frederick_douglass.txt data/moby_dick.txt data/sense_and_sensibility.txt data/time_machine.txt
```

We now have the names of our input files,
but what we want is the names of the corresponding output files.
Make has another function called `patsubst` (short for "**pat**tern **subst**itution")
that uses the same kind of patterns used in rules to do exactly this:

<!-- src="automate/patsubst.mk" -->
```
RESULTS=$(patsubst data/%.txt,results/%.csv,${DATA})
```

`$(patsubst ...)` calls the pattern substitution function.
The first argument is what to look for:
in this case,
a text file in the `raw` directory.
As in a pattern rule,
we use `%` to match the [stem](glossary.html#filename-stem) of the file's name,
which is the part we want to keep.

The second argument is the replacement we want.
Ours uses the stem matched by `%` to construct the name of a CSV file in the `results` directory.
Finally,
the third argument is what we're doing substitutions in,
which is our list of books' names.

Let's check that this has worked by adding to the `settings` target

<!-- src="automate/patsubst.mk" -->
```
settings :
        @echo COUNT: ${COUNT}
        @echo DATA: ${DATA}
        @echo RESULTS: ${RESULTS}
```
```shell
$ make -f patsubst.mk settings
```
```text
COUNT: bin/countwords.py
DATA: data/common_sense.txt data/jane_eyre.txt data/life_of_frederick_douglass.txt data/moby_dick.txt data/sense_and_sensibility.txt data/time_machine.txt
RESULTS: results/common_sense.csv results/jane_eyre.csv results/life_of_frederick_douglass.csv results/moby_dick.csv results/sense_and_sensibility.csv results/time_machine.csv
```

Excellent:
`DATA` has the names of all of the files we want to process
and `RESULTS` automatically has the corresponding names of the files we want to generate.
Let's test it:

```shell
$ make -f patsubst.mk clean
```
```text
rm -f results/*.csv
```
```shell
$ make -f patsubst.mk all
```
```text
python bin/countwords.py data/common_sense.txt > results/common_sense.csv
python bin/countwords.py data/jane_eyre.txt > results/jane_eyre.csv
python bin/countwords.py data/life_of_frederick_douglass.txt > results/life_of_frederick_douglass.csv
python bin/countwords.py data/moby_dick.txt > results/moby_dick.csv
python bin/countwords.py data/sense_and_sensibility.txt > results/sense_and_sensibility.csv
python bin/countwords.py data/time_machine.txt > results/time_machine.csv
```

Our workflow is now just two steps:
add a data file and run Make.
As we'll see in Chapter \@ref(integrate),
we can even automate the second half in some cases,
but this is still a big improvement over running things manually,
particularly as we start to add more steps
(such as merging data files and generating plots).

## How can I document my workflow? {#automate-doc}

Every well-behaved program can tell people how to use it.
If we run `make --help`, for example,
we get a (long) list of things Make can do for us.

But how can we document the workflow that our Makefile now embodies?
One common choice is to provide a special target like `settings`
that prints a description of available targets:

<!-- src="automate/makefile_help.mk" -->
```
.PHONY: all clean help settings

# ...other definitions...

# Show help.
help :
	@echo "all : regenerate all out-of-date results files."
	@echo "results/*.csv : regenerate a particular results file."
	@echo "clean : remove all generated files."
	@echo "settings : show the values of all variables."
	@echo "help : show this message."
```

This is easy to set up and does the job,
but once again its redundancy should worry us:
the same information appears in both the comments on rules and the help,
which means that authors have to remember to update the help when adding or changing rules.

A better approach,
which we will explore in more depth in Chapter \@ref(docs),
is to have people format some comments in a special way
and then extract and display those comments when asked for help.
We'll use `##` (a double comment marker) to indicate the lines we want displayed
and use `grep` to extract lines that start with that marker.
We will use Make's `MAKEFILE_LIST` variable to get the path to the Makefile,
since we may be using the `-f` flag to specify which Makefile we're using.
With all that in place,
our finished Makefile is:

<!-- src="automate/makefile_grep.mk" -->
```
.PHONY: all clean help settings

COUNT=bin/countwords.py
DATA=$(wildcard data/*.txt)
RESULTS=$(patsubst data/%.txt,results/%.csv,${DATA})

## all : regenerate all results.
all : ${RESULTS}

## results/*.csv : regenerate result for any book.
results/%.csv : data/%.txt ${COUNT}
	python ${COUNT} $< > $@

## clean : remove all generated files.
clean :
	rm -f results/*.csv

## settings : show variables' values.
settings :
	@echo COUNT: ${COUNT}
	@echo DATA: ${DATA}
	@echo RESULTS: ${RESULTS}

## help : show this message.
help :
	@grep '^##' ${MAKEFILE_LIST}
```

Let's test:

```shell
$ make -f makefile_grep.mk
```
```text
## all : regenerate all results.
## results/*.csv : regenerate result for any book.
## clean : remove all generated files.
## settings : show variables' values.
## help : show this message.
```

With a bit more work we could strip off the leading `##` markers,
but this is a good start.

> **How did you know that?**
>
> FIXME: keep this personal?
>
> I had never used the variable `MAKEFILE_LIST` before writing this lesson.
> In fact, until about 15 minutes ago, I didn't even know it existed:
> I always had my `help` target's action `grep` for `##` in `Makefile`.
> Once I realized that wouldn't work in this example
> (because I'm writing lots of little Makefiles to demonstrate ideas step by step)
> I searched online for "how to get name of Makefile in make".
> The second hit took me to [the GNU Make documentation for other special variables][gnu-make-other-vars],
> which told me exactly what I needed.
> I spend anywhere from a quarter to three quarters of my time searching for things when I program these days;
> one of the goals of these lessons is to give you an idea of what you ought to be searching for yourself
> so that you can do this more efficiently.

## Summary {#automate-summary}

```{r automate-concept, echo=FALSE, fig.cap="Automation Concept Map"}
if (knitr::is_latex_output()) {
  knitr::include_graphics("figures/automate/concept.pdf")
} else {
  knitr::include_graphics("figures/automate/concept.svg")
}
```

-   @Smit2011 describes the design and implementation of several build tools in detail.

## Exercises {#automate-exercises}

### Create a summary CSV file

-   Add rule to Makefile to create a summary CSV file from all of the book CSV files.
-   Careful how to write the dependencies so that it doesn't depend on itself.

### Generate a plot for the top N words

-   FIXME: make it depend on the summary.

## Key Points {#automate-keypoints}

```{r, child="keypoints/automate.md"}
```
