# A Branching Workflow {#s:branches}

## Objectives

```{r, child="objectives/branches.md"}
```

A common Git workflow for single-author/single-user projects is:

1.  Make some changes, running `git add` and `git commit` as you go along.
2.  `git push` to send those changes to a remote repository for backup and sharing.
3.  Switch computers (e.g., go from your laptop to the cluster where you're going to process full-sized data sets).
4.  `git pull` to get your changes from the remote repository,
    and possibly resolve merge conflicts if you made some changes on the cluster the last time you were there
    but forgot to push them to the remote repo.

Occasionally,
you'll run `git checkout -- .` or `git reset --hard VERSION` to discard changes.
Less frequently,
you'll use `git diff -r abcd1234 path/to/file` or `git show abcd1234:path/to/file`
to look at what's changed in files,
or `git checkout abcd1234 path/to/file` followed by `git add` and `git commit`
to restore an old version of a file.
And every once in a while,
you'll [jenny](glossary.html#jenny) the repository and clone a new copy because everything is messed up.

This workflow essentially uses Git to do backups and as a replacement for `scp`
(or `ftp`, if you're of an older generation).
But sometimes you have to work on several things at once in a single project,
or need to set aside current work for a high-priority interrupt.
And even if those situations don't arise,
this workflow doesn't provide guidance for collaborating with others.

This lesson introduces a few working practices that will help you do these things.
Along the way, it shows you how to use some of Git's more advanced capabilities.
As with everything involving Git,
the syntax for commands is more complicated than it needs to be @Pere2013,@Pere2016,
but if used carefully,
they can make you much more productive.

## How can I use branches to manage development of new features? {#s:branches-branch}

Experienced developers tend to use a [branch-per-feature workflow](glossary.html#branch-per-feature-workflow),
even when they are working on their own.
Whenever they want to create a new feature,
they create a new branch from `master` and do the work there.
When it's done,
they merge that branch back into `master.

More specifically,
a branch-per-feature workflow looks like this:

1.  `git checkout master` to make sure you're working in the `master` branch.

2.  `git pull origin master` (or `git pull upstream master`) to make sure that
    your starting point is up to date
    (i.e., that the changes you committed while working on the cluster yesterday are now in your local copy,
    or that the fix your colleague did last week has been incorporated).

3.  <code>git checkout -b <em>name-of-feature</em></code> to create a new branch.
    The branch's name should be as descriptive as the name of a function
    or of a file containing source code:
    `"stuff"` or `"fixes"` saves you a couple of seconds of typing when you create the branch,
    and will cost you much more time than that later on
    when you're juggling half a dozen branches
    and trying to figure out what's in each.

4.  Do the work you want to do.
    If some other task occurs to you while you're doing it---for example,
    if you're writing a new function and realize that the documentation for some other function should be updated---*don't*
    do that work in this branch just because you happen to be there.
    Instead,
    commit your changes,
    switch back to `master`,
    create a new branch for the other work,
    and carry on.

5.  Periodically,
    you may <code>git push origin <em>name-of-feature</em></code> to save your work to GitHub.

6.  When the work is done,
    you `git pull origin master` to get any changes that anyone else has put in the master copy
    into your branch
    and merge any conflicts.
    This is an important step:
    you want to do the merge *and test that everything still works* in *your* branch,
    not in `master`.

7.  Finally,
    use `git checkout master` to switch back to the `master` branch on your machine
    and <code>git merge <em>name-of-feature</em> master</code> to merge your changes into `master`.
    Test it one more time,
    and then `git push origin master` to send your changes to the remote repository
    for everyone else to admire and use.

It takes longer to describe this workflow than to use it.
Right now,
for example,
the repository for this material currently has six active branches:
`master` (which is the published version),
`explain-branch-per-feature` (where this is being written),
and four others created from `master` as reminders to fix a glossary entry,
rewrite the description in `README.md` of how to generate a PDF of these notes,
and so on.
This approach essentially uses branches as a to-do list,
which takes no more time than typing a brief note about the feature into a text file
or writing it in a lab notebook,
and is much easier to track.
Chapter \@ref(s:backlog) will look at better ways to manage this.

## How can I switch between branches when work is only partly done? {#s:branches-switching}

So what happens if we are in the middle of making changes on branch X
and realize that we need to do some reorganizing that has nothing to do with
the feature we're currently working on?
The easy answer is to commit our half-finished work,
switch branches,
do the other task,
and then switch back.
However,
that workflow leaves a lot of commits in our history corresponding to half-done work,
and ideally we want every commit to move the system from one useful state to another.
And if we make a dozen commits while working on a feature and then merge that branch into `master`,
the overall history of the project can become very difficult to read.

There is a command called `git stash` that will temporarily save work on a branch.
We don't recommend using it,
since it creates yet another place where valuable information might be stored (or forgotten).
Instead,
commit changes in the branch as you work,
then squash those commits into one single comprehensible commit
using the command `git rebase` (discussed in [the next section](#s:branches-rebase)).
After squashing,
you can merge that single large (meaningful) commit into `master`.

Like many things involving Git,
this is easier to understand with a diagram.
Suppose we have created a branch to work on a feature:

FIXME: figure

<!-- == noindent -->
We make several small changes,
then realize we need to fix something else,
so we commit those changes:

FIXME: figure

<!-- == noindent -->
switch to `master`, and create a new branch to do the other work.
This happens several times,
eventually leaving the repository in this state:

FIXME: figure

<!-- == noindent -->
If we were to merge now,
four commits would wind up in the history of the `master` branch.
Instead,
we squash those four commits into one:

FIXME: figure

<!-- == noindent -->
and then merge that:

FIXME: figure

## How can I keep my project's history clean when working on many branches? {#s:branches-rebase}

The workflow described above depends on [rebasing](glossary.html#rebase),
which means moving or combining some commits from one branch to another.
`git rebase` is a powerful command:
it can replay changes made in one branch on top of changes made to another:

FIXME: figure

<!-- == noindent -->
or collapse several consecutive commits into a single commit,
as we saw in the [previous section](#s:branches-switching).
We will only use it for the latter ability.

The command we want is `git rebase -i BASE`,
where `BASE` is the most recent commit *before* the sequence to be compressed
(i.e., the one that everything else will be based on).
Many people find this confusing,
and frequently ask `git rebase` to start with the first commit they want changed
rather than the last one that they don't.

When we run `git rebase -i`,
it brings up a display of recent commits
in the same editor we would use for writing commit messages:

FIXME: figure

The help text in this display tells us what we can do;
in most cases,
people choose to `pick` the first commit and `squash` the rest.
When we save this file,
Git immediately launches the editor again to show us
the combination of recent changes and messages:

FIXME: figure

<!-- == noindent -->
We can now edit this text to create a commit message for the unified commit.
After we have done this,
`git log` will only show us this single commit,
because it's the only one left in our history:

FIXME: figure

`git rebase` is a complex command,
and if we have merged changes from other branches into the branch we're rebasing,
Git can become confused
(or rather, it can confuse us).
A good rule to follow is,
"Don't rebase branches that are shared with other people."
Instead,
do all the work required for a feature,
rebase,
and then merge in recent changes from `master` and merge back to `master`:

FIXME: figure

What if you have pushed a branch to GitHub (or elsewhere)
and then combine the commits to change its history
as shown below:

FIXME: figure

<!-- == noindent -->
In this case,
Git realizes that your merge would lose information and prevents the operation from going through.
You can use `git push --force` to overwrite the remote history,
but this is usually a sign that you should have done something differently a while back.
Remember,
`git push --force` will also overwrite any work that other people have pushed to the repository,
so it's a good way to end friendships.

## How can I make it easy for people to review my work before I merge it? {#s:branches-pull-request}

The biggest benefit of having a second person work on a programming project
is not getting twice as much code written,
but having a second pair of eyes look at the software.
Chapter \@ref(s:style) will discuss how to do code review,
but a good first step depends on using a branch-per-feature workflow.

In order for someone to review your change,
that change and the original have to somehow be put side-by-side.
The usual way to do this is to create a [pull request](glossary.html#pull-request),
which is a marker saying, "Someone would like to merge branch A into branch X".
The pull request does *not* contain the changes:
instead,
it points at two branches,
so that if either branch is changed,
the differences displayed are always up to date.

FIXME: figure

Pull requests can be created between two branches in a single repository,
or between branches of different repositories.
The latter is common in projects with many contributors:
each contributor works in a branch in their own [fork](glossary.html#fork) of the repository,
and when they're done,
they create a pull request offering to merge their changes into the `master` branch
of the original repository.
In order to update their own repository,
they pull changes from the original:

FIXME: figure

A pull request can store more than the location of the source and destination branch.
In particular,
it can store comments that people have made about the proposed merge.
GitHub and other [forges](glossary.html#forge) allow users to add comments on the pull request as a whole,
or on particular lines,
and can mark old comments as out of date if the lines of code the comment is attached to are updated.

Pull requests aren't just for code.
If you are using a [typesetting language](glossary.html#typesetting-language) like Markdown or LaTeX to write your papers,
you can create pull requests for changes
so that your colleagues can comment on them,
then revise and push again when you incorporate their suggestions.
This workflow has all the benefits of adding comments to Google Docs,
but scales to large projects or large numbers of contributors.
The downside,
of course,
is that you have to use a typesetting language rather than a [WYSIWYG](glossary.html#wysiwyg) editor,
because the programmers who created Git and other version control systems
still don't support anything that couldn't be put on a punchcard in 1965.

## What *is* a feature? {#s:branches-thinking}

But what is a "feature", exactly?
What's large enough to merit creation of a new branch?
These rules make sense for small projects with or without collaborators:

1.  Anything cosmetic that is only one or two lines long can be done in `master` and committed right away.
    "Cosmetic" means changes to comments or documentation:
    nothing that affects how code runs---not even a simple variable renaming---is done this way,
    because experience has taught that things that aren't supposed to often do.

2.  A pure addition that doesn't change anything else is a feature and goes into a branch.
    For example,
    if you run a new analysis and save the results,
    that should be done on its own branch
    because it might take several tries to get the analysis to run,
    and you might interrupt yourself several times to fix things that you've discovered aren't working.

3.  Every change to code that someone might want to undo later in one step gets done as a feature.
    For example,
    if a parameter is added to a function,
    every call to the function has to be updated;
    since neither alteration makes sense without the other,
    it's considered a single feature and should be done in one branch.

The hardest thing about using a branch-per-feature workflow is actually doing it for small changes.
As the first point in the list above suggests,
most people are pragmatic about this on small projects;
on large ones,
where dozens of people might be committing,
even the smallest and most innocuous change needs to be in its own branch
so that it can be reviewed (which we discuss below).

The other thing that's hard to do with a branch-per-feature workflow is a major code reorganization.
If many files are being moved, renamed, and altered in order to restructure the project,
merging branches where those changes *haven't* been made can be tedious and error-prone.
The solution is to not get into this situation:
as Chapter \@ref(s:refactor) says,
code should be reorganized in many small steps,
not one big one.

## How can I label specific versions of my work? {#s:branches-tag}

A [tag](glossary.html#git-tag) is a permanent label on a particular state of the repository.
Tags are theoretically redundant,
since the [commit hash](glossary.html#commit-hash) identifies that state as well,
but commit hashes are (deliberately) random and therefore hard to remember or find.

Experienced developers Use [annotated tags](glossary.html#annotated-tag)
to mark every major event in the project's history.
These tags are called "annotated" because they allow their creator to specify a message,
just like a commit.
Research projects often use `report-date-event` for tag names,
such as `jse-2018-06-23-response` or `pediatrics-2018-08-15-summary`.
If you do this,
please don't tempt fate by calling something `-final`.

Most software projects use [semantic versioning](glossary.html#semantic-versioning) for software releases,
which produces three-part version numbers `major.minor.patch`:

-   Increment `major` every time there's an incompatible externally-visible change.
-   Increment `minor` when adding functionality without breaking any existing code.
-   Increment `patch` for bug fixes that don't add any new features ("now works as previously documented").

Simple projects only tag the `master` branch
because everything that is finished is merged to `master`.
Larger software projects may create a branch for each released version and do minor or patch updates on that branch,
but this outside the scope of this lesson.

## Summary {#s:branches-summary}

FIXME: create concept map for workflow

## Exercises {#s:branches-exercises}

FIXME: exercises for branching

## Key Points

```{r, child="keypoints/branches.md"}
```

```{r, child="etc/links.md"}
```
