# Including Everyone {#inclusive}

## Questions {#inclusive-questions}

```{r, child="questions/inclusive.md"}
```

## Objectives {#inclusive-objectives}

```{r, child="objectives/inclusive.md"}
```

## Introduction {#inclusive-intro}

The previous lesson talked about the physical organization of projects.
This one talks about the social structure,
which is more important to the project's success.
A project can survive badly-organized code;
none will survive for long if people are confused,
pulling in different directions,
or hostile.
This lesson therefore talks about what projects can do to make newcomers feel welcome
and to make things run smoothly after that.
It draws on @Foge2005,
which describes how good open source software projects are run,
and on @Boll2014,
which explains what a [commons](glossary.html#commons) is and when it's the right model to use.

## Why does a project need explicit rules? {#inclusive-rules}

Jo Freeman's influential essay
"[The Tyranny of Structurelessness][tyranny-structurelessness]" pointed out that
every group has a power structure;
the only question is whether it is formal and accountable
or informal and unaccountable @Free1972.
Thirty-five years after the free software movement took on its modern, self-aware form,
its successes and failures have shown that if a project doesn't clearly state
who has the right to do what,
it will wind up being run by whoever argues loudest and longest.

## How should I license my software? {#inclusive-software-license}

It might seem strange to put licensing under discussion of inclusivity,
but if the law or a publication agreement prevents people from reading your work or using your software,
you're excluding them
(and probably hurting your own career).
You may need to do this in order to respect personal or commercial confidentiality,
but the first and most important rule of inclusivity
is to be open by default.

However,
that is easier said than done,
not least because the law hasn't kept up with everyday practice.
@Mori2012 and [this blog post][vanderplas-licensing] are good starting points from a scientist's point of view,
while @Lind2008 is a deeper dive for those who want details.
In brief,
creative works are automatically eligible for intellectual property (and thus copyright) protection.
This means that every creative work has some sort of license:
the only question is whether authors and users know what it is.

Every project should therefore include an explicit license.
This license should be chosen early:
if you don't set it up right at the start,
then each collaborator will hold copyright on their work
and will need to be asked for approval when a license *is* chosen.
By convention,
the license is usually put in a file called `LICENSE` or `LICENSE.txt` in the project's root directory.
This file should clearly state the license(s) under which the content is being made available;
the plural is used because code, data, and text may be covered by different licenses.

> *Don't write your own license*,
> even if you are a lawyer:
> legalese is a highly technical language,
> and words don't mean what you think they do.

To make license selection as easy as possible,
GitHub allows you to select one of the most common licenses when creating a repository.
The Open Source Initiative maintains [a list of licenses][osi-license-list],
and [choosealicense.com][choose-license] will help you find a license that suits your needs.
Some of the things you will need to think about are:

1.  Do you want to license the code at all?
2.  Is the content you are licensing source code?
3.  Do you require people distributing derivative works to also distribute their code?
4.  Do you want to address patent rights?
5.  Is your license compatible with the licenses of the software you depend on?
    For example, as we will discuss below,
    you can use MIT-licensed code in a GPL-licensed project but not vice versa.

The two most popular licenses for software are
the [MIT license](glossary.html#mit-license) and the [GNU Public License](glossary.html#gpl) (GPL).
The MIT license (and its close sibling the BSD license)
say that people can do whatever they want to with the software as long as they cite the original source,
and that the authors accept no responsibility if things go wrong.
The GPL gives people similar rights,
but requires them to share their own work on the same terms:

> You may copy, distribute and modify the software as long as you track changes/dates in source files.
> Any modifications to or software including (via compiler) GPL-licensed code must also be made available under the GPL
> along with build & install instructions.
>
> --- [tl;dr][tldr-gpl]

We recommend the MIT license:
it places the fewest restrictions on future action,
it can be made stricter later on,
and the last thirty years shows that it's good enough to keep work open.

## How should I license my data and reports? {#inclusive-software-text}

The MIT license and the GPL apply to software.
When it comes to data and reports,
the most widely used family of licenses are those produced by [Creative Commons][creative-commons],
which have been written and checked by lawyers and are well understood by the community.

The most liberal license is referred to as [CC-0](glossary.html#cc-0),
where the "0" stands for "zero restrictions".
CC-0 puts work in the public domain,
i.e.,
allows anyone who wants to use it to do so however they want with no restrictions.
This is usually the best choice for data,
since it simplifies aggregate analysis.
For example,
if you choose a license for data that requires people to cite their source,
then anyone who uses that data in an analysis must cite you;
so must anyone who cites *their* results,
and so on,
which quickly becomes unwieldy.

The next most common license is the Creative Commons - Attribution license,
usually referred to as [CC-BY](glossary.html#cc-by).
This allows people to do whatever they want to with the work
as long as they cite the original source.
This is the best license to use for manuscripts,
since you *want* people to share them widely
but also want to get credit for your work.

Other Creative Commons licenses incorporate various restrictions on specific use cases:

-   ND (no derivative works) prevents people from creating modified versions of your work.
    Unfortunately, this also inhibits translation and reformatting.
-   NC (no commercial use) does *not* mean that people cannot charge money for something that includes your work,
    though some publishers still try to imply that in order to scare people away from open licensing.
    Instead,
    the NC clause means that people cannot charge for something that uses your work without your explicit permission,
    which you can give under whatever terms you want.
-   Finally,
    SA (share-alike) requires people to share work that incorporates yours
    on the same terms that you used.
    Again,
    this is fine in principle,
    but in practice makes aggregation a headache.

## Why should I establish a code of conduct for my project? {#inclusive-conduct}

You don't expect to have a fire,
but every large building or event should have a fire safety plan.
Similarly,
having a Code of Conduct for your project
reduces the uncertainty that participants face about what is acceptable and unacceptable behavior.
You might think this is obvious,
but long experience shows that articulating it clearly and concisely reduces problems caused by have different expectations,
particularly when people from very different cultural backgrounds are trying to collaborate.
An explicit Code of Conduct is particularly helpful for newcomers,
so having one can help your project grow
and encourage people to give you feedback.

Having a Code of Conduct is particularly important for people from marginalized or under-represented groups,
who have probably experienced harassment or unwelcoming behavior before.
By adopting one,
you signal that your project is trying to be a better place than YouTube,
Twitter,
and other online cesspools.
Some people may push back claiming that it's unnecessary,
or that it infringes freedom of speech,
but in our experience,
what they often mean is that thinking about how they might have benefited from past inequity makes them feel uncomfortable,
or that they like to argue for the sake of arguing.
If having a Code of Conduct leads to them going elsewhere,
that will probably make your project run more smoothly.

Just as you shouldn't write your own license for a project,
you probably shouldn't write your own Code of Conduct.
We recommend using the [Contributor Covenant][covenant] for development projects
and the [model code of conduct][model-coc] from the [Geek Feminism Wiki][geek-feminism] for in-person events.
Both have been thought through carefully and revised in the light of experience,
and both are now used widely enough that
many potential participants in your project will not need to have them explained.

Rules are meaningless if they aren't enforced.
If you adopt a Code of Conduct,
it is therefore important to be clear about how to report issues and who will handle them.
@Auro2018 is a short, practical guide to handling incidents;
like the Contributor Covenant and the model code of conduct,
it's better to start with something that other people have thought through and refined
than to try to create something from scratch.

## Why can I be a good ally for members of marginalized groups? {#inclusive-why-ally}

Setting out rules and handling incidents when they arise is what projects can do;
if you have power
(even or especially the power that comes from being a member of the majority group),
what you can do personally is be a good ally for members of marginalized groups.
Much of this discussion is drawn from the [Frameshift Consulting Ally Skills workshop][ally-skills],
which you should attend if you can.

First,
some definitions.
[Privilege](glossary.html#privilege) is an unearned advantage given to some people but not all;
[oppression](glossary.html#oppression) is systemic, pervasive inequality that benefits the privileged
and harms those without privilege.
A straight, white, physically able, economically secure male
is less likely to be interrupted when speaking,
more likely to be called on in class,
and more likely to get a job interview based on an identical CV
than someone who is perceived as being outside these categories.
The unearned advantage may be small in any individual case,
but compound interest quickly amplifies these differences:
someone who is called on more often in class is more likely to be remembered by a professor,
who in turn is therefore more likely to recommend them to a potential employer,
who is more likely to excuse the poor grades on their transcripts,
and on and on it goes.
People who are privileged are often not aware of it
for the same reason that most fish don't know what water tastes like.

A [target](glossary.html#target-oppression) is someone who suffers from oppression.
Targets are often called "members of a marginalized group",
but that phrasing is deliberately passive.
Targets don't choose to be marginalized:
those with privilege marginalize them.
An [ally](glossary.html#ally) is a member of a privileged group
who is working to understand their own privilege and end oppression.
For example,
privilege is being able to walk into a store and have the owner assume you're there to buy things,
not to steal them.
Oppression is the stories told about (for example) indigenous people being thieves,
and the actions people take as a result of them.
A target is an indigenous person who wants to buy milk,
and an ally is a white person who pays attention to a lesson like this one (raising their own awareness),
calls out peers who spread racist stories (a [peer action](glossary.html#peer-action)),
or asks the shopkeeper whether they should leave too (a [situational action](glossary.html#situational-action)).

Why should you be an ally?
You could do it out of a sense of fairness
because you realize that you have benefited from oppression
even if you haven't taken part (or don't think you have).
And you should do it because you can:
taking action to value diversity results in worse performance ratings for minority and female leaders,
while ethnic majority or male leaders who do this aren't penalized @Hekm2017.
As soon as you acknowledge that (for example) women are called on less often than men,
or are less likely to get an interview or a publication given identical work,
you have to acknowledge that white and Asian males are *more* likely to get these benefits than their performance alone deserves.

## How can I be a good ally for members of marginalized groups? {#inclusive-how-ally}

So much for the theory:
what should you actually do?
A few simple rules will go a long way:

1.  Be short, simple, and firm.
2.  Don't try to be funny: it almost always backfires, or will later be used against you.
3.  Play for the audience:
    you probably won't change the mind of the oppressor you're calling out,
    but you might change the minds or give heart to people who are observing.
4.  Pick your battles.
    You can't challenge everyone, every time,
    without exhausting yourself and deafening your audience.
    An occasional sharp retort will be much more effective than constant criticism.
5.  Don't shame or insult one group when trying to help another.
    For example,
    don't call someone stupid when what you really mean is that they're racist or homophobic.
6.  Change the terms of the debate.

The last rule is best explained by example.
Suppose someone says,
"Why should we take diversity into account when hiring?
Why don't we just hire the best candidate?"
Your response could be,
"Because taking diversity into account *is* hiring the best candidate.
If you can run a mile in four minutes and someone else can do it in 4:15 with a ball and chain on their leg,
who the better athlete?
Who will perform better *if the impediment is removed*?
If you intend to preserve an exclusionary culture in this lab,
considering how much someone has achieved despite systemic unfairness might not make sense,
but you're not arguing for that,
are you?"
And if someone then says,
"But it's not fair to take anything other than technical skill into account when hiring for a technical job,"
you can say,
"You're right,
which means that what you're *really* upset about is the thought that
you might be treated the way targets have been treated their whole lives."

[Captain Awkward][captain-awkward] has useful advice,
and [Charles' Rules of Argument][charles-rules] are very useful online:

1.  Don't go looking for an argument.
2.  State your position once, speaking to the audience.
3.  Wait for absurd replies.
4.  Reply once more to correct any misunderstandings of your original statement.
5.  Do not reply again---go do something fun instead.

Finally,
it's important to recognize that good principles sometimes conflict.
For example,
consider this scenario:

> A manager consistently uses male pronouns to refer to software and people of unknown gender.
> When you tell them it makes you uncomfortable to treat maleness as the norm,
> they say that male is the default gender in their first language
> and you should be more considerate of people from other cultures.

On the one hand,
you want to respect other people's cultures;
on the other hand,
you want to be inclusive of women.
In this case,
the manager's discomfort about changing pronouns
matters less than the career harm caused by them being exclusionary,
but many cases are not this clear cut.
Like any written rules,
a Code of Conduct requires constant interpretation;
like everything else,
discussion about specific cases becomes easier with practice.

## Summary {#inclusive-summary}

FIXME: create concept map for making an inclusive project

## Exercises {#inclusive-exercises}

FIXME: exercises for creating an inclusive project.

## Key Points {#inclusive-keypoints}

```{r, child="keypoints/inclusive.md"}
```

```{r, child="etc/links.md"}
```
