# Key Points {#s:keypoints}

## Introduction
```{r, child="keypoints/intro.md"}
```

## The Unix Shell
```{r, child="keypoints/shell.md"}
```

## Automating Analyses
```{r, child="keypoints/automate.md"}
```

## Syndicating Data
```{r, child="keypoints/syndicate.md"}
```

## Configuring Software
```{r, child="keypoints/configure.md"}
```

## Logging
```{r, child="keypoints/logging.md"}
```

## Unit Testing
```{r, child="keypoints/unit.md"}
```

## Verification
```{r, child="keypoints/verify.md"}
```

## A Branching Workflow
```{r, child="keypoints/branches.md"}
```

## Managing Backlog
```{r, child="keypoints/backlog.md"}
```

## Code Style and Review
```{r, child="keypoints/style.md"}
```

## Development Process
```{r, child="keypoints/process.md"}
```

## Continuous Integration
```{r, child="keypoints/integrate.md"}
```

## Working Remotely
```{r, child="keypoints/remote.md"}
```

## Other Tools
```{r, child="keypoints/tools.md"}
```

## Documentation
```{r, child="keypoints/docs.md"}
```

## Refactoring
```{r, child="keypoints/refactor.md"}
```

## Project Structure
```{r, child="keypoints/project.md"}
```

## Including Everyone
```{r, child="keypoints/inclusive.md"}
```

## Packaging
```{r, child="keypoints/package.md"}
```

## Publishing
```{r, child="keypoints/publish.md"}
```

## Teamwork
```{r, child="keypoints/teamwork.md"}
```

## Pacing
```{r, child="keypoints/pacing.md"}
```

## Finale
```{r, child="keypoints/finale.md"}
```
