# Key Points {#keypoints}

## Novice R {#keypoints-novice-r}

FIXME: fill in keypoints for novice R

## Novice Python {#keypoints-novice-py}

FIXME: fill in keypoints for novice Python

## RSE {#keypoints-rse}

### The Unix Shell
```{r, child="keypoints/bash.md"}
```

### Automating Analyses
```{r, child="keypoints/automate.md"}
```

### Syndicating Data
```{r, child="keypoints/syndicate.md"}
```

### Configuring Software
```{r, child="keypoints/configure.md"}
```

### Logging
```{r, child="keypoints/logging.md"}
```

### Unit Testing
```{r, child="keypoints/unit.md"}
```

### Verification
```{r, child="keypoints/verify.md"}
```

### A Branching Workflow
```{r, child="keypoints/branches.md"}
```

### Managing Backlog
```{r, child="keypoints/backlog.md"}
```

### Code Style and Review
```{r, child="keypoints/style.md"}
```

### Development Process
```{r, child="keypoints/process.md"}
```

### Continuous Integration
```{r, child="keypoints/integrate.md"}
```

### Working Remotely
```{r, child="keypoints/remote.md"}
```

### Other Tools
```{r, child="keypoints/tools.md"}
```

### Documentation
```{r, child="keypoints/docs.md"}
```

### Refactoring
```{r, child="keypoints/refactor.md"}
```

### Project Structure
```{r, child="keypoints/project.md"}
```

### Including Everyone
```{r, child="keypoints/inclusive.md"}
```

### Python Packaging
```{r, child="keypoints/py-package.md"}
```

### Publishing
```{r, child="keypoints/publish.md"}
```

### Teamwork
```{r, child="keypoints/teamwork.md"}
```

### Pacing
```{r, child="keypoints/pacing.md"}
```

### Finale
```{r, child="keypoints/finale.md"}
```
