# Learning Objectives {#objectives}

## Novice R {#objectives-novice-r}

FIXME: fill in learning objectives for novice R

## Novice Python {#objectives-novice-py}

FIXME: fill in learning objectives for novice Python

## RSE {#objectives-rse}

### The Unix Shell
```{r shell, child="objectives/shell.md"}
```

### Automating Analyses
```{r automate, child="objectives/automate.md"}
```

### Syndicating Data
```{r syndicate, child="objectives/syndicate.md"}
```

### Configuring Software
```{r configure, child="objectives/configure.md"}
```

### Logging
```{r logging, child="objectives/logging.md"}
```

### Unit Testing
```{r unit, child="objectives/unit.md"}
```

### Verification
```{r verify, child="objectives/verify.md"}
```

### A Branching Workflow
```{r branches, child="objectives/branches.md"}
```

### Managing Backlog
```{r backlog, child="objectives/backlog.md"}
```

### Code Style and Review
```{r style, child="objectives/style.md"}
```

### Development Process
```{r process, child="objectives/process.md"}
```

### Continuous Integration
```{r integrate, child="objectives/integrate.md"}
```

### Working Remotely
```{r remote, child="objectives/remote.md"}
```

### Other Tools
```{r tools, child="objectives/tools.md"}
```

### Documentation
```{r docs, child="objectives/docs.md"}
```

### Refactoring
```{r refactor, child="objectives/refactor.md"}
```

### Project Structure
```{r project, child="objectives/project.md"}
```

### Including Everyone
```{r inclusive, child="objectives/inclusive.md"}
```

### Packaging
```{r package, child="objectives/package.md"}
```

### Publishing
```{r publish, child="objectives/publish.md"}
```

### Teamwork
```{r teamwork, child="objectives/teamwork.md"}
```

### Pacing
```{r pacing, child="objectives/pacing.md"}
```

### Finale
```{r finale, child="objectives/finale.md"}
```
