# Packaging {#s:package}

## Objectives

```{r, child="objectives/package.md"}
```

> Another response of the wizards,
> when faced with a new and unique situation,
> was to look through their libraries to see if it had ever happened before.
> This was...a good survival trait.
> It meant that in times of danger you spent the day sitting very quietly in a building with very thick walls.
>
> -- Terry Pratchett

The more software you write,
the more you realize that a programming language is a way to build and combine software libraries.
Every widely-used language now has an online repository
from which people can download and install those libraries.
This lesson shows you how to use Python's tools to create and share libraries of your own.

This material is based in part on [Python 102][python-102] by [Ashwin Srinath][srinath-ashwin].

## How can I turn a set of Python source files into a module? {#s:package-modules}

Any Python source file can be imported by any other.
When a file is imported,
the statements in it are executed as it loads.
Variables defined in the file are then available as `module.name`.
(This is why Python files should be named using [`pothole_case`](glossary.html#pothole-case)
instead of [`kebab-case`](glossary.html#kebab-case):
an expression like `import some-thing` isn't allowed
because `some-thing` isn't a legal variable name.)

As an example,
we can put a constant and two functions used in our Zipf's Law study in a file called `zipf.py`:

```python
from pytest import approx


RELATIVE_ERROR = 0.05


def make_zipf(length):
    assert length > 0, 'Zipf distribution must have at least one element'
    result = [1/(1 + i) for i in range(length)]
    return result


def is_zipf(hist, rel=RELATIVE_ERROR):
    assert len(hist) > 0, 'Cannot test Zipfiness without data'
    scaled = [h/hist[0] for h in hist]
    perfect = make_zipf(len(hist))
    return scaled == approx(perfect, rel=rel)
```
{: title="package/01/zipf.py"}

<!-- == noindent -->
and then use `import zipf`,
`from zipf import is_zipf`,
and so on:

```python
from zipf import make_zipf, is_zipf

generated = make_zipf(5)
print('generated distribution: {}'.format(generated))
generated[-1] *= 2
print('passes test with default tolerance: {}'.format(is_zipf(generated)))
print('passes test with tolerance of 1.0: {}'.format(is_zipf(generated, rel=1.0)))
```
{: title="package/01/use.py"}

Running this program produces the following output:

```text
generated distribution: [1.0, 0.5, 0.3333333333333333, 0.25, 0.2]
passes test with default tolerance: False
passes test with tolerance of 1.0: True
```

<!-- == noindent -->
It also creates a sub-directory called `__pycache__`
that holds the compiled versions of the imported files.
The next time Python imports `zipf`,
it checks the timestamp on `zipf.py` and the timestamp on the corresponding file in `__pycache__`.
If the latter is more recent,
Python doesn't bother to recompile the file:
it just loads the bytes in the cached version and uses those.
To avoid confusing it,
we (almost) always put `__pycache__` in `.gitignore`.

## How can I control what is executed during import and what isn't? {#s:package-import}

Sometimes it's handy to be able to import code and also run it as a program.
For example,
we may have a file full of useful functions for extracting keywords from text
that we want to be able to use in other programs,
but also want to be able to run `keywords somefile.txt` to get a listing.

To help us do this (and other things we'll see later),
Python automatically creates a variable called `__name__` in each module.
If the module is the main program,
that variable is assigned the string `'__main__'`.
Otherwise, it is assigned the module's name.
Using this leads to modules like this:

```python
import sys
from pytest import approx


USAGE = '''zipf num [num...]: are the given values Zipfy?'''
RELATIVE_ERROR = 0.05


def make_zipf(length):
    ...as before...


def is_zipf(hist, rel=RELATIVE_ERROR):
    ...as before...


if __name__ == '__main__':
    if len(sys.argv) == 1:
        print(USAGE)
    else:
        values = [int(a) for a in sys.argv[1:]]
        result = is_zipf(values)
        print('{}: {}'.format(result, values))
    sys.exit(0)
```
{: title="package/02/zipf.py"}

<!-- used="package/02/use.py" -->

Here,
the code guarded by `if __name__ == '__main__'` isn't executed when the file loaded by something else.
We can test this by re-running `use.py` as before:
the usage message doesn't appear,
which means the main block wasn't executed,
which is what we want.

## How can I install a Python package? {#s:package-install}

The most common way to install Python packages is to use a tool called `pip`.
The command <code>pip install <em>package</em></code>
checks to see if the package is already installed (or needs to be upgraded);
if so,
it downloads the package from [PyPI][pypi] (the Python Package Index),
unpacks it,
and installs it.
Depending on where Python is installed,
completing that installation may require administrative privileges;
for example,
if Python is installed in `/usr/bin/python`,
you may need to run `sudo` to overwrite previously-installed libraries.
*This is a bad idea*,
since system tools may depend on particular versions of those packages,
and may break if you overwrite them.
Section \@ref(s:package-virtualenv) shows how to avoid these problems.

Since a project may depend on many packages,
developers frequently put a list of those dependencies in a file called `requirements.txt`.
`pip install -r requirements.txt` will then install the dependencies listed in that file.
(The file can be called anything,
but everyone uses `requirements.txt`,
so you should too.)
This file can just list package names,
or it can specify exact versions, minimum versions, etc.:

```text
request
scipy==1.1.0
tdda>=1.0
```

If you want to create a file like this,
`pip freeze` will print the exact versions of all installed packages.
You usually won't use this directly in `requirements.txt`,
since your project probably doesn't depend on all of the listed files,
but it's a good practice to save this in version control when producing reports
so that you can reproduce your results later
(Chapter \@ref(s:publish)).

## How can I create an installable Python package? {#s:package-package}

Packages have to come from somewhere,
and that "somewhere" is mostly developers like you.
Creating a Python package is fairly straightforward,
and mostly comes down to having the right directory structure.

A [package](glossary.html#package) is a directory that contains a file called `__init__.py`,
and may contain other files or sub-directories containing files.
`__init__.py` can contain useful code or be empty,
but either way,
it has to be there to tell Python that the directory is a package.
For our Zipf example,
we can reorganize our files as follows:

```text
+- use.py
+- zipf
    +- __init__.py
```

<!-- used="package/03/use.py" -->
<!-- used="package/03/zipf/__init__.py" -->

<!-- == noindent -->
`zipf/__init__.py` contains `RELATIVE_ERROR` and the functions we've seen before.
`use.py` doesn't change---in particular,
it can still use `import zipf` or `from zipf import is_zipf`
even though there isn't a file called `zipf.py` any longer.
When we run `use.py`,
it loads `zipf/__init__.py` when `import zipf` executes
and creates a `__pycache__` directory inside `zipf`.

## How can I manage the source code of large packages? {#s:package-large}

As our package grows,
we should split its source code into multiple files.
To show how this works,
we can put the Zipf generator in `zipf/generate.py`:

```python
def make_zipf(length):
    assert length > 0, 'Zipf distribution must have at least one element'
    result = [1/(1 + i) for i in range(length)]
    return result
```
{: title="package/04/zipf/generate.py"}

<!-- == noindent -->
and then import that file in `__init__.py`.

```python
import sys
from pytest import approx
from . import generate


RELATIVE_ERROR = 0.05


def is_zipf(hist, rel=RELATIVE_ERROR):
    assert len(hist) > 0, 'Cannot test Zipfiness without data'
    scaled = [h/hist[0] for h in hist]
    perfect = generate.make_zipf(len(hist))
    return scaled == approx(perfect, rel=rel)
```
{: title="package/04/zipf/__init__.py"}

We write that import as `from . import generate` to make sure that
we will get the `generate.py` file in the same directory as `__init__.py`
(the `.` means "current directory", just as it does in the Unix shell).

When we arrange our code like this,
the code that *uses* the library must refer to `zipf.generate.make_zipf`:

```python
import zipf

generated = zipf.generate.make_zipf(5)
print('generated distribution: {}'.format(generated))
generated[-1] *= 2
print('passes test with default tolerance: {}'.format(zipf.is_zipf(generated)))
print('passes test with tolerance of 1.0: {}'.format(zipf.is_zipf(generated, rel=1.0)))
```
{: title="package/04/use.py"}

## How can I distribute software packages that I have created? {#s:package-distribute}

People can always get your package by cloning your repository and copying files from that
(assuming your repository is accessible,
which is should be for published research),
but it's much friendlier to create something they can install.
For historical reasons,
Python has several ways to do this.
We will show how to use [setuptools][setuptools],
which is the lowest common denominator;
[conda][conda] is a modern does-everything solution,
but has larger startup overhead.

To use `setuptools`,
we must create a file called `setup.py` in the directory *above* the root directory of the package:

```text
+- setup.py
+- use.py
+- zipf
    +- __init__.py
    +- generate.py
```

<!-- == noindent -->
The file `setup.py` must have exactly that name,
and must contain these lines:

```python
from setuptools import setup, find_packages

setup(
    name='zipf',
    version='0.1',
    author='Greg Wilson',
    packages=find_packages()
)
```
{: title="package/05/setup.py"}

<!-- used="package/05/use.py" -->
<!-- used="package/05/zipf/__init__.py" -->
<!-- used="package/05/zipf/generate.py" -->

<!-- == noindent -->
The `name`, `version`, and `author` parameters to `setup` are self-explanatory;
you should modify the values assigned to them for your package;
the function `find_packages` returns a list of things worth packaging.

Once you have created this file,
you can run `python setup.py sdist` to create your package.
The verb `sdist` stands for "source distribution",
meaning that the source of the Python files is included in the package:

```shell
$ python setup.py sdist
```
```text
running sdist
running egg_info
creating zipf.egg-info
writing zipf.egg-info/PKG-INFO
writing dependency_links to zipf.egg-info/dependency_links.txt
writing top-level names to zipf.egg-info/top_level.txt
writing manifest file 'zipf.egg-info/SOURCES.txt'
reading manifest file 'zipf.egg-info/SOURCES.txt'
writing manifest file 'zipf.egg-info/SOURCES.txt'
warning: sdist: standard file not found: should have one of README, README.rst, README.txt, README.md

running check
warning: check: missing required meta-data: url

warning: check: missing meta-data: if 'author' supplied, 'author_email' must be supplied too

creating zipf-0.1
creating zipf-0.1/zipf
creating zipf-0.1/zipf.egg-info
copying files to zipf-0.1...
copying setup.py -> zipf-0.1
copying zipf/__init__.py -> zipf-0.1/zipf
copying zipf/generate.py -> zipf-0.1/zipf
copying zipf.egg-info/PKG-INFO -> zipf-0.1/zipf.egg-info
copying zipf.egg-info/SOURCES.txt -> zipf-0.1/zipf.egg-info
copying zipf.egg-info/dependency_links.txt -> zipf-0.1/zipf.egg-info
copying zipf.egg-info/top_level.txt -> zipf-0.1/zipf.egg-info
Writing zipf-0.1/setup.cfg
creating dist
Creating tar archive
removing 'zipf-0.1' (and everything under it)
```

<!-- == noindent -->
We will look at how to clean up the warnings about `README.md`, `url`, and `author_email` in the exercises.

`python setup.py sdist` creates a compressed file `dist/zipf-0.1.tar.gz` that contains the following:

```shell
$ tar ztvf dist/zipf-0.1.tar.gz
```
```text
drwxr-xr-x  0 pterry staff       0 20 Aug 15:36 zipf-0.1/
-rw-r--r--  0 pterry staff     180 20 Aug 15:36 zipf-0.1/PKG-INFO
-rw-r--r--  0 pterry staff      38 20 Aug 15:36 zipf-0.1/setup.cfg
-rw-r--r--  0 pterry staff     145 20 Aug 13:40 zipf-0.1/setup.py
drwxr-xr-x  0 pterry staff       0 20 Aug 15:36 zipf-0.1/zipf/
-rw-r--r--  0 pterry staff     317 20 Aug 13:34 zipf-0.1/zipf/__init__.py
-rw-r--r--  0 pterry staff     163 20 Aug 13:34 zipf-0.1/zipf/generate.py
drwxr-xr-x  0 pterry staff       0 20 Aug 15:36 zipf-0.1/zipf.egg-info/
-rw-r--r--  0 pterry staff       1 20 Aug 15:36 zipf-0.1/zipf.egg-info/dependency_links.txt
-rw-r--r--  0 pterry staff     180 20 Aug 15:36 zipf-0.1/zipf.egg-info/PKG-INFO
-rw-r--r--  0 pterry staff     154 20 Aug 15:36 zipf-0.1/zipf.egg-info/SOURCES.txt
-rw-r--r--  0 pterry staff       5 20 Aug 15:36 zipf-0.1/zipf.egg-info/top_level.txt
```

<!-- == noindent -->
The source files `__init__.py` and `generate.py` are in there,
along with the odds and ends that `pip` will need to install this package properly when the time comes.

## How can I manage the packages my projects need? {#s:package-virtualenv}

We want to test the package we just created,
but we *don't* want to affect the packages we already have installed,
and as noted earlier,
we may not have permission to write into the directory that contains system-wide packages.
(For example, we may be testing something out on a cluster shared by our whole department.)
The solution is to use a [virtual environment](glossary.html#virtual-environment).
These are slowly being superceded by more general solutions like [Docker][docker],
but they are still the easiest solution for most of us.

A virtual environment is a layer on top of an existing Python installation.
Whenever Python needs to find a library,
it looks in the virtual environment first.
If it can satisfy its needs there, it's done;
otherwise,
it looks in the underlying environment.
This gives us a place to install packages that only some projects need,
or that are still under development,
without affecting the main installation.

FIXME: figure

We can create and manage virtual environments using a tool called `virtualenv`.
To install it,
run `pip install virtualenv`.
Once we have done that,
we can create a new virtual environment called `test` by running:

```shell
$ virtualenv test
```
```text
Using base prefix '/Users/pterry/anaconda3'
New python executable in /Users/pterry/test/bin/python
Installing setuptools, pip, wheel...
done.
```

<!-- == noindent -->
`virtualenv` creates a new directory called `test`,
which contains sub-directories called `bin`, `lib`, and so on---everything
needed for a minimal Python installation.
Crucially,
`test/bin/python` checks for packages in `test/lib` *before* checking the system-wide install.

We can switch to the `test` environment by running:

```shell
$ source test/bin/activate
```

<!-- == noindent -->
`source` is a Unix shell command meaning "run all the commands from a file in this currently-active shell".
We use it because typing `test/bin/activate` on its own would run those commands in a sub-shell,
which would have no effect on the shell we're in.
Once we have done this,
we're running the Python interpreter in `test/bin`:

```shell
$ which python
```
```text
/Users/pterry/test/bin/python
```

We can now install packages to our heart's delight.
Everything we install will go under `test`,
and won't affect the underlying Python installation.
When we're done,
we can switch back to the default environment with `deactivate`.
(We don't need to `source` this.)

Most developers create a directory called `~/envs`
(i.e., a directory called `envs` directly below their home directory)
to store their virtual environments:

```shell
$ cd ~
$ mkdir envs
$ which python
```
```text
/Users/pterry/anaconda3/bin/python
```
```shell
$ virtualenv envs/test
```
```text
Using base prefix '/Users/pterry/anaconda3'
New python executable in /Users/pterry/envs/test/bin/python
Installing setuptools, pip, wheel...done.
```
```shell
$ which python
```
```text
/Users/pterry/anaconda3/bin/python
```
```shell
$ source envs/test/bin/activate
```
```text
(test)
```
```shell
$ which python
```
```text
/Users/pterry/envs/test/bin/python
(test)
```
```shell
$ deactivate
```
```shell
$ which python
```
```text
/Users/pterry/anaconda3/bin/python
```

Notice how every command now displays `(test)` when that virtual environment is active.
Between Git branches and virtual environments,
it can be very easy to lose track of what exactly you're working on and with.
Having prompts like this can make it a little less confusing;
using virtual environment names that match the names of your projects
(and branches, if you're testing different environments on different branches)
quickly becomes essential.

## How can I test package installation? {#s:package-test-install}

Now that we have a virtual environment set up,
we can test the installation of our Zipf package:

```shell
$ pip install ./src/package/05/dist/zipf-0.1.tar.gz
```
```text
Processing ./src/package/05/dist/zipf-0.1.tar.gz
Building wheels for collected packages: zipf
  Running setup.py bdist_wheel for zipf ... done
  Stored in directory: /Users/pterry/Library/Caches/pip/wheels/6b/de/80/d72bb0d6e7c65b6e413f0cf10f04b4bbccb329767853fe1644
Successfully built zipf
Installing collected packages: zipf
Successfully installed zipf-0.1
(test)
```
```shell
$ python
>>> import zipf
>>> zipf.RELATIVE_ERROR
```
```text
0.05
```
```shell
$ pip uninstall zipf
```
```text
Uninstalling zipf-0.1:
  Would remove:
    /Users/pterry/envs/test/lib/python3.6/site-packages/zipf-0.1.dist-info/*
    /Users/pterry/envs/test/lib/python3.6/site-packages/zipf/*
Proceed (y/n)? y
  Successfully uninstalled zipf-0.1
(test)
```

Again,
one environment per project and one project per environment might use more disk space than is absolutely necessary,
but it will still be less than most of your data sets,
and will save a lot of debugging.

## Announcing Work {#s:package-announce}

FIXME: <https://medium.com/indeed-engineering/marketing-for-data-science-a-7-step-go-to-market-plan-for-your-next-data-product-60c034c34d55>

## Summary {#s:package-summary}

FIXME: create concept map for packages

## Exercises {#s:package-exercises}

### Clean up warning messages

-   FIXME: clean up warning messages from `python setup.py sdist`

## Key Points

```{r, child="keypoints/package.md"}
```

```{r, child="etc/links.md"}
```
