# Development Process {#s:process}

## Objectives

```{r, child="objectives/process.md"}
```

A [software development process](glossary.html#dev-process) is
the steps a team goes through to create, deliver, and maintain software.
Broadly speaking,
software development processes can be divided into three groups:

-   Chaotic: everyone's doing something,
    but there's no overall plan or consistency.
-   An [agile](glossary.html#agile) process
    based on lots of short steps with frequent feedback and course correction.
-   A [sturdy](glossary.html#sturdy) process
    that invests a lot of effort in planning out work.
    (This label is made up:
    historically,
    planning-based approaches were described first and didn't have a particular name,
    but then agile came along and we needed one.)

Most self-taught teams start with a chaotic process.
This material is meant to help you move to something better.

## What is agile development? {#s:process-agile}

Agile development starts from three related premises:

-   You *can't* plan a software project very far in advance
    because requirements and technology are constantly changing.
-   You *shouldn't* try to plan everything in advance
    because you're going to learn what's possible,
    and your users are going to learn what they actually want,
    as the work progresses.
-   You can *afford* not to lock yourself into a long-term plan
    because software is much more malleable than concrete or steel.

This approach came into prominence with the rise of the web in the 1990s:

1.  The web made it possible to release software weekly, daily, or even hourly,
    since updating a server is a lot faster,
    and a lot less expensive,
    than shipping CDs to thousands of people.
2.  Multi-year development plans don't make a lot of sense
    when everything they depend on  will be obsolete by the time work starts,
    much less by the time it finishes.
3.  The growth of the web was aided by, and fuelled, the growth of the open source movement.
    People couldn't help noticing that most open source projects didn't have long-range plans,
    but nevertheless produced high-quality software faster than many closed-source commercial projects.

## What are the feedback loops in agile development? {#s:process-feedback}

At its core,
agile development relies on continuous feedback.
Agile methods break development down into short [iterations](glossary.html#iteration),
typically no more than two weeks long,
and often as short as a single day.
In each iteration,
the team decides what to build next,
then designs it, builds it, tests it, and delivers it.
Feedback loops at several scales help them get this right and do it better next time.

The iteration itself is the primary feedback loop.
Users often don't know what they want until they see it,
so short cycles are a way to avoid spending too much time building what turns out to be the wrong thing.
Short iterations help improve efficiency in two other ways as well.
First,
most people can keep track of what they're doing for a few days at a time without elaborate Gantt charts,
so short cycles allow them to spend proportionally less time coordinating with one another.
Second, finding bugs becomes easier:
instead of looking through weeks' or months' worth of software to find out where the problem is,
developers usually only have to look at what's been written in the last few days.

A typical working day starts with a [stand-up meeting](glossary.html#stand-up-meeting)
where everyone in the team reports what they did since the last meeting,
what they're planning to do next,
and what's blocking them (if anything).
(It's called a "stand-up" meeting because it's usually held standing up,
which encourages people to stay focused.)
For example,
my report earlier this week was:

-   Yesterday:
    fixed a bug in citation handling;
    got the Shiny example to update properly;
    added slides to the tutorial on loading data in Shiny applications.
-   Today:
    update the script that checks for unused bibliography entries;
    add tests for handling missing data files in the Shiny tutorial.
-   Blockers:
    should we report "no permission" separately from "missing file" in the Shiny tutorial?

Stand-up meetings are another agile feedback loop.
Each day,
the team gets feedback on the progress they're making,
whether they're still on track to meet the iteration's goals,
whether the technical decisions they're making are paying off, and so on.
The key to making this work is that each task is at most a day long.
Anything longer is broken into sub-tasks so that there's something to report at every meeting.
Without this rule,
it's all too easy for someone to say, "Still working on X," several days in a row,
so feedback and the possibility of early course correction are lost.

Once the stand-up meeting is over, everyone gets back to work.
In many agile teams,
this means sitting with a partner and doing [pair programming](glossary.html#pair-programming).
One person, called the driver, does the typing,
while the other person, called the navigator,
watches and comments;
every hour or so, the pair switch roles.

Pair programming is beneficial for several reasons.
First,
the navigator will often notice mistakes in the driver's code,
or remember design decisions that the driver is too busy typing to recall.
This is the tightest of the feedback loops that make agile work,
since feedback is nearly continuous.

Second,
pair programming spreads knowledge around:
every piece of code has been seen by at least two people,
which reduces the risk of "But I didn't know" mistakes.
It also helps people pick up new skills:
if you have just seen someone do something with two clicks,
you will probably do it that way when it's your turn to drive
rather than spending two minutes doing it the way you always have.
And finally,
most people are less likely to check Facebook every five minutes if someone else is working with them...

As well as pair programming, most agile teams use two other practices.
[Test-driven development](glossary.html#tdd) (TDD),
discussed in Chapter \@ref(s:unit),
is the practice of writing unit tests *before* writing application code.
The second is [continuous integration](glossary.html#continuous-integration).
Every time someone commits code to the version control repository,
an automated process checks out a clean copy of the code,
builds it,
runs all the tests,
and posts the results somewhere for the whole team to see.
If any of the tests fail,
the continuous integration system notifies people by sending out email or texting them.

We will explain how to implement continuous integration in Chapter \@ref(s:integrate).
From a developer's point of view,
its key benefit is that it ensures the project is always in a runnable state.
This may not seem important if you don't have to hand it in until next week,
but is critical as soon as two or more people are involved in development.
It's very frustrating (and unproductive) for someone to be blocked
because someone else has broken something but not realized it.

The final key practice in agile development is the [post-mortem](glossary.html#post-mortem).
At the end of every sprint,
the team should get together and ask themselves what went well,
what could be improved,
and what new things they want to try (if any).

## What are the risks associated with agile development? {#s:process-agile-risks}

The biggest risk associated with agile is that
people can use the term to excuse a process that's still actually chaotic.
Most people don't like writing plans before they code or documenting what they've done;
coincidentally,
agile doesn't require them to do much of either.
It's therefore all too common for developers to say "we're agile"
when what they mean is "we're not going to bother doing anything we don't want to".
In reality,
agile requires *more* discipline, not less,
just as improvising well requires even more musical talent than playing a score exactly.

At the same time,
many people don't like having to make decisions:
after two decades of schooling,
they want to be told what the assignment is and exactly what they have to do
to get an 'A', a 'B', or whatever grade they're shooting for.
Some become quite defensive when told that figuring out what to do is now part of their job,
but that's as essential to agile development as it is to scientific research.

Finally, FIXME: customer buy-in.

## What is sturdy development? {#s:process-sturdy}

As we said in the introduction,
the alternative to agile development doesn't have a widely-accepted name;
we call it "sturdy",
since "traditional", "old-fashioned", and "process-heavy" all sound pejorative.

Whatever it's called,
it's key feature is planning work in advance.
If you're going to spend three days driving across the country,
it makes sense to spend half an hour figuring out a good route.
Equally,
if you're going to spend several months building a complex piece of software,
*and you know what the final result is supposed to look like*,
it makes sense to spend some time figuring out what you're going to do
and how long it ought to take.
In order to explain how to do this systematically,
we need to look at two roles in large software projects:
the product manager and the project manager.

## What is the role of a product manager? {#s:process-product}

The [product manager](glossary.html#product-manager) is
the person responsible for the software's feature list.
Typically,
while developers are building Version N,
she is talking to users in order to find out what should go into Version N+1.
She doesn't ask them what features they want,
because if she does,
what she'll get is a mish-mash of buzzwords plucked from
the lead paragraphs of popular articles.
(Blockchain! Blockchain all the things!)

Instead, she asks, "What can't you do right now that you want to?",
"What do you find irritating in the current product?",
and, "Why are you using other software instead of ours?"
She then translates the answers into a list of changes
to be considered for Version N+1.
The product manager also talks to the software's developers
to find out what *they* don't like about the current software
and adds their wishes to the pile.
Typically,
these are things like
"reorganize the database interface" and "clean up the build process".

So, it's Monday morning.
Version N shipped last Thursday
(because you should never, ever ship on a Friday).
the team has had a weekend to catch its collective breath
and is ready to start work once again.
(If people are so burned out from the previous round of work
that they need a whole week to recover,
read the discussion of crunch mode in Chapter \@ref(s:pacing).)
At this point,
the product manager divides up the list of desired features
and assigns a few to each developer.
They then have some time---typically a few days
to a couple of weeks---to do a little research,
write some throwaway prototype code,
and most importantly to think.
How could this feature be implemented?
Is there a quick-and-dirty alternative that would only deliver half of what was asked for
but could be done in a tenth of the time?
What impact will each alternative have on the installation process?
And how will the new feature be tested?

This process is called [analysis and estimation](glossary.html#ae) (A&E).
The result is a collection of short proposals,
each typically half a page to half a dozen pages long.
There's no set form for these,
but they usually include whatever background information a well-informed developer is unlikely to already know,
a discussion of the alternatives,
lessons learned from any prototyping that was done,
and most importantly,
an estimate of how much time would be needed to build each alternative.
This time includes estimates from QA (for testing),
the technical writer (for documenting),
the people responsible for the build and creating the installer,
and so on.

So now it's Monday morning again.
Three weeks have gone by and all the A&E's are done.
When the time estimates are totalled,
they come to 700 developer-days.
Unfortunately, there are only 240 available:
the size of the team is fixed,
and the next release has to be available in May.
*This is normal:*
there is *never* enough time to add everything that everyone wants.

What the product manager does now is draw a 3x3 grid.
The X axis is labeled "effort", the Y axis, "importance",
and each axis is divided into "low", "medium", and "high".
(Finer-grained divisions, such as a 1--10 scale, add no value:
nobody has an algorithm for distinguishing priority 6 items from priority 7 items,
and anyway, the grid is just to get conversation going.)
Each feature's name now goes in one of the nine grid cells,
and the product manager steps out of the way.

## What is the role of a project manager? {#s:process-project}

The [project manager](glossary.html#project-manager)'s role is
to turn priorities and effort estimates into a schedule
and then ensure that the schedule is met.
First,
the high-effort, low-importance features in the grid are crossed off:
they're simply not worth doing.
The project manager may decide to tackle some low-effort, high-value items first
in order to ensure that if things go badly wrong,
something useful will still have been accomplished
Alternatively,
she may decide to go after some high-effort items at the start,
on the theory that if they aren't tackled early,
they'll keep getting pushed off.

Either way,
the items on the diagonal are the ones that have to be argued over.
Should the team tackle Feature 14 (high effort, high importance)
or Features 18, 19, and 22 (individually lower importance, but the same total effort)?
It can take rounds of discussion to sort this out;
what's important is that people don't tweak effort estimates in order to squeeze things in,
because if they do,
the people doing the estimating will start padding their numbers in self defense.
Since project managers aren't stupid,
they'll shave the estimates even more,
so developers will add even more padding,
and pretty soon the whole thing becomes science fiction.

## What are the risks associated with sturdy development? {#s:process-sturdy-risks}

The hardest step in this process for beginners is coming up with time estimates.
How can you possibly guess how long it will take to write a database interface
if you've never used one before?
There are two answers:

-   You're not expected to pull an number out of thin air
    (at least, not by managers who know what they're doing).
    Instead,
    you should budget enough time to write some throwaway code,
    or download and try out an open source tool,
    in order to get a feel for it.
-   You've had to learn other new technologies before.
    A guess based on that experience might be off by a factor of two or three,
    but it probably won't be off by a factor of ten.
    Even if it is,
    it's better than no guess at all,
    provided everyone involved knows that it's a guess.
    (You will meet people who will be very critical every time one of your estimates is wrong.
    In my experience,
    they are no better at estimating than anyone else.)

The more estimating you do, the better you'll get,
and the better you are,
the more smoothly your projects will run.
If you're an undergraduate,
your project will probably have to fit in one or two terms.
You will therefore probably be asked to go around the planning loop once or twice,
which in turn determines how much you'll be expected to deliver in each iteration.
This is called [time boxing](glossary.html#time-boxing):
you specify how long a cycle will last,
then see how much work you can fit into that interval.

The alternative is [feature boxing](glossary.html#feature-boxing):
decide what you want to do,
then build a schedule that gives you enough time to do it.
Time boxing generally works better
even in projects that don't have to line up with semester boundaries,
since it encourages developers to take smaller steps,
and allows them to give customers more frequent demos
(which serve as course corrections).

## Why, when, and how should our team start to cut corners? {#s:process-corners}

Contrary to popular belief,
a schedule's primary purpose is not to tell you what you're supposed to be doing on any given day.
Instead,
it is is to tell you when you should start cutting corners.
Suppose,
for example,
that you have ten weeks in order to accomplish some task.
Five weeks after you start,
you've only done the first third of work.
You have several options:

Ignore the problem.
:   This is very popular, but doesn't actually solve the problem.

Work longer hours.
:   This is also very popular,
    but as Chapter \@ref(s:pacing) explains,
    it is self-defeating.

Enlarge the team.
:   This is a good strategy if you're peeling carrots,
    but usually doesn't work on other kinds of projects
    because it takes more time to bring someone up to speed
    than it does to do the work yourself.

Move the deadline back.
:   Product groups and research teams often do this
    (usually in combination with the previous solution),
    but it usually isn't an option for course projects.
    Instructors have to submit marks at the end of the term,
    and conference deadlines are fixed as well;
    sometimes,
    whatever hasn't been done by the due date
    might as well not be done at all.

Cut corners.
:   The best approach is to update the schedule
    to reflect the rate at which you're *actually* working,
    then drop features that it tells you can't be finished in time.

Let's go back to our example.
At the start of the project you believed it would take ten weeks.
You're now at week five,
but you've done only the first four weeks' worth of work,
so your estimates for how long tasks would take
were too optimistic by about 25%.
You should therefore go back to your schedule and add 1/4 to each task's estimate.
That inevitably means that some of the things you originally planned to do
now spill off the end of your ten-week window.
That's OK;
it's a shame you won't get to them,
but at least you can start taking action now
rather than trying to recover from a disaster.

People will thank you for this.
"I'm sorry, we're not going to have the frobnosticator for May 1"
is OK in January,
since it gives whoever was counting on the frobnosticator time to make other plans.
It is *not* OK on April 30;
neither is saying that it's "done", but full of bugs.

These calculations are the responsibility of the project manager.
Her job is to make sure everyone is doing what they're supposed to be doing,
to shield the team from interruptions (there are *always* interruptions),
and to track the team's progress.
She periodically compares how much has actually been done
with how much was supposed to be done and adjusts plans accordingly.

Taking scheduling seriously is one of the things that distinguishes good teams from bad ones.
It's unfortunate that most students only get to do it once or twice in their courses,
since you only really see the benefits with practice,
but even a couple of rounds of practice can make a big difference.

## Summary {#s:process-summary}

FIXME: summarize process

## Key Points

```{r, child="keypoints/process.md"}
```

```{r, child="etc/links.md"}
```
