# Project Structure {#project}

## Questions {#project-questions}

```{r questions, child="questions/project.md"}
```

## Objectives {#project-objectives}

```{r objectives, child="objectives/project.md"}
```

## Introduction {#project-intro}

Project organization is like a diet:
there is no such thing as "no diet",
just a good one or a bad one.
Similarly, there is no such thing as "no project organization":
your project is either organized well or poorly.
As with coding style (Chapter \@ref(style)),
small pieces in predictable places with readable names are easier to find and use
than large chunks that vary from project to project
and have names like `stuff`.

## What are Noble's Rules? {#project-noble}

@Nobl2009 described a way to organize small bioinformatics projects
that is equally useful for other kinds of research computing.
Each project is put in a separate Git repository,
and the directories in the root of this repository are organized according to purpose.
The original specification included five top-level directories:

-   The `./src/` directory (short for "source") holds source code
    for programs written in languages like C or C++ that need to be compiled.
    Many projects don't have this directory
    because all of their code is written in languages that don't need compilation.
-   Runnable programs go in `./bin/` (an old Unix abbreviation for "binary", meaning "not text").
    This includes the compiled and runnable versions of C and C++ programs,
    and also shell scripts,
    Python or R programs,
    and everything else that can be executed.
-   Raw data goes in in `./data/` and is never modified after being stored.
-   Results are put in `./results/`.
    This includes cleaned-up data,
    figures,
    and everything else that can be rebuilt using what's in `./bin/` and `./data/`.
    If intermediate results can be re-created quickly and easily,
    they might not be stored in version control,
    but anything that is included in a manuscript should be here.
-   Finally,
    documentation and manuscripts go in `./doc/`.

```{r project-noble, echo=FALSE, fig.cap="Project Layout"}
if (knitr::is_latex_output()) {
  knitr::include_graphics("figures/project/noble.pdf")
} else {
  knitr::include_graphics("figures/project/noble.svg")
}
```

Figure \@ref(fig:project-noble) below shows this layout for a project called `g-trans`.
A few things to notice are:

-   The documentation for the `regulate` script appears in the root of `./doc/`,
    while the paper for JCMB is stored in a sub-directory,
    since it contains several files.
-   The `./src/` directory contains a Makefile to re-build the `regulate` program (Chapter \@ref(automate)).
    Some projects put the Makefile in the root directory,
    reasoning that since it affects both `./src/` and `./bin/`,
    it belongs above them both rather than in either one.
-   There are several sub-directories underneath `./data/` and `./results/`,
    which we will discuss in [the next section](#project-filenames).
    Each of the sub-directories in `./results/` has its own Makefile,
    which will re-create the contents of that directory.

## How should files and sub-directories be named? {#project-filenames}

While the directories in the top level of each project are organized by purpose.
the directories within `./data/` and `./results/` are organized chronologically
so that it's easy to see when data was gathered
and when results were generated.
These directories all have names in [ISO date format](glossary.html#iso-date-format) like `YYYY-MM-DD`
to make it easy to sort them chronologically.
This naming is particularly helpful when data and results are used in several reports.

At all levels,
filenames should be easy to match with simple shell wildcards.
For example,
a project might use <code><em>species</em>_<em>organ</em>_<em>treatment</em>.csv</code>
as a file-naming convention,
giving filenames like `human_kidney_cm200.csv`.
This allows `human_*_cm200.csv` to match all human organs
or `*_kidney_*.csv` to match all kidney data.
It does produce long filenames,
but [tab completion](glossary.html#tab-completion) means you only have to type them once.
Long filenames are just as easy to match in programs:
Python's `glob` and R's `Sys.glob` will both take a pattern and return a list of matching filenames.

## How should I manage a mix of compiled programs and scripts? {#project-scripts}

Noble's Rules puts everything runnable in a single directory called `./bin/`.
That makes things easy to find,
but most software engineers would say that only the source code for programs should be in version control,
not the output of the compiler.
There are three ways to handle this:

1.  Put the compiled program under version control.
    In theory this makes research more [reproducible](glossary.html#reproducible-research),
    since anyone who wants to re-run the analysis can be sure they're using exactly the same program as the author,
    but in practice,
    many "compiled" programs load libraries dynamically,
    so their results can still be affected by changes to their environment.
    Putting the compiled programs in version control *does* make it easier for people to re-run analyses,
    though,
    since they don't need to be able to compile things themselves.

2.  Put everything in `./bin/`,
    then edit `.gitignore` to tell Git to ignore the compiled programs.
    This keeps everything runnable in one place,
    but requires a little bit of extra bookkeeping.

3.  Put compiled programs in `./bin/` and everything that doesn't require compilation in `./scripts/`.
    This makes management simpler (just ignore everything in `./bin/`),
    but means there are two places where runnable programs might live.

We usually go with the third option,
but they're all good as long as you are consistent between projects
and [document your rule](#project-boilerplate).

## Should I separate documentation from manuscripts? {#project-docs}

Noble's Rules place documentation and manuscripts in `./docs/`.
As with compiled programs and scripts,
some people separate these,
so that (for example) `./docs/` has the project's website and the documentation for its software,
while `./reports/` has one sub-directory for each paper, thesis chapter, or other manuscript.
As more researchers [work in the open(#g:open-science),
and as tools like [R Markdown][r-markdown] and the [Jupyter Notebook][jupyter]
blur the distinction between software, documentation, and reports,
separating the two makes less sense.

## How should I handle data can't be stored in version control? {#project-big-data}

Small datasets that don't contain sensitive information can and should be stored in version control.
As a rule of thumb,
anything that you would send in an email attachment is probably small enough to be put into Git,
while anything that might reveal someone's identity when combined with other data sets should not be.

If data is large or sensitive,
there should still be something in `./data/` to show its existence,
and that "something" should be easy for programs to read.
One option is a CSV file whose columns are:

-   the name of the dataset,
-   its URL or other unique identifier,
-   the date it was last checked, and
-   its size (so that users will have some idea of how much work is involved in processing it).

Another option is to have one file per dataset,
so that instead of reading `human_genome.bam`,
the program reads `human_genome.yml` and then uses the `url` key in that file to find the data it actually wants.
If your data is complicated,
you may also want to include a `README.md` file in `./data/` modelled on the one that accompanies
The Pudding article "[Women's Pockets are Inferior][womens-pockets]"
(see Chapter \@ref(publish)).

## What other files should every project contain? {#project-boilerplate}

Most projects' repositories contain a few files that weren't mentioned in Noble's paper,
but which have become conventional in open source projects.
All of these files may be plain text or Markdown,
and may have a `.txt` or `.md` suffix (or no suffix at all),
but please use the principal names given,
in all caps,
since a growing number of tools expect them.

-   `README`:
    the project's title and a one-paragraph description of its purpose or content.
    This file is displayed by GitHub Pages as the project's home page.

-   `LICENSE`:
    the project's license (discussed in Chapter \@ref(inclusive)).

-   `CONDUCT`:
    its code of conduct
    (also discussed in Chapter \@ref(inclusive)).

-   `CITATION`:
    how the work should be cited.
    This file usually contains a plain text citation,
    and may also include entries formatted for various bibliographic systems like [BibTeX][bibtex].
    Some projects also have a separate `CONTRIBUTORS` file listing everyone who has contributed to it,
    while others include this as a section in `CITATION`.

If setting up the project or contributing to it are complex,
there may also be a file called `BUILD` that explains how to install the required software,
the formatting conventions for new data entries,
and so on.
These instructions can also be included as a section in `README`.
Whichever convention is used,
remember that the easier it is for people to get set up and contribute,
the more likely they are to do so @Stei2014.

## What *is* a project? {#project-thinking}

Like features (Chapter \@ref(branches)),
what exactly constitutes a "project" requires a bit of judgment,
and different people will make different decisions.
Some common criteria are one project per publication,
one project per deliverable piece of software,
or one project per team.
The first tends to be too small:
a good data set will result in several reports,
and the goal of some projects is to produce a steady stream of reports (such as monthly forecasts).
The second is a good fit for software engineering projects
whose primary aim is to produce tools rather than results,
but is inappropriate for most data analysis work.
The third tends to be too large:
a team of half a dozen people may work on many different things at once,
and a repository that holds them all quickly looks like someone's basement.

The best rule of thumb we have found for deciding what is and isn't a project
is to ask what you have meetings about.
If the same set of people need to get together on a regular basis to talk about something,
that "something" probably deserves its own repository.
And if the list of people changes slowly over time but the meetings continue,
that's an even stronger sign.

## Summary {#project-summary}

FIXME: create concept map for project structure

## Exercises {#project-exercises}

FIXME: exercises for structure chapter.

## Key Points {#project-keypoints}

```{r keypoints, child="keypoints/project.md"}
```

```{r links, child="etc/links.md"}
```
