---
title: "Merely Useful: RSE Python"
author: "Madeleine Bonsma-Fisher, Kate Hertweck, Damien Irving, Luke Johnston, Christina Koch, Sara Mahallati, Joel Ostblom, Elizabeth Wickes, Charlotte Wickham, and Greg Wilson"
documentclass: book
bibliography: book.bib
citation-package: none
link-citations: yes
github-repo: merely-useful/merely-useful.github.io
site: bookdown::bookdown_site
date: "`r Sys.Date()`"
lot: yes
lof: yes
---

# Introduction {#intro}

```{r intro-setup, include=FALSE}
source(here::here("_common.R"))
```

> It's still magic even if you know how it's done.
>
> — Terry Pratchett

Software is now as essential to research as telescopes, test tubes, and reference libraries,
which means that researchers need need to know how to build, check, use, and share programs.
However,
most introductions to programming focus on developing commercial applications,
not on exploring problems whose answers aren't yet known.
Our goal is show you how to do that,
both on your own and as part of a team.

We believe every researcher should know
how to write short programs that clean and analyze data in a reproducible way,
and how to use version control to keep track of what they have done.
But just as some astronomers spend their careers designing telescopes,
some researchers focus on building the software that makes research possible.
People who do this are called [research software engineers][rse];
the aim of this book is to get you ready for this role,
i.e.,
to help you go from writing code for yourself
to creating tools to help your entire field advance.

All of this material can be freely re-used
under the terms of the Creative Commons–Attribution License (CC-BY 4.0);
please see Appendix \@ref(license) for details.
The source for the book lives in a public Git repository;
corrections and additions are very welcome,
and everyone whose work is included will be credited in the acknowledgments.

## What's the big picture? {#intro-big-picture}

Our approach to research software engineering is based on three related concepts:

-   [Open science][open-science]
    focuses on making data, methods, and results freely available to all
    by publishing them under [open licenses][open-license].
-   [Reproducible research][reproducible-research]
    means ensuring that anyone with access to data and software can feasibly reproduce results,
    both to check them and to build on them.
-   Software is [sustainable][sustainable-software]
    if it's easier for people to maintain it and extend it than to replace it.
    However,
    sustainability isn't just a property of the software:
    it also depends on the skills and culture of its users.

People often conflate these three ideas,
but they are distinct.
For example,
if you share your data and the programs that analyze it,
but don't document what steps to take in what order,
your work is open but not reproducible.
Conversely,
if you completely automate your analysis,
but your data is only available to people in your lab,
your work is reproducible but not open.
Finally,
if a software package is being maintained by a couple of post-docs
who are being paid a fraction of what they could earn in industry
and have no realistic hope of promotion because their field doesn't value tool building,
then sooner or later it will become [abandonware][abandonware],
at which point openness and reproducibility become moot points.

Nobody argues that research should be irreproducible or unsustainable,
but "not against it" and actively supporting it are very different things.
Academia doesn't yet know reward people for writing useful software,
so while you may be thanked,
the effort you put in may not translate into job security or decent pay.

And some people still worry that if they make their data and code generally available,
someone else will use it and publish a result they have come up with themselves.
This is almost unheard of in practice,
but that doesn't stop it being used as a scare tactic.
Other people are afraid of looking foolish or incompetent by sharing code that might contain bugs.
This isn't just [impostor syndrome][impostor-syndrome]:
members of marginalized groups are frequently judged more harshly than others,
so being wrong in public is much riskier for them.

## Who are these lessons for? {#intro-personas}

Amira Khan
:   completed a master's in library science five years ago
    and has since worked for a small aid organization.
    She did some statistics during her degree,
    and has learned some R and Python by doing data science courses online,
    but has no formal training in programming.
    Amira would like to tidy up the scripts, data sets, and reports she has created
    in order to share them with her colleagues.
    These lessons will show her how to do this and what "done" looks like.

Jun Hsu
:   completed an [Insight Data Science][insight] fellowship last year after doing a PhD in Geology
    and now works for a company that does forensic audits.
    He uses a variety of machine learning and visualization packages,
    and would now like to turn some of his own work into an open source project.
    This book will show him how such a project should be organized
    and how to encourage people to contribute to it.

Sami Virtanen
:   became a competent programmer during a bachelor's degree in applied math
    and was then hired by the university's research computing center.
    The kinds of applications they are being asked to support
    have shifted from fluid dynamics to data analysis;
    this guide will teach them how to build and run data pipelines
    so that they can pass those skills on to their users.

### Prerequisites

Readers should already be using Python regularly for data analysis,
and should be comfortable reading data from files
and writing loops, conditionals, and functions.

Learners will need:

-   a personal computer with Internet access;
-   a [Unix shell][unix-shell]
    (which is available on recent versions of Windows as well as Mac and Linux);
-   [Python 3][python] with [Jupyter][jupyter] and various scientific libraries; and
-   a [GitHub][github] account.

Please see Appendix \@ref(install) for instructions on how to set all of this up.

## Syllabus {#intro-syllabus}

This book uses data analysis as a motivating example,
and assumes that the learner's goal is to answer questions
rather than deliver commercial software products.
The data analysis task that we focus on
relates to a fascinating result in the field of quantitative linguistics.
[Zipf's Law][zipfs-law] states that the second most common word in a body of text
appears half as often as the most common,
the third most common appears a third as often, and so on.
To test Zipf's Law,
we analyze the distribution of word frequencies
in a collection of classic English novels
that are freely available from [Project Gutenberg][project-gutenberg].

Appendix \@ref(install) explains
how to download the novels and other files you will need.
When you are done,
you should have a directory (also called a [folder][folder])
containing a single sub-directory called `data`
with the following contents:

```text
└── data
    ├── README.md
    ├── dracula.txt
    ├── frankenstein.txt
    ├── jane_eyre.txt
    ├── moby_dick.txt
    ├── sense_and_sensibility.txt
    ├── sherlock_holmes.txt
    └── time_machine.txt
```

In the process of writing and publishing a Python package to verify Zipf's Law,
we will show you how to:

-   Use the Unix shell to efficiently manage your data and code.
-   Write Python programs that can be run at the command line.
-   Use Git and GitHub to track and share your work.
-   Use Make to automate complex workflows.
-   Enable users to configure your software without modifying it directly.
-   Find, handle, and fix errors in your code.
-   Work productively in a small team where everyone is welcome.
-   Write and review code to make it readable as well as correct.
-   Organise small and medium-sized data science projects.
-   Create Python packages that can be installed in standard ways.
-   Test your software and know which parts have not yet been tested.
-   Publish your code and research in open and reproducible ways.

## Acknowledgments {#intro-ack}

This book owes its existence to
everyone we met through [the Carpentries][carpentries].
We are also grateful to [Insight Data Science][insight] for sponsoring the early stages of this work,
to the authors of [@Nobl2009; @Hadd2010; @Wils2014; @Scop2015; @Tasc2017; @Wils2017; @Brow2018; @Deve2018; @Shol2019; @Wils2019],
and to everyone who has contributed, including:

-   Jonathan Dursi
-   Brandeis Marshall

```{r, child="./links.md"}
```
