# Version Control {#py-version-control}

## Questions {#py-version-control-questions}


## Objectives {#py-version-control-objectives}

```{r, child="objectives/py-version-control.md"}
```

## Exercises

1. Abbie has started working on a data analysis project. She has a copy of the main data set, and sent her supervisor a preliminary report based on a script she has been writing. The response to her report was a lot of comments, and an updated version of the dataset. Meanwhile, she had a colleague look at her script and he refactored it into a more conventional coding standard. 
Where are some areas where Rhonda could use version control strategies to track her work? What strategies could she use? 

1. Which of these is/are examples of remote storage? 
* GitHub
* Google Drive
* an external hard drive

1. Make a change to (file) and commit the changes to the repository. How can you check that the changes were committed? 

1. [TBD: Log output of previous commits]
What command would he use to look at the difference between the repository in its current state and when the commit happened? 

1. Manuel has a copy of his main repository on his laptop, work computer, and Github. If he has done work on his laptop while traveling, what steps does he need to go through to update the copy on his work computer?  What should he have done before starting to work on his laptop to make this go as smoothly as possible? 

[TBD: messages with a merge conflict]
What does this message indicate about the state of the repository? What would you do next? 

## Key Points {#py-version-control-keypoints}

```{r, child="keypoints/py-version-control.md"}
```

