# Joining data sets {#r-joins}

## Questions {#r-joins-questions}

```{r, child="questions/r-joins.md"}
```

```{r, message = FALSE}
library(tidyverse)
library(here)
dog_licenses <- read_csv(here("data/nyc-dog-licenses.csv.gz"))
```


## How can I work with two or more datasets?

So far all the manipulation verbs you have seen work on a single table of data.  But what if your questions involve data spread across more than one dataset?  As an example, consider the question: which neighborhood has the highest proportion of dog ownership in 2016?  You can get a handle on how many dog are licensed during 2016 for each neighborhood using the manipulations you've seen so far:
```{r}
dogs_by_nta <- dog_licenses %>% 
  filter(
    license_issued_date >= as.Date("2016-01-01"), 
    license_issued_date <= as.Date("2016-12-31")
    ) %>% 
  group_by(neighborhood_tabulation_area) %>% 
  summarise(n_licences = n()) %>% 
  arrange(desc(n_licences))
dogs_by_nta
```

Looks like MN12 has the most licenses, but what neighborhood is this?  And are there many licenses because a high proportion of people own dogs, or just that there are lots of people in this neighborhood?  To answer both of these questions you would need to combine this data with additional data, say that contained in `nyc-nta-pop`
```{r, eval = FALSE}
nta_pop <- read_csv("nyc-nta-population.csv")
```
```{r, echo = FALSE}
nta_pop <- read_csv(here("data", "nyc-nta-population.csv"))
```

```{r}
nta_pop
```

This data includes populations for each NTA and a full name for the NTA.  If you look at the neighborhoods with the largest populations:
```{r}
nta_pop %>% 
  arrange(desc(population)) %>% 
  select(nta_code, nta_name, year, population, year)
```
You'll see that MN12, the Upper West Side, in fact has the largest population of all the neighborhoods (at least in 2000 and 2010), so perhaps it isn't surprising it also has the most dog licenses issued in 2016.

Joins are a way to combine two datasets together by using a column the datasets have in common - like the columns that contain the NTA codes.  Let's work with some smaller datasets to see how this works.  
```{r, include = FALSE}
dogs <- tribble(
  ~animal_name,  ~nta, ~breed_name,
  "PRECIOUS",  "QN12",  "Yorkshire Terrier",
  "TOBY",      "QN55",  "Jack Russell Terrier",
  "GUCCI",     "QN01",  "Yorkshire Terrier",
  "HUNTER",    "QN12",  "Unknown", 
)
neighborhoods <- nta_pop %>% 
  filter(nta_code %in% c("QN01", "QN02", "QN03", "QN12"), year == 2010) %>% 
  select(nta = nta_code, population, nta_name)
```

One dataset might have rows that correspond to dogs, and their neighborhood:
```{r}
dogs
```

Another dataset might have rows that correspond to neighborhoods and their populations:
```{r}
neighborhoods 
```
Notice the datasets both have a column called `nta`, and this column contains the NTA code. {TODO: either point out this isn't true in full size data, or make it true by changing data.} This provides a way to match rows in `dogs` to the rows in `neighborhoods`: for each row of `dogs`, look up the value of `nta` in `neighborhoods` and return the values of `population` and `nta_name`.  This might result in a new dataset that looks like this: 
```{r, echo = FALSE, message = FALSE}
dogs %>% left_join(neighborhoods)
```
Both "PRECIOUS" and "HUNTER" live in Hammels-Arverne-Edgemere with a population of 36,885 and "GUCCI" lives in South Jamaica with a population of 38,894.  Notice that "TOBY" has missing values for both `population` and `nta_name` because his value of `nta`, `"QN55"`, wasn't in this small `neighborhoods` data.  

This join can be performed with the `left_join()` function in dplyr:
```{r}
dogs %>% left_join(neighborhoods)
```
It's called a `left_join()` because we will get (at least) one row back for every row in the dataset on the left, in this case, `dogs`.  I like to think about `left_join()` working just as we described above: for each row in the dataset on the left, look up the value in the dataset on the right. 

Notice that `left_join()` gave us the message `Joining, by = "nta"`.  The join functions in dplyr will use any columns with the same name in both datasets to do the join.  If the datasets have columns for the same quantity but with different names, for instance, if instead of `neighborhoods`, you had the dataset `neighborhoods_2` with a column called `nta_code` instead of `nta`:
```{r, include = FALSE}
neighborhoods_2 <- neighborhoods %>% 
  rename(nta_code = nta)
```{r}
neighborhoods_2 
```

Then attempting the join will result in an error - `data sources have no common variables`:
```{r, error = TRUE}
dogs %>% left_join(neighborhoods_2)
```
You can use the `by` argument to the join function to specify which columns are to be used to do the join:
```{r}
dogs %>% 
  left_join(neighborhoods_2, by = c("nta" = "nta_code"))
```
The value passed to `by` is a named character vector and is created using the form
```
c("<< Name of column in LHS dataset >>" = "<< Name of column in RHS dataset >>")
```
The `by` argument in the example above, 
```{r, results = "hide"}
c("nta" = "nta_code")
```
specifies the join column is `"nta"` in `dogs`, and `"nta_code"` in `neighborhoods_2`.

### Exercise: Predicting the output of a join

Try to predict what this code will return, then run it and check your prediction:
```{r, results = "hide", message = FALSE}
neighborhoods %>% left_join(dogs)
```

### Exercise: Dog licenses per capita

* Use `left_join()` to join `dogs_by_nta` with `nta_pop`. 

* `filter()` the result to only use population from `year` 2010.

* Add a `mutate()` step to find the number of dog licenses per person in each NTA.

* Which neighborhood has the highest dog licenses per person based on the 2010 population?

```{r, include = FALSE}
dogs_by_nta %>% 
  left_join(nta_pop, by = c("neighborhood_tabulation_area" = "nta_code")) %>% 
  mutate(licenses_per_person = n_licences/population) %>% 
  filter(year == 2010) %>% 
  arrange(desc(licenses_per_person)) %>% 
  select(nta_name, licenses_per_person)
# Stuyvesant Town-Cooper Village averages 0.045 dogs per person
# Or ~45 dogs per 1000 people
```

### Other join functions

Like `left_join()` all the join functions in dplyr also end in `_join` and take two datasets as arguments.  You can see the full list in:
```{r, eval = FALSE}
?dplyr::join
```

The functions `inner_join()`, `left_join()`, `right_join()` and `full_join()` all take two datasets and return a single dataset with **all the columns** from both.  They differ in the treatment of rows with no matches:

* `left_join()`, as you've seen, keeps rows in the left hand dataset, but if there is no match in the right hand dataset, columns from the right hand dataset are given missing values.
    ```{r}
    dogs %>% left_join(neighborhoods)
    ```

* `right_join()`, is just like `left_join()` but with the roles of the data sets reversed.
    ```{r}
    dogs %>% right_join(neighborhoods)
    ```

* `inner_join()` returns only rows that have matches.
    ```{r}
    dogs %>% inner_join(neighborhoods)
    ```

* `full_join()` returns all rows from both datasets, using missing values in columns where a row doesn't have a match in the other dataset.
    ```{r}
    dogs %>% full_join(neighborhoods)
    ```

Another useful join is `anti_join()`.  It doesn't actually return a combined dataset, you only get columns from one dataset back, but it returns the rows in the left hand dataset with no matches in the right hand dataset.  This is a great way to explore what you might be missing.  For example, it makes it easy to ask which rows in `dogs` don't have matches in `neighborhoods`:
```{r}
dogs %>% anti_join(neighborhoods)
```

With the full size data, there aren't any NTAs for licensed dogs that aren't in our population data:
```{r}
dogs_by_nta %>% 
  anti_join(nta_pop, by = c("neighborhood_tabulation_area" = "nta_code")) 
```

But there are a few NTAs in the population data, with no licensed dogs:
```{r}
nta_pop %>% 
  anti_join(dogs_by_nta, by = c("nta_code" = "neighborhood_tabulation_area")) 
```
Perhaps it's not surprising that there aren't any dogs licensed in the airport or cemetery.  What about Rikers Island? You might try a web search to see if you can figure out why none of the ~12,000 residents have dogs.

### Exercise

```{r, eval = FALSE, include = FALSE}
set.seed(01819)
library(babynames)

dog_counts <- dog_licenses %>% 
  filter(animal_gender == "F", lubridate::year(animal_birth_month) == 2016) %>% 
  group_by(animal_name) %>% 
  summarise(n_dogs = n()) %>% 
  rename(name = animal_name)

five_baby_names <- 
  babynames %>% 
  filter(year == 2016, sex == "F") %>% 
  mutate(name = toupper(name)) %>% 
  rename(n_babies = n) %>% 
  semi_join(dog_counts) %>% 
  sample_n(size = 5) %>% 
  select(name, n_babies)

# datapasta::tribble_paste(five_baby_names)
# datapasta::tribble_paste(dog_counts %>% semi_join(five_baby_names))
```
Consider the following two datasets, one with a count of the number of babies (human) born in 2016 for five names:
```{r}
babynames <- tribble(
      ~name, ~n_babies,
   "CLAIRE",     5207L,
     "ZURI",      889L,
     "SUZY",       20L,
    "VICKI",       13L,
  "CHELSEA",      929L
  )
babynames
```

And one with the number of dogs licensed in New York whose birth month is in 2016:
```{r}
dognames <- tribble(     
      ~name, ~n_dogs,
  "CHELSEA",      1L,
   "CLAIRE",      2L,
     "SUZY",      3L,
    "VICKI",      1L,
     "ZURI",      1L
  )
dognames
```

* Use `left_join()` to combine the datasets.

* Why doesn't this join work?  Can you fix it?
```{r}
babynames_2 <- tribble(
      ~name, ~n,
   "CLAIRE",     5207L,
     "ZURI",      889L,
     "SUZY",       20L,
    "VICKI",       13L,
  "CHELSEA",      929L
  )
dognames_2 <- tribble(     
      ~name,      ~n,
  "CHELSEA",      1L,
   "CLAIRE",      2L,
     "SUZY",      3L,
    "VICKI",      1L,
     "ZURI",      1L
  )
left_join(babynames_2, dognames_2)
```

* Why doesn't this join work?  Can you fix it?
```{r}
babynames_3 <- tribble(
      ~name, ~n_babies,
   "Claire",     5207L,
     "Zuri",      889L,
     "Suzy",       20L,
    "Vicki",       13L,
  "Chelsea",      929L
  )
dognames_3 <- tribble(     
      ~name,      ~n_dogs,
  "CHELSEA",      1L,
   "CLAIRE",      2L,
     "SUZY",      3L,
    "VICKI",      1L,
     "ZURI",      1L
  )
left_join(babynames_3, dognames_3)
```


### Exercise

Find the dog licences per capita, but now at the community district level, by joining the `dog_licenses` data with the `nyc-cd-pop` data.
