---
title: "Merely Useful: Research Software Engineering with R"
author: "Madeleine Bonsma-Fisher, Kate Hertweck, Damien Irving, Luke Johnston, Christina Koch, Sara Mahallati, Joel Ostblom, Elizabeth Wickes, Charlotte Wickham, and Greg Wilson"
documentclass: book
bibliography: book.bib
citation-package: none
link-citations: yes
github-repo: merely-useful/merely-useful.github.io
site: bookdown::bookdown_site
date: "`r Sys.Date()`"
lot: yes
lof: yes
---

# Introduction {#rse-intro}

```{r r-rse-index-setup, include=FALSE}
source(here::here("_common.R"))
knitr::opts_chunk$set(eval = FALSE)
```

> It's still magic even if you know how it's done.
>
> — Terry Pratchett

Computers are now as essential for research as telescopes, test tubes, and libraries,
which means that researchers need need to know how to build, use, and share software.
However,
most introductions to programming focus on developing commercial applications,
not on exploring problems and answering questions.
This series of books will show you how to do the latter
both on your own and as part of a team.

The prequels to this book introduce three core skills:
how to write short programs to clean and analyze data in a reproducible way,
how to keep track of what you have done,
and how to share your software and your results with other people.
But just as some astronomers spend their careers designing telescopes,
some researchers focus on building software to make their colleagues' lives easier.
People who do this are called [research software engineers][rse],
and the aim of this book is to get you ready for this role—to help you go from writing code for your own use
to creating tools to help your entire field advance.
All of this material can be freely re-used under the terms of the Creative Commons Attribution License (CC-BY 4.0);
please see Appendix \@ref(license) for details.

## Who are these lessons for? {#rse-intro-personas}

Amira
:   completed a master's in library science five years ago
    and has since worked for a small aid organization.
    She did some statistics during her degree,
    and has learned some R and Python by doing data science courses online,
    but has no formal training in programming.
    Amira would like to tidy up the scripts, data sets, and reports she has created
    in order to share them with her colleagues.
    These lessons will show her how to do this and what "done" looks like.

Jun
:   completed an [Insight Data Science][insight] fellowship last year after doing a PhD in Geology
    and now works for a company that does forensic audits.
    He uses a variety of machine learning and visualization packages,
    and would now like to turn some of his own work into an open source project.
    This book will show him how such a project should be organized
    and how to encourage people to contribute to it.

Sami
:   became a competent programmer during a bachelor's degree in applied math
    and was then hired by the university's research computing center.
    The kinds of applications they are being asked to support
    have shifted from fluid dynamics to data analysis;
    this guide will teach them how to build and run data pipelines
    so that they can pass those skills on to their users.

### Summary

For researchers and data scientists who can write functions to create programs that are several pages long,
and who want to be more productive and have more confidence in their results,
this book provides a pragmatic, tools-based introduction to research software engineering.
Unlike material courses aimed at computer scientists and professional software developers,
this book uses data analysis as a motivating example
and assumes that the learner's ultimate goal is to answer questions rather than ship products.

### Prerequisites

Learners must be able to:

-   Write reusable functions to analyze tabular data using [R][r].
-   Keep track of their work using [Git][git] and [GitHub][github].
-   Create reproducible reports using [R Markdown][r-markdown].
-   Publish those reports with appropriate metadata such as a [DOI][doi] and [ORCID][orcid].

Learners will need:

-   A personal computer with Internet access.
-   A [Unix shell][unix-shell] (see Appendix \@ref(install) for how to set this up on Windows).
-   [R][r] and [RStudio][rstudio].
-   A [GitHub][github] account.

If you are using Windows,
please install [Chocolatey][chocolatey]:
you can use it to install other programs as you go along.

## What's the big picture? {#rse-intro-big-picture}

To make sense of what follows,
we must distinguish two key ideas:

-   [Open science][open-science]
    focuses on making data, methods, and results freely available to all
    by publishing them under [open licenses][open-license].
-   [Reproducible research][reproducible-research]
    means ensuring that anyone with access to data and software can feasibly reproduce results,
    both to check them and to build on them.

People often conflate these terms,
but they are distinct:

-   If you share your data and the programs that analyze it,
    but don't document what steps to take in what order,
    your work is open but not reproducible.
-   If you completely automate your analysis,
    but your data is only available to people in your lab,
    your work is reproducible but not open.

The third key idea is [sustainability][sustainable-software].
Software is sustainable if it's easier for people to maintain it and extend it than to replace it.
However,
sustainability isn't just a property of the software:
it also depends on the skills and culture of its users.
If a software package is being maintained by a couple of post-docs
who are being paid a fraction of what they could earn in industry
and have no realistic hope of promotion because their field doesn't value tool building,
it doesn't matter whether the package is well tested and easy to install:
sooner or later (probably sooner)
it will become [abandonware][abandonware].

### Why isn't all of this already normal?

Nobody argues that research should be irreproducible or unsustainable,
but "not against it" and actively supporting it are very different things.
Academia doesn't yet know how to reward people for writing useful software,
so while you may be thanked,
the effort you put in may not translate into job security or decent pay.

And some people still worry that if they make their data and code generally available,
someone else will use it and publish a result they have come up with themselves.
This is almost unheard of in practice,
but that doesn't stop people using it as a scare tactic.
Other people are afraid of looking foolish or incompetent by sharing code that might contain bugs.
This isn't just [impostor syndrome][impostor-syndrome]:
members of marginalized groups are frequently judged more harshly than others,
so being wrong in public is much riskier for them.

### Syllabus

This book covers the tools and practices needed to create work that is open, reproducible, and sustainable.
You can use this book for self-study
or as the foundation of a one-semester course for graduate students or senior undergraduates.
By its end,
you will be able to:

-   Use [branches][branch] to coordinate development of complex software.
-   Enable users to configure software without modifying it.
-   Re-run analyses and rebuild packages reproducibly with a single command.
-   Test your software and know which parts have not yet been tested.
-   Keep track of what still needs to be done and who's working on it.
-   Work on remote computers such as computing clusters or the cloud.
-   Document software so it can be easily used, maintained, and extended.
-   Create packages that can be installed in standard ways and be used by others.
-   Work productively in a small team where everyone is welcome.

## Acknowledgments {#rse-intro-ack}

This book owes its existence to
everyone we met through [the Carpentries][carpentries].
We are also grateful to [Insight Data Science][insight] for sponsoring the early stages of this work
and to everyone who has contributed,
particularly Jonathan Dursi
and the authors of [@Nobl2009; @Hadd2010; @Wils2014; @Scop2015; @Tasc2017; @Wils2017; @Brow2018; @Deve2018; @Shol2019; @Wils2019].

```{r, child="./links.md"}
```
