# Working Remotely {#remote}

## Questions {#remote-questions}

-   How can I work with remote computers?

## Objectives {#remote-objectives}

```{r, child="objectives/remote.md"}
```

## Introduction {#remote-intro}

Computers keep getting faster,
but data keeps getting bigger.
While today's laptops would have been considered supercomputers a decade ago,
they cannot store or process all of the information that researchers have available to them.
Even if they could,
much of that information is too sensitive to be put on anyone's personal machine.
This lesson will therefore introduce you to some tools that can be used
to connect to and work on remote computers.

## How are connections to remote computers implemented? {#remote-implement}

When we use the Unix shell on a desktop or laptop computer,
the first step is to log in
so that the operating system knows who we are and what we're allowed to do.
We do this by typing our username and password;
the operating system checks those values against its records,
and if they match,
runs a shell for us.

As we type commands,
the characters we're typing are sent from the keyboard to the shell.
The shell displays those characters on the screen to represent what we type,
and then,
if what we typed was a command,
the shell executes it and displays its output (if any).

FIXME: diagram

What if we want to run some commands on another machine,
such as the cluster in the basement?
To do this,
we have to first log in to that machine.
We call this a [remote login](glossary.html#remote-login).

In order for us to do this,
we have to run a client program that can talk to that server.
The client program passes our login credentials to the remote server;
if we are allowed to login,
that server then runs a shell for us on the remote computer.

Once our local client is connected to the remote server,
everything we type is passed to
the shell running on the remote computer.
That remote shell runs those commands on our behalf,
just as a local shell would,
then sends back output, via the server, to our client, for our computer to display:

FIXME: diagram

The most important thing about the connection between the local and remote computers
isn't its reliability,
but how secure it is.
In the early days of the Internet,
two programs called `rsh` and `rcp` were frequently used to run an interactive shell on a remote computer
and copy files back and forth between computers.
However,
since the messages they sent were not encrypted,
anyone who wanted to snoop on the traffic between the two machines could do so.

The [SSH protocol](glossary.html#ssh-protocol) was invented to solve this problem.
It uses several sophisticated (and heavily tested) encryption protocols
to keep the traffic between computers confidential.
It can be used interactively via two programs called `ssh` (which stands for "secure shell")
and `scp` (which stands for "secure copy").

## How can I run commands on a remote computer? {#remote-ssh}

To log in to a remote computer,
we use the command <code>ssh <em>username</em>@<em>hostname</em></code>.
If the login is successful,
we will be presented with a shell prompt that just happens to be coming from somewhere else.
Typing `exit` or Control-D terminates the remote shell,
just as it would terminate a local shell,
and returns us to our previous shell.

FIXME: diagram

In the example below,
the local machine's command prompt is `earth$` and the remote computer's is `moon$`.

```bash
# Check current directory on local computer.
earth$ pwd
```
```text
/users/pterry
```

```bash
# Log in to remote computer `moon.euphoric.edu` as `tp`.
earth$ ssh tp@moon.euphoric.edu
Password: ********
```

```bash
# Check name of computer on which remote shell is running.
moon$ hostname
```
```text
moon
```

```bash
# Check current directory on remote computer.
moon$ pwd
```
```text
/home/tp
```

```bash
# List contents of directory on remote computer.
moon$ ls -F
```
```text
bin/     cheese.txt   dark_side/   rocks.cfg
```

```bash
# End remote shell session.
moon$ exit
```

```bash
# Confirm that we are back in the starting directory on the local computer.
earth$ pwd
```
```text
/users/pterry
```

## How can I execute commands on a remote computer without logging in? {#remote-command}

Suppose we want to check whether the file `backups/results-2011-11-12.dat` exists on a remote computer.
Instead of logging in and then typing `ls`,
we can tell `ssh` to run just that single command:

```bash
$ ssh pterry@moon.euphoric.edu "ls results*"
Password: ********
```
``` text
results-2018-09-18.dat
results-2018-10-04.dat
results-2018-10-28.dat
results-2018-11-11.dat
```

`ssh` takes the argument after `username@computer` and passes it to the remote shell;
in this case the command contains a space,
so we have to enclose it in quotes.
The remote shell runs `ls results` for us
and sends the output back to our local shell for display.

## How can I copy files to and from remote computers? {#remote-scp}

To copy a file with `cp`,
we specify the source and destination paths.
The secure copying command `scp` works exactly the same way,
except either or both of the paths may contain a hostname.
If we leave out the hostname,
`scp` assumes we mean the machine we're running on.

For example,
this command copies our latest results to the backup server,
printing out the status as it does so:

```bash
$ scp results.dat pterry@moon.euphoric.edu:backups/results-2018-11-11.dat
Password: ********
```
```text
results.dat              100%  9  1.0 MB/s 00:00
```

Note the colon `:`, separating the hostname of the server
and the pathname of  the file we are copying to.
It is this character that informs `scp` that
the source or target of the copy is on the remote machine.

Copying a whole directory uses the same syntax as the `cp` command:
we just add the `-r` option to signal that we want to copy recursively.
For example,
this command copies all of our results from the backup server to our laptop:

```bash
$ scp -r pterry@moon.euphoric.edu:backups ./backups
Password: ********
```
```text
results-2018-09-18.dat              100%  7  1.0 MB/s 00:00
results-2018-10-04.dat              100%  9  1.0 MB/s 00:00
results-2018-10-28.dat              100%  8  1.0 MB/s 00:00
results-2018-11-11.dat              100%  9  1.0 MB/s 00:00
```

`ssh` can even be used in a pipe:

```bash
earth$ ssh pterry@moon.euphoric.edu "ls results" | sort -r
```
```text
results-2018-11-11.dat
results-2018-10-28.dat
results-2018-10-04.dat
results-2018-09-18.dat
```

FIXME: diagram

## How can I avoid typing my password over and over again? {#remote-keys}

Typing our password over and over again is annoying,
especially if the commands we want to run remotely are in a loop.
To avoid doing this,
we can create an [SSH key](glossary.html#ssh-key)
to tell the remote machine that it should always trust us.

SSH keys come in pairs:
a public key that gets shared with services like GitHub,
and a private key that is stored only on your computer.
If the keys match, you're granted access.
The cryptography behind SSH keys ensures that
no one can reverse engineer your private key from the public one.

You might already have an SSH key pair on your machine.
To check,
see if the directory `~/.ssh` exists
(i.e., if there is a directory called `.ssh` with a leading `.` in its name
directly below your home directory):

```bash
$ ls ~/.ssh
```

<!-- == noindent -->
If a file called `id_rsa.pub` is there,
you already have a key pair and don't need to create a new one.

If you don't see `id_rsa.pub`,
use the following command to generate a new key pair
(making sure to replace `your@email.com` with your own email address):

```bash
$ ssh-keygen -t rsa -C "your@email.com"
```

<!-- == noindent -->
When asked where to save the new key, hit enter to accept the default location.
You should see a message like this,
with your username instead of `pterry`:

```text
Generating public/private rsa key pair.
Enter file in which to save the key (/Users/pterry/.ssh/id_rsa):
```

You will then be asked to provide an optional passphrase.
This can be used to make your key even more secure,
but if what you want is avoiding type your password every time you can skip it by hitting enter twice.

```bash
Enter passphrase (empty for no passphrase):
Enter same passphrase again:
```

When the key generation is complete you should see the following confirmation:

```text
Your identification has been saved in /Users/pterry/.ssh/id_rsa.
Your public key has been saved in /Users/pterry/.ssh/id_rsa.pub.
The key fingerprint is:
01:0f:f4:3b:ca:85:d6:17:a1:7d:f0:68:9d:f0:a2:db your@email.com
The key's randomart image is:
+--[ RSA 2048]----+
|                 |
|                 |
|        . E +    |
|       . o = .   |
|      . S =   o  |
|       o.O . o   |
|       o .+ .    |
|      . o+..     |
|       .+=o      |
+-----------------+
```

<!-- == noindent -->
(The random art image is an alternative way to match keys,
which we won't use.)

You now you need to place a copy of your public key
on any remote computer you would like to connect to
without typing your password.
The content of your public key is in `~/.ssh/id_rsa.pub`:

```bash
$ cat ~/.ssh/id_rsa.pub
```
```text
ssh-rsa AAAAB3NzaC1yc2EAAAABIwAAAQEA879BJGYlPTLIuc9/R5MYiN4yc/YiCLcdBpSdzgK9Dt0Bkfe3rSz5cPm4wmehdE7GkVFXrBJ2YHqPLuM1yx1AUxIebpwlIl9f/aUHOts9eVnVh4NztPy0iSU/Sv0b2ODQQvcy2vYcujlorscl8JjAgfWsO3W4iGEe6QwBpVomcME8IU35v5VbylM9ORQa6wvZMVrPECBvwItTY8cPWH3MGZiK/74eHbSLKA4PY3gM4GHI450Nie16yggEg2aTQfWA1rry9JYWEoHS9pJ1dnLqZU3k/8OWgqJrilwSoC5rGjgp93iu0H8T6+mEHGRQe84Nk1y5lESSWIbn6P636Bl3uQ== your@email.com
```

Copy this to your clipboard,
then log in to the remote server with your username and password:

```bash
$ ssh pterry@moon.euphoric.edu
Password: ********
```

<!-- == noindent -->
and paste the content of the clip at the end of `~/.ssh/authorized_keys`
on the remote computer.

After you have appended the public key,
logout of the remote machine and try login again.
If you set up your SSH key correctly you won't need to type your password.

```bash
$ ssh pterry@moon.euphoric.edu "ls results"
```
``` text
results-2018-09-18.dat
results-2018-10-04.dat
results-2018-10-28.dat
results-2018-11-11.dat
```

There are several other ways to accomplish this task.
For example,
you can `scp` the file `~/.ssh/id_rsa.pub` from your local computer to the remote computer,
then use a text editor to add it to `~/.ssh/authorized_keys` on the remote computer.
Whatever you do,
be careful not to overwrite the content of the remote `authorized_keys` file.

## Summary {#remote-summary}

FIXME: create concept map for SSH and its kin.

## Exercises {#remote-exercises}

FIXME: create exercises for SSH and its kin.

## Key Points {#remote-keypoints}

```{r, child="keypoints/remote.md"}
```

```{r, child="etc/links.md"}
```
