# Solutions {#rse-solutions}

```{r rse-solutions-setup, include=FALSE}
source("_common.R")
```

## Chapter \@ref(rse-bash)

### Exercise \@ref(#rse-bash-ex-more-ls)

The `-l` option makes `ls` use a **l**ong listing format, showing not only
the file/directory names but also additional information such as the file size
and the time of its last modification. If you use both the `-h` option and the `-l` option,
this makes the file size "**h**uman readable", i.e. displaying something like `5.3K`
instead of `5369`.

### Exercise \@ref(#rse-bash-ex-ls-rt)

The files/directories in each directory are sorted by time of last change.

### Exercise \@ref(#rse-bash-ex-paths)

1. No: `.` stands for the current directory.
2. No: `/` stands for the root directory.
3. No: Amanda's home directory is `/Users/amanda`.
4. No: this goes up two levels, i.e. ends in `/Users`.
5. Yes: `~` stands for the user's home directory, in this case `/Users/amanda`.
6. No: this would navigate into a directory `home` in the current directory if it exists.
7. Yes: unnecessarily complicated, but correct.
8. Yes: shortcut to go back to the user's home directory.
9. Yes: goes up one level.

### Exercise \@ref(#rse-bash-ex-resolve-rel-path)

1. No: there *is* a directory `backup` in `/Users`.
2. No: this is the content of `Users/thing/backup`,
   but with `..` we asked for one level further up.
3. No: see previous explanation.
4. Yes: `../backup/` refers to `/Users/backup/`.

### Exercise \@ref(#rse-bash-ex-reading-ls)

 1. No: `pwd` is not the name of a directory.
 2. Yes: `ls` without directory argument lists files and directories
    in the current directory.
 3. Yes: uses the absolute path explicitly.

### Exercise \@ref(#rse-bash-ex-touch)

1.  The `touch` command generates a new file called `my_file.txt` in
    your current directory.  You
    can observe this newly generated file by typing `ls` at the
    command line prompt. `my_file.txt` can also be viewed in your
    GUI file explorer.

2.  When you inspect the file with `ls -l`, note that the size of
    `my_file.txt` is 0 bytes.  In other words, it contains no data.
    If you open `my_file.txt` using your text editor it is blank.

3.  Some programs do not generate output files themselves, but
    instead require that empty files have already been generated.
    When the program is run, it searches for an existing file to
    populate with its output.  The touch command allows you to
    efficiently generate a blank text file to be used by such
    programs.

### Exercise \@ref(#rse-bash-ex-move-dot)

```shell
$ mv ../analyzed/sucrose.dat ../analyzed/maltose.dat .
```

Recall that `..` refers to the parent directory (i.e. one above the current directory)
and that `.` refers to the current directory.

### Exercise \@ref(#rse-bash-ex-renaming-files)

1. No.  While this would create a file with the correct name,
the incorrectly named file still exists in the directory and would need to be deleted.
2. Yes, this would work to rename the file.
3. No, the period(.) indicates where to move the file, but does not provide a new file name;
identical file names cannot be created.
4. No, the period(.) indicates where to copy the file, but does not provide a new file name;
identical file names cannot be created.

### Exercise \@ref(#rse-bash-ex-last-ls)

We start in the `/Users/jamie/data` directory, and create a new folder called `recombine`.
The second line moves (`mv`) the file `proteins.dat` to the new folder (`recombine`).
The third line makes a copy of the file we just moved.  The tricky part here is where the file was
copied to.  Recall that `..` means "go up a level", so the copied file is now in `/Users/jamie`.
Notice that `..` is interpreted with respect to the current working
directory, **not** with respect to the location of the file being copied.
So, the only thing that will show using ls (in `/Users/jamie/data`) is the recombine folder.

1. No, see explanation above.  `proteins-saved.dat` is located at `/Users/jamie`
2. Yes
3. No, see explanation above.  `proteins.dat` is located at `/Users/jamie/data/recombine`
4. No, see explanation above.  `proteins-saved.dat` is located at `/Users/jamie`

### Exercise \@ref(#rse-bash-ex-safe-rm)

```shell
$ rm: remove regular file 'thesis_backup/quotations.txt'? y
```

The `-i` option will prompt before (every) removal
(use <kbd>Y</kbd> to confirm deletion or <kbd>N</kbd> to keep the file).
The Unix shell doesn't have a trash bin, so all the files removed will disappear forever.
By using the `-i` option, we have the chance to check that we are deleting
only the files that we want to remove.

### Exercise \@ref(#rse-bash-ex-copy-multi)

If given more than one file name followed by a directory name (i.e. the destination directory must 
be the last argument), `cp` copies the files to the named directory.

If given three file names, `cp` throws an error such as the one below, because it is expecting a directory
name as the last argument.

```text
cp: target ‘morse.txt’ is not a directory
```

### Exercise \@ref(#rse-bash-ex-ls-match)

The solution is `3.`

`1.` shows all files whose names contain zero or more characters (`*`) followed by the letter `t`,
then zero or more characters (`*`) followed by `ane.pdb`.
This gives `ethane.pdb  methane.pdb  octane.pdb  pentane.pdb`. 

`2.` shows all files whose names start with zero or more characters (`*`) followed by the letter `t`,
then a single character (`?`), then `ne.` followed by zero or more characters (`*`).
This will give us `octane.pdb` and `pentane.pdb` but doesn't match anything which ends in `thane.pdb`.

`3.` fixes the problems of option 2 by matching two characters (`??`) between `t` and `ne`.
This is the solution.

`4.` only shows files starting with `ethane.`.

### Exercise \@ref(#rse-bash-ex-more-wildcards)

```shell
$ cp *calibration.txt backup/calibration
$ cp 2015-11-* send_to_bob/all_november_files/
$ cp *-23-dataset* send_to_bob/all_datasets_created_on_a_23rd/
```

### Exercise \@ref(#rse-bash-ex-organizing)

```shell
mv *.dat analyzed
```

Jamie needs to move her files `fructose.dat` and `sucrose.dat` to the `analyzed` directory.
The shell will expand *.dat to match all .dat files in the current directory.
The `mv` command then moves the list of .dat files to the "analyzed" directory.

### Exercise \@ref(#rse-bash-ex-reproduce-structure)

The first two sets of commands achieve this objective.
The first set uses relative paths to create the top level directory before
the subdirectories.

The third set of commands will give an error because `mkdir` won't create a subdirectory
of a non-existant directory: the intermediate level folders must be created first.

The final set of commands generates the 'raw' and 'processed' directories at the same level
as the 'data' directory.

### Exercise \@ref(#rse-bash-ex-sort-n)

The `-n` option specifies a numerical rather than an alphanumerical sort.

### Exercise \@ref(#rse-bash-ex-redirect-append)

In the first example with `>`, the string "hello" is written to `testfile01.txt`,
but the file gets overwritten each time we run the command.

We see from the second example that the `>>` operator also writes "hello" to a file
(in this case`testfile02.txt`),
but appends the string to the file if it already exists (i.e. when we run it for the second time).

### Exercise \@ref(#rse-bash-ex-append-data)

Option 3 is correct.
For option 1 to be correct we would only run the `head` command.
For option 2 to be correct we would only run the `tail` command.
For option 4 to be correct we would have to pipe the output of `head` into `tail -n 2`
by doing `head -n 3 animals.txt | tail -n 2 > animals-subset.txt`

### Exercise \@ref(#rse-bash-ex-piping)

Option 4 is the solution.
The pipe character `|` is used to feed the standard output from one process to
the standard input of another.
`>` is used to redirect standard output to a file.
Try it in the `data-shell/molecules` directory!

### Exercise \@ref(#rse-bash-ex-uniq-adjacent)

```shell
$ sort salmon.txt | uniq
```

### Exercise \@ref(#rse-bash-ex-reading-pipes)

The `head` command extracts the first 5 lines from `animals.txt`.
Then, the last 3 lines are extracted from the previous 5 by using the `tail` command.
With the `sort -r` command those 3 lines are sorted in reverse order and finally,
the output is redirected to a file `final.txt`.
The content of this file can be checked by executing `cat final.txt`.
The file should contain the following lines:
```text
2012-11-06,rabbit
2012-11-06,deer
2012-11-05,raccoon
```

### Exercise \@ref(#rse-bash-ex-pipe-construction)

```shell
$ cut -d , -f 2 animals.txt | sort | uniq
```

### Exercise \@ref(#rse-bash-ex-which-pipe)

Option 4. is the correct answer.
If you have difficulty understanding why, try running the commands, or sub-sections of
the pipelines (make sure you are in the `data-shell/data` directory).

### Exercise \@ref(#rse-bash-ex-wildcard-expressions)

1. A solution using two wildcard expressions:
    ```shell
    $ ls *A.txt
    $ ls *B.txt
    ```
2. The output from the new commands is separated because there are two commands.
3. When there are no files ending in `A.txt`, or there are no files ending in
`B.txt`.

### Exercise \@ref(#rse-bash-ex-remove-unneeded)

1. This would remove `.txt` files with one-character names
2. This is correct answer
3. The shell would expand `*` to match everything in the current directory,
so the command would try to remove all matched files and an additional
file called `.txt`
4. The shell would expand `*.*` to match all files with any extension,
so this command would delete all files

### Exercise \@ref(#rse-bash-ex-loop-dry-run)

The second version is the one we want to run.
This prints to screen everything enclosed in the quote marks, expanding the
loop variable name because we have prefixed it with a dollar sign.

The first version redirects the output from the command `echo analyze $file` to
a file, `analyzed-$file`. A series of files is generated: `analyzed-cubane.pdb`,
`analyzed-ethane.pdb` etc.

Try both versions for yourself to see the output! Be sure to open the
`analyzed-*.pdb` files to view their contents.

### Exercise \@ref(#rse-bash-ex-nested-loops)

We have a nested loop, i.e. contained within another loop, so for each species
in the outer loop, the inner loop (the nested loop) iterates over the list of
temperatures, and creates a new directory for each combination.

Try running the code for yourself to see which directories are created!

### Exercise \@ref(#rse-bash-ex-loop-variables)

The first code block gives the same output on each iteration through
the loop.
Bash expands the wildcard `*.pdb` within the loop body (as well as
before the loop starts) to match all files ending in `.pdb`
and then lists them using `ls`.
The expanded loop would look like this:
```shell
$ for datafile in cubane.pdb  ethane.pdb  methane.pdb  octane.pdb  pentane.pdb  propane.pdb
> do
>	ls cubane.pdb  ethane.pdb  methane.pdb  octane.pdb  pentane.pdb  propane.pdb
> done
```

```text
cubane.pdb  ethane.pdb  methane.pdb  octane.pdb  pentane.pdb  propane.pdb
cubane.pdb  ethane.pdb  methane.pdb  octane.pdb  pentane.pdb  propane.pdb
cubane.pdb  ethane.pdb  methane.pdb  octane.pdb  pentane.pdb  propane.pdb
cubane.pdb  ethane.pdb  methane.pdb  octane.pdb  pentane.pdb  propane.pdb
cubane.pdb  ethane.pdb  methane.pdb  octane.pdb  pentane.pdb  propane.pdb
cubane.pdb  ethane.pdb  methane.pdb  octane.pdb  pentane.pdb  propane.pdb
```

The second code block lists a different file on each loop iteration.
The value of the `datafile` variable is evaluated using `$datafile`,
and then listed using `ls`.

```text
cubane.pdb
ethane.pdb
methane.pdb
octane.pdb
pentane.pdb
propane.pdb
```

### Exercise \@ref(#rse-bash-ex-limiting-file-sets)

**Part 1**

4 is the correct answer. `*` matches zero or more characters, so any file name starting with
the letter c, followed by zero or more other characters will be matched.

**Part 2**

4 is the correct answer. `*` matches zero or more characters, so a file name with zero or more
characters before a letter c and zero or more characters after the letter c will be matched.

### Exercise \@ref(#rse-bash-ex-loop-save)

**Part 1**

1. The text from each file in turn gets written to the `alkanes.pdb` file.
However, the file gets overwritten on each loop interation, so the final content of `alkanes.pdb`
is the text from the `propane.pdb` file.

**Part 2**

3 is the correct answer. `>>` appends to a file, rather than overwriting it with the redirected
output from a command.
Given the output from the `cat` command has been redirected, nothing is printed to the screen.

### Exercise \@ref(#rse-bash-ex-list-unique)

```text
# Script to find unique species in csv files where species is the second data field
# This script accepts any number of file names as command line arguments

# Loop over all files
for file in $@ 
do
	echo "Unique species in $file:"
	# Extract species names
	cut -d , -f 2 $file | sort | uniq
done
```

### Exercise \@ref(#rse-bash-ex-history-order)

If a command causes something to crash or hang, it might be useful
to know what that command was, in order to investigate the problem.
Were the command only be recorded after running it, we would not
have a record of the last command run in the event of a crash.

### Exercise \@ref(#rse-bash-ex-script-variables)

The correct answer is 2. 

The special variables $1, $2 and $3 represent the command line arguments given to the
script, such that the commands run are:

```shell
$ head -n 1 cubane.pdb ethane.pdb octane.pdb pentane.pdb propane.pdb
$ tail -n 1 cubane.pdb ethane.pdb octane.pdb pentane.pdb propane.pdb
```

The shell does not expand `'*.pdb'` because it is enclosed by quote marks.
As such, the first argument to the script is `'*.pdb'` which gets expanded within the
script by `head` and `tail`.

### Exercise \@ref(#rse-bash-ex-longest-with-extension)

```text
# Shell script which takes two arguments: 
#    1. a directory name
#    2. a file extension
# and prints the name of the file in that directory
# with the most lines which matches the file extension.

wc -l $1/*.$2 | sort -n | tail -n 2 | head -n 1
```

### Exercise \@ref(#rse-bash-ex-reading-scripts)

In each case, the shell expands the wildcard in `*.pdb` before passing the resulting
list of file names as arguments to the script.

Script 1 would print out a list of all files containing a dot in their name.
The arguments passed to the script are not actually used anywhere in the script.

Script 2 would print the contents of the first 3 files with a `.pdb` file extension.
`$1`, `$2`, and `$3` refer to the first, second, and third argument respectively.

Script 3 would print all the arguments to the script (i.e. all the `.pdb` files),
followed by `.pdb`.
`$@` refers to *all* the arguments given to a shell script.
```text
cubane.pdb ethane.pdb methane.pdb octane.pdb pentane.pdb propane.pdb.pdb
```

### Exercise \@ref(#rse-bash-ex-using-grep)

The correct answer is 3, because the `-w` option looks only for whole-word matches.
The other options will also match "of" when part of another word.

### Exercise \@ref(#rse-bash-ex-tracking-species)
 
```shell
grep -w $1 -r $2 | cut -d : -f 2 | cut -d , -f 1,3  > $1.txt
```

You would call the script above like this:

```shell
$ bash count-species.sh bear .
```

### Exercise \@ref(#rse-bash-ex-little-women)

```text
for sis in Jo Meg Beth Amy
do
	echo $sis:
>	grep -ow $sis LittleWomen.txt | wc -l
done
```

Alternative, slightly inferior solution:
```text
for sis in Jo Meg Beth Amy
do
	echo $sis:
>	grep -ocw $sis LittleWomen.txt
done
```

This solution is inferior because `grep -c` only reports the number of lines matched.
The total number of matches reported by this method will be lower if there is more
than one match per line.

### Exercise \@ref(#rse-bash-ex-match-subtract)

The correct answer is 1. Putting the match expression in quotes prevents the shell
expanding it, so it gets passed to the `find` command.

Option 2 is incorrect because the shell expands `*s.txt` instead of passing the wildcard
expression to `find`.

Option 3 is incorrect because it searches the contents of the files for lines which
do not match "temp", rather than searching the file names.

### Exercise \@ref(#rse-bash-ex-reading-find)

1. Find all files with a `.dat` extension recursively from the current directory
2. Count the number of lines each of these files contains
3. Sort the output from step 2. numerically

### Exercise \@ref(#rse-bash-ex-find-properties)
 
Assuming that Nelle's home is our working directory we type:

```shell
$ find ./ -type f -mtime -1 -user ahmed
```

### Exercise \@ref(#rse-bash-ex-combining-options)

FIXME: Note that in most command line tools, multiple options can be combined
with a single `-` and no spaces between the options: `ls -F -a` is equivalent to
`ls -Fa`.

### Exercise \@ref(#rse-bash-ex-other-wildcards)

FIXME: introduce ? and other wildcards.

## Chapter \@ref(rse-automate)

### Exercise \@ref(rse-automate-ex-create-summary-results) {-}

-   Add a rule to Makefile to create a summary CSV file from all of the book CSV files.
-   Be careful about writing the prerequisites so that it doesn't depend on itself.

### Exercise \@ref(rse-automate-ex-plot-top-n) {-}

-   Make it depend on the summary.

### Exercise \@ref(rse-automate-ex-mkdir) {-}

-   Why is `mkdir -p` useful?

### Exercise \@ref(rse-automate-ex-report-change) {-}

-   Write a rule to report which result files would actually change.
-   Hint: use `diff`.

### Exercise \@ref(rse-automate-ex-readable-help) {-}

-   Modify the command in the `help` action to remove the leading '##' markers from the output.

### Exercise \@ref(rse-automate-ex-wildcard-perils) {-}

What is wrong with writing the rule for `results/collated.csv` like this:

```make
results/collated.csv : results/*.csv
	$(RUN_COLLATE) $^ > $@
```

Hint: the fact that the result no longer depends on the program used to create it isn't the only problem.

## Chapter \@ref(rse-git-cmdline)

### Exercise \@ref(rse-git-cmdline-ex-places) {-}

Frances does not need to make the `univac` sub-directory a Git repository 
because the `eniac` repository will track everything inside it regardless of how deeply nested.

Frances *shouldn't* run `git init` in `univac` because nested Git repositories can interfere with each other.
If someone commits something in the inner repository,
Git will not know whether to record the changes in that repository,
the outer one,
or both.
 
### Exercise \@ref(rse-git-cmdline-ex-commit) {-}

1.  Would only create a commit if files have already been staged.  
2.  Would try to create a new repository.  
3.  Is correct: first add the file to the staging area, then commit.  
4.  Would try to commit a file "my recent changes" with the message myfile.txt.  

### Exercise \@ref(rse-git-cmdline-ex-multiple) {-}

1.  Change `names.txt` and `old-computers.txt` using an editor like Nano.
2.  Add both files to the staging area with `git add *.txt`.
3.  Check that both files are there with `git status`.
4.  Commit both files at once with `git commit`.

### Exercise \@ref(rse-git-cmdline-ex-bio) {-}

1.  Go into your home directory with `cd ~`.
2.  Create a new folder called `bio` with `mkdir bio`.
3.  Go into it with `cd bio`.
4.  Turn it into a repository with `git init`.
5.  Create your biography using Nano or another text editor.
6.  Add it and commit it in a single step with `git commit -a -m "Some message"`.
7.  Modify the file.
8.  Use `git diff` to see the differences.

### Exercise \@ref(rse-git-cmdline-ex-ignore-nested) {-}

To ignore only the contents of `results/plots`,
add this line to `.gitignore`:

```text
results/plots/
```

### Exercise \@ref(rse-git-cmdline-ex-include) {-}

Add the following two lines to `.gitignore`:

```text
*.dat           # ignore all data files
!final.dat      # except final.data
```

The exclamation point `!` includes a previously-excluded entry.

Note also that if we have previously committed `.dat` files in this repository
they will not be ignored once these rules are added to `.gitignore`.
Only future `.dat` files will be ignored.

### Exercise \@ref(rse-git-cmdline-ex-github-interface) {-}

The left button (with the picture of a clipboard)
copies the full identifier of the commit to the clipboard.
In the shell,
`git log` shows the full commit identifier for each commit.

The middle button shows all of the changes that were made in that particular commit;
green shaded lines indicate additions and red lines indicate removals.
We can show the same thing in the shell using `git diff`
or <code>git diff <em>from</em>..<em>to</em></code>
(where <code><em>from</em></code> and <code><em>to</em></code> are commit identifiers).

The right button lets us view all of the files in the repository at the time of that commit.
To do this in the shell,
we would need to check out the repository as it was at that commit
using <code>git checkout <em>id</em></code>.
If we do this,
we need to remember to put the repository back to the right state afterward.

### Exercise \@ref(rse-git-cmdline-ex-timestamp) {-}

GitHub displays timestamps in a human-readable relative format
(i.e. "22 hours ago" or "three weeks ago").
However, if we hover over the timestamp
we can see the exact time at which the last change to the file occurred.

### Exercise \@ref(rse-git-cmdline-ex-push-commit) {-}

Committing updates our local repository.
Pushing sends any commits we have made locally
that aren't yet in the remote repository
to the remote repository.

### Exercise \@ref(rse-git-cmdline-ex-boilerplate) {-}

When GitHub creates a `README.md` file while setting up a new repository,
it actually creates the repository and then commits the `README.md` file.
When we try to pull from the remote repository to our local repository,
Git detects that their histories do not share a common origin and refuses to merge them.

```shell
$ git pull origin master
```

```text
warning: no common commits
remote: Enumerating objects: 3, done.
remote: Counting objects: 100% (3/3), done.
remote: Total 3 (delta 0), reused 0 (delta 0), pack-reused 0
Unpacking objects: 100% (3/3), done.
From https://github.com/frances/eniac
 * branch            master     -> FETCH_HEAD
 * [new branch]      master     -> origin/master
fatal: refusing to merge unrelated histories
```

We can force git to merge the two repositories with the option `--allow-unrelated-histories`. 
Please check the contents of the local and remote repositories carefully before doing this.

### Exercise \@ref(rse-git-cmdline-ex-recover) {-}

The answer is (5)-Both 2 and 4. 

The `checkout` command restores files from the repository,
overwriting the files in our working directory.
Answers 2 and 4 both restore the latest version in the repository of the file `data_cruncher.sh`.
Answer 2 uses `HEAD` to indicate the latest,
while answer 4 uses the unique ID of the last commit,
which is what `HEAD` means. 

Answer 3 gets the version of `data_cruncher.sh` from the commit before `HEAD`,
which is not what we want.

Answer 1 can be dangerous:
without a filename,
`git checkout` will restore all files in the current directory (and all directories below it)
to their state at the commit specified. 
This command will restore `data_cruncher.sh` to the latest commit version,
but will also reset any other files we have changed to that version,
which will erase any unsaved changes you may have made to those files.

### Exercise \@ref(rse-git-cmdline-ex-history) {-}

The answer is 2. 
 
The command `git add history.txt` adds the current version of `history.txt` to the staging area. 
The changes to the file from the second `echo` command are only applied to the working copy, 
not the version in the staging area.
 
As a result,
when `git commit -m "Origins of ENIAC"` is executed, 
the version of `history.txt` committed to the repository is the one from the staging area
with only one line.
  
However,
the working copy still has the second line.
(`git status` will show that the file is modified.)
`git checkout HEAD history.txt` therefore replaces the working copy with
the most recently committed version of `history.txt`,
so `cat history.txt` prints:
 
```text
ENIAC was the world's first general-purpose electronic computer.
```

### Exericse \@ref(rse-git-cmdline-ex-diff) {-}

FIXME

### Exercise \@ref(rse-git-cmdline-ex-unstage) {-}

FIXME

### Exercise \@ref(rse-git-cmdline-ex-blame) {-}

FIXME

## Chapter \@ref(rse-git-advanced)

### Exercise \@ref(rse-git-advanced-ex-additional-pr) {-}

```shell
$ git checkout master
$ git checkout -b addItaly
$ cp united_states.txt italy.txt
$ nano italy.txt #Add the right info into the file
$ git add italy.txt
$ git commit -m "Added file on Italy"
$ git push origin addItaly
```

```{r, child="./links.md"}
```
