# Command line programs {#rse-py-scripting}

## Questions {#rse-py-scripting-questions}

```{r, child="questions/rse-py-scripting.md"}
```


## Introduction {#rse-py-scripting-intro}

The Jupyter Notebook, PyCharm and other interactive tools
are great for prototyping code and exploring data,
but in many cases you will ultimately want to take the code you've developed
using those tools and apply it to thousands of data files, 
run it with many different parameters,
and/or combine it with other programs in a data analysis pipeline. 
In order to do these things,
it is often necessary/preferable to turn the code you've developed into a program 
that works like any other Unix command-line tool.

In this chapter we take the code we developed earlier (TODO: Link)
and turn it into a Python script (i.e. a command line program written in Python).


## A template script {#rse-py-scripting-template}

In general, the first thing that gets added to any Python script is the following:

```python
if __name__ == '__main__':
    main()
```

The reason we need these two lines of code
is that running a Python script in bash is very similar to importing that file in Python. 
The biggest difference is that we don’t expect anything to happen when we import a file, 
whereas when running a script we expect to see some output
(e.g. an output file, figure and/or some text printed to the screen).

The `__name__` variable exists to handle these two situations.
When you import a Python file `__name__` is set to the name of that file
(e.g. when importing script.py, `__name__` is `script`),
but when running a script in bash `__name__` is always set to `__main__`.
The convention is to call the function that produces the output `main()`,
but you can call it whatever you like.

The next thing you'll need is a library to parse the command line for input arguments.
The most widely used option is 
[argparse](https://docs.python.org/3/library/argparse.html).

Putting those together,
here's a basic python script for us to start with:

```shell
$ cat script_template.py
```

```python
import argparse

#
# All your functions (that will be called by main()) go here.
#

def main(inargs):
    """Run the program."""

    print('Input file: ', inargs.infile)
    print('Output file: ', inargs.outfile)


if __name__ == '__main__':

    description='Print the input arguments to the screen.'
    parser = argparse.ArgumentParser(description=description)
    
    parser.add_argument("infile", type=str, help="Input file name")
    parser.add_argument("outfile", type=str, help="Output file name")

    args = parser.parse_args()            
    main(args)
```

By running `script_template.py` at the command line
we'll see that `argparse` handles all the input arguments:

```shell
$ python script_template.py in.csv out.png
```

```text
Input file:  in.csv
Output file:  out.png
```

It also generates help information for the user:

```shell
$ python script_template.py -h
```

```text
usage: script_template.py [-h] infile outfile

Print the input arguments to the screen.

positional arguments:
  infile      Input file name
  outfile     Output file name

optional arguments:
  -h, --help  show this help message and exit
```

and issues errors when users give the program invalid arguments:

```shell
$ python script_template.py in.csv
```

```text
usage: script_template.py [-h] infile outfile
script_template.py: error: the following arguments are required: outfile
```

## Adding our code {#rse-py-scripting-code}

Using this template as a starting point,
we can add the functions we developed previously to a script called...

TODO: In this section we will take the code developed in the novice book
and turn it into a Python script that can be executed at the command line.
See [this lesson](https://carpentrieslab.github.io/python-aos-lesson/04-cmdline/index.html)
for an example of this approach in action.


## Key Points {#rse-pyscripts-keypoints}

```{r, child="keypoints/rse-py-scripting.md"}
```

```{r, child="./links.md"}
```