# Git at the command line {#rse-cmdline-git}

## Questions {#rse-cmdline-git-questions}

```{r, child="questions/rse-cmdline-git.md"}
```

## Introduction {#rse-cmdline-git-intro}

In a previous lesson (TODO: provide link),
we learned the basics of Git using a graphical interface (GitKraken or R Studio).
Like many computer programs,
using Git at the command line is a potentially more powerful option,
as it allows you to access all Git's features.
This chapter describes how the basic "add, commit, push, pull" cycle works at the command line.


## Setting up Git {#rse-cmdline-git-setup}

On a command line, Git commands are written as `git verb options`,
where `verb` is what we actually want to do
and `options` is additional optional information which may or may not be required for the `verb`.

Following this syntax,
the first thing we need to do is configure Git:

```shell
$ git config --global user.name "Vlad Dracula"
$ git config --global user.email "vlad@tran.sylvan.ia"
```

(Please use your own name and email address instead of Dracula's.)

For these lessons we will be interacting with [GitHub](https://github.com/),
so the email address should be the same as the one used when setting up your GitHub account.

The commands we just ran above only need to be run once: the flag `--global` tells Git
to use the settings for every project in your user account on this computer.
You can re-run them at any time if you want to change your details.

You can also check your settings at any time:

```shell
$ git config --list
```

> **Git Help and Manual**
>
> Always remember that if you forget a `git` command,
> you can access the list of commands by using `-h` and access the Git manual by using `--help` :
>
> ```shell
> $ git config -h
> $ git config --help
> ```


## Creating a repository {#rse-cmdline-git-repos}

Once Git is configured, we can start using it.
This section will cover how to create a Git repository associated with a project folder.

We will continue with the story of Wolfman and Dracula who are investigating if it
is possible to send a planetary lander to Mars. 

*TODO: Figure - Motivating Example [../fig/motivatingexample.png](https://github.com/swcarpentry/git-novice/blob/gh-pages/fig/motivatingexample.png)*

First, let's create a directory in `Desktop` folder for our work and then move into that directory:

```shell
$ cd ~/Desktop
$ mkdir planets
$ cd planets
```

Then we tell Git to make `planets` a repository — a place where Git can store versions of our files:

```shell
$ git init
```

If we use `ls` to show the directory's contents,
it appears that nothing has changed:

```shell
$ ls
```

But if we add the `-a` flag to show everything,
we can see that Git has created a hidden directory within `planets` called `.git`:

```shell
$ ls -a
```

```text
.	..	.git
```

Git uses this special sub-directory to store all the information about the project, 
including all files and sub-directories located within the project's directory.
If we ever delete the `.git` sub-directory,
we will lose the project's history.

We can check that everything is set up correctly
by asking Git to tell us the status of our project:

```shell
$ git status
```

```text
# On branch master
#
# Initial commit
#
nothing to commit (create/copy files and use "git add" to track)
```


## Tracking changes {#rse-cmdline-git-changes}

Now that we have a folder and repository for our work,
we'll want to add files to it and track those changes using Git.
Most of the commands we'll use (e.g. `git add`, `git commit`) will sound familiar,
as the same terminology is used in graphical interfaces like GitKraken and R Studio.

To start, let's make sure we're still in the right directory.
You should be in the `planets` directory.

```shell
$ cd ~/Desktop/planets
```

Let's create a file called `mars.txt` that contains some notes
about the Red Planet's suitability as a base.

Using any text editor you like,
type the text below into the `mars.txt` file:

```text
Cold and dry, but everything is my favorite color
```

If we check the status of our project again,
Git tells us that it's noticed the new file:

```shell
$ git status
```

```text
On branch master

Initial commit

Untracked files:
   (use "git add <file>..." to include in what will be committed)

	mars.txt
nothing added to commit but untracked files present (use "git add" to track)
```

The "untracked files" message means that there's a file in the directory
that Git isn't keeping track of.
We can tell Git to track a file using `git add`:

```shell
$ git add mars.txt
```

and then check that the right thing happened:

```shell
$ git status
```

```text
On branch master

Initial commit

Changes to be committed:
  (use "git rm --cached <file>..." to unstage)

	new file:   mars.txt

```

Git now knows that it's supposed to keep track of `mars.txt`,
but it hasn't recorded these changes as a commit yet.
To get it to do that,
we need to run one more command:

```shell
$ git commit -m "Start notes on Mars as a base"
```

```text
[master (root-commit) f22b25e] Start notes on Mars as a base
 1 file changed, 1 insertion(+)
 create mode 100644 mars.txt
```

When we run `git commit`,
Git takes everything we have told it to save by using `git add`
and stores a copy permanently inside the special `.git` directory.
This permanent copy is called a **commit**
(or **revision**) and its short identifier is `f22b25e`.
Your commit may have another identifier.

We use the `-m` flag (for "message")
to record a short, descriptive, and specific comment that will help us remember later on what we did and why.
If we just run `git commit` without the `-m` option,
Git will launch `nano`
(or whatever other default editor is configured using `git config --global core.editor`)
so that we can write a longer message.

If we run `git status` now:

```shell
$ git status
```

```text
On branch master
nothing to commit, working directory clean
```

it tells us everything is up to date.
If we want to know what we've done recently,
we can ask Git to show us the project's history using `git log`:

```shell
$ git log
```

```text
commit f22b25e3233b4645dabd0d81e651fe074bd8e73b
Author: Vlad Dracula <vlad@tran.sylvan.ia>
Date:   Thu Aug 22 09:51:46 2013 -0400

    Start notes on Mars as a base
```

`git log` lists all commits  made to a repository in reverse chronological order.
The listing for each commit includes
the commit's full identifier
(which starts with the same characters as
the short identifier printed by the `git commit` command earlier),
the commit's author,
when it was created,
and the log message Git was given when the commit was created.

Now suppose Dracula adds more information to the file:

```text
Cold and dry, but everything is my favorite color
The two moons may be a problem for Wolfman
```

When we run `git status` now,
it tells us that a file it already knows about has been modified:

```shell
$ git status
```

```text
On branch master
Changes not staged for commit:
  (use "git add <file>..." to update what will be committed)
  (use "git checkout -- <file>..." to discard changes in working directory)

	modified:   mars.txt

no changes added to commit (use "git add" and/or "git commit -a")
```

The last line is the key phrase:
"no changes added to commit".
We have changed this file,
but we haven't told Git we will want to save those changes
(which we do with `git add`)
nor have we saved them (which we do with `git commit`).
So let's do that now. It is good practice to always review
our changes before saving them. We do this using `git diff`.
This shows us the differences between the current state
of the file and the most recently saved version:

```shell
$ git diff
```

```text
diff --git a/mars.txt b/mars.txt
index df0654a..315bf3a 100644
--- a/mars.txt
+++ b/mars.txt
@@ -1 +1,2 @@
 Cold and dry, but everything is my favorite color
+The two moons may be a problem for Wolfman
```

The output is cryptic because
it is actually a series of commands for tools like editors and `patch`
telling them how to reconstruct one file given the other.
If we break it down into pieces:

1.  The first line tells us that Git is producing output similar to the Unix `diff` command
    comparing the old and new versions of the file.
2.  The second line tells exactly which versions of the file
    Git is comparing;
    `df0654a` and `315bf3a` are unique computer-generated labels for those versions.
3.  The third and fourth lines once again show the name of the file being changed.
4.  The remaining lines are the most interesting, they show us the actual differences
    and the lines on which they occur.
    In particular,
    the `+` marker in the first column shows where we added a line.

After reviewing our change, it's time to commit it:

```shell
$ git commit -m "Add concerns about effects of Mars' moons on Wolfman"
$ git status
```

```text
On branch master
Changes not staged for commit:
  (use "git add <file>..." to update what will be committed)
  (use "git checkout -- <file>..." to discard changes in working directory)

	modified:   mars.txt

no changes added to commit (use "git add" and/or "git commit -a")
```

Whoops:
Git won't commit because we didn't use `git add` first.
Let's fix that:

```shell
$ git add mars.txt
$ git commit -m "Add concerns about effects of Mars' moons on Wolfman"
```

```text
[master 34961b1] Add concerns about effects of Mars' moons on Wolfman
 1 file changed, 1 insertion(+)
```

> **Ignoring things**
>
> We're only working with one file so far (`mars.txt`),
> but what if we had lots of files in this directory?
> We might want to have git track our text (`.txt`) files, for instance,
> but not our data files (`.dat`).
>
> So that those data files don't show up as "untracked files" every time we call `git status`,
> we can create a file in the root directory of our project called `.gitignore`.
> In this file we can list all the files we want git to ignore.
> For example, we could add the line `*.dat` to `.gitignore` to ignore all the data files.
> 
> Don't forget to add and commit the `.gitignore` file to your repository so that git
> knows to use it!


## Remotes in GitHub {#rse-cmdline-git-remotes}

One of the first things we did when learning Git via GitKraken or R Studio
was to link our local repository with an external hosting service (GitHub).
We will do that again here,
so login to GitHub and create a new repository called `planets`.

The next step is to connect the two repositories.
We do this by making the GitHub repository a remote for the local repository.
The home page of the repository on GitHub includes the string we need to
identify it:

*TODO: Figure - Where to Find Repository URL on GitHub [../fig/github-find-repo-string.png](https://github.com/swcarpentry/git-novice/blob/gh-pages/fig/github-find-repo-string.png)*

Click on the 'HTTPS' link to change the protocol from SSH to HTTPS.

> **HTTPS vs. SSH**
>
> We use HTTPS here because it does not require additional configuration.  After
> the workshop you may want to set up SSH access, which is a bit more secure, by
> following one of the great tutorials from
> [GitHub](https://help.github.com/articles/generating-ssh-keys),
> [Atlassian/Bitbucket](https://confluence.atlassian.com/bitbucket/set-up-ssh-for-git-728138079.html)
> and [GitLab](https://about.gitlab.com/2014/03/04/add-ssh-key-screencast/)
> (this one has a screencast).

*TODO: Figure - Changing the Repository URL on GitHub [../fig/github-change-repo-string.png](https://github.com/swcarpentry/git-novice/blob/gh-pages/fig/github-change-repo-string.png)*

Copy that URL from the browser, go into the local `planets` repository, and run
this command:

```shell
$ git remote add origin https://github.com/vlad/planets.git
```

(Make sure to use the URL for your repository rather than Vlad's: the only
difference should be your username instead of `vlad`.)

`origin` is a local name used to refer to the remote repository. It could be called
anything, but `origin` is a convention that is often used by default in Git
and GitHub, so it's helpful to stick with this unless there's a reason not to.

We can check that the command has worked by running `git remote -v`:

```shell
$ git remote -v
```

```text
origin   https://github.com/vlad/planets.git (push)
origin   https://github.com/vlad/planets.git (fetch)
```

Once the remote is set up, this command will push the changes from
our local repository to the repository on GitHub:

```shell
$ git push origin master
```

```text
Enumerating objects: 16, done.
Counting objects: 100% (16/16), done.
Delta compression using up to 8 threads.
Compressing objects: 100% (11/11), done.
Writing objects: 100% (16/16), 1.45 KiB | 372.00 KiB/s, done.
Total 16 (delta 2), reused 0 (delta 0)
remote: Resolving deltas: 100% (2/2), done.
To https://github.com/vlad/planets.git
 * [new branch]      master -> master
```

We can pull changes from the remote repository to the local one as well:

```shell
$ git pull origin master
```

```text
From https://github.com/vlad/planets
 * branch            master     -> FETCH_HEAD
Already up-to-date.
```

Pulling has no effect in this case because the two repositories are already
synchronized.


## Exploring history {#rse-cmdline-git-history}

Now that we've synced our local repository with GitHub, we can resume work on the planets project.

At this stage, we're starting to build up a history of changes to the project.
Using Git, we have the ability to look at previous versions of the project files, and,
if desired, restore a specific file to a previous state.
To do this, we need to use an identifier to indicate which version of the file we want to examine/use.
As we saw in a previous section, we can refer to commits by their identifiers.
You can refer to the most recent commit of the working directory by using the identifier `HEAD`.

For example, let's make a change to `mars.txt` adding yet another line.

```text
Cold and dry, but everything is my favorite color
The two moons may be a problem for Wolfman
An ill-considered change
```

We can now use `HEAD` to see the change we made:

```shell
$ git diff HEAD mars.txt
```

```text
diff --git a/mars.txt b/mars.txt
index b36abfd..0848c8d 100644
--- a/mars.txt
+++ b/mars.txt
@@ -1,3 +1,4 @@
 Cold and dry, but everything is my favorite color
 The two moons may be a problem for Wolfman
+An ill-considered change.
```

which is the same as what you would get if you leave out `HEAD` (try it).
The real goodness in all this is when referring to previous commits.
We do that by adding `~1` 
(where "~" is "tilde", pronounced [**til**-d*uh*]) 
to refer to the commit one before `HEAD`.

```shell
$ git diff HEAD~1 mars.txt
```

If we want to see the differences between older commits we can use `git diff`
again, but with the notation `HEAD~1`, `HEAD~2`, and so on, to refer to them:

```shell
$ git diff HEAD~2 mars.txt
```

```text
diff --git a/mars.txt b/mars.txt
index df0654a..b36abfd 100644
--- a/mars.txt
+++ b/mars.txt
@@ -1 +1,4 @@
 Cold and dry, but everything is my favorite color
+The two moons may be a problem for Wolfman
+An ill-considered change
```

We could also use `git show` which shows us what changes we made at an older commit as 
well as the commit message, rather than the _differences_ between a commit and our 
working directory that we see by using `git diff`.

```shell
$ git show HEAD~2 mars.txt
```

```text
commit f22b25e3233b4645dabd0d81e651fe074bd8e73b
Author: Vlad Dracula <vlad@tran.sylvan.ia>
Date:   Thu Aug 22 09:51:46 2013 -0400

    Start notes on Mars as a base

diff --git a/mars.txt b/mars.txt
new file mode 100644
index 0000000..df0654a
--- /dev/null
+++ b/mars.txt
@@ -0,0 +1 @@
+Cold and dry, but everything is my favorite color
```

We can also refer to commits using
those long strings of digits and letters
that `git log` displays.
These are unique IDs for the changes,
and "unique" really does mean unique:
every change to any set of files on any computer
has a unique 40-character identifier.
Our first commit was given the ID
`f22b25e3233b4645dabd0d81e651fe074bd8e73b`,
so let's try this:

```shell
$ git diff f22b25e3233b4645dabd0d81e651fe074bd8e73b mars.txt
```

```text
diff --git a/mars.txt b/mars.txt
index df0654a..93a3e13 100644
--- a/mars.txt
+++ b/mars.txt
@@ -1 +1,4 @@
 Cold and dry, but everything is my favorite color
+The two moons may be a problem for Wolfman
+An ill-considered change
```

That's the right answer,
but typing out random 40-character strings is annoying,
so Git lets us use just the first few characters:

```shell
$ git diff f22b25e mars.txt
```

```text
diff --git a/mars.txt b/mars.txt
index df0654a..93a3e13 100644
--- a/mars.txt
+++ b/mars.txt
@@ -1 +1,4 @@
 Cold and dry, but everything is my favorite color
+The two moons may be a problem for Wolfman
+An ill-considered change
```

All right!
So we can save changes to files and see what we've changed—now how
can we restore older versions of things?
Let's suppose we change our mind about the last update to
`mars.txt` (the "ill-considered change").

`git status` now tells us that the file has been changed,
but those changes haven't been staged:

```shell
$ git status
```

```text
On branch master
Changes not staged for commit:
  (use "git add <file>..." to update what will be committed)
  (use "git checkout -- <file>..." to discard changes in working directory)

    modified:   mars.txt

no changes added to commit (use "git add" and/or "git commit -a")
```

We can put things back the way they were
by using `git checkout`:

```shell
$ git checkout HEAD mars.txt
$ cat mars.txt
```

```text
Cold and dry, but everything is my favorite color
The two moons may be a problem for Wolfman
```

As you might guess from its name,
`git checkout` checks out (i.e., restores) an old version of a file.
In this case,
we're telling Git that we want to recover the version of the file recorded in `HEAD`,
which is the last saved commit.
If we want to go back even further,
we can use a commit identifier instead:

```shell
$ git checkout f22b25e mars.txt
```

```shell
$ cat mars.txt
```

```text
Cold and dry, but everything is my favorite color
```

```shell
$ git status
```

```text
On branch master
Changes to be committed:
  (use "git reset HEAD <file>..." to unstage)

    modified:   mars.txt

```

Notice that the changes are on the staged area.
Again, we can put things back the way they were
by using `git checkout`:

```shell
$ git checkout HEAD mars.txt
```

*TODO: Figure - Git Checkout [../fig/git-checkout.svg](https://github.com/swcarpentry/git-novice/blob/gh-pages/fig/git-checkout.svg)*


## Exercises {#rse-cmdline-git-exercises}

### Creating a repository

**Places to Create Git Repositories**

Along with tracking information about planets (the project we have already created), 
Dracula would also like to track information about moons.
Despite Wolfman's concerns, Dracula creates a `moons` project inside his `planets` 
project with the following sequence of commands:

```shell
$ cd ~/Desktop   # return to Desktop directory
$ cd planets     # go into planets directory, which is already a Git repository
$ ls -a          # ensure the .git sub-directory is still present in the planets directory
$ mkdir moons    # make a sub-directory planets/moons
$ cd moons       # go into moons sub-directory
$ git init       # make the moons sub-directory a Git repository
$ ls -a          # ensure the .git sub-directory is present indicating we have created a new Git repository
```

Is the `git init` command, run inside the `moons` sub-directory, required for 
tracking files stored in the `moons` sub-directory?
 
> *Solution*
>
> No. Dracula does not need to make the `moons` sub-directory a Git repository 
> because the `planets` repository will track all files, sub-directories, and 
> sub-directory files under the `planets` directory.  Thus, in order to track 
> all information about moons, Dracula only needed to add the `moons` sub-directory
> to the `planets` directory.
> 
> Additionally, Git repositories can interfere with each other if they are "nested":
> the outer repository will try to version-control
> the inner repository. Therefore, it's best to create each new Git
> repository in a separate directory. To be sure that there is no conflicting
> repository in the directory, check the output of `git status`. If it looks
> like the following, you are good to go to create a new repository as shown
> above:
>
> ```shell
> $ git status
> ```
>
> ```text
> fatal: Not a git repository (or any of the parent directories): .git
> ```


**Correcting `git init` Mistakes**

Wolfman explains to Dracula how a nested repository is redundant and may cause confusion
down the road. Dracula would like to remove the nested repository. How can Dracula undo 
his last `git init` in the `moons` sub-directory?

> *Background*
> 
> Removing files from a git repository needs to be done with caution. To remove files from the working tree and not from your working directory, use
> 
> ```shell
> $ rm filename
> ```
> 
> The file being removed has to be in sync with the branch head with no updates.
> If there are updates, the file can be removed by force by using the `-f` option.
> Similarly a directory can be removed from git using `rm -r dirname` or `rm -rf dirname`.
>
> *Solution - USE WITH CAUTION*
>
> Git keeps all of its files in the `.git` directory.
> To recover from this little mistake, Dracula can just remove the `.git`
> folder in the moons subdirectory by running the following command from inside the `planets` directory:
>
> ```shell
> $ rm -rf moons/.git
> ```
>
> But be careful! Running this command in the wrong directory, will remove
> the entire Git history of a project you might want to keep.
> Therefore, always check your current directory using the command `pwd`.


### Tracking changes

**Committing Changes to Git**

Which command(s) below would save the changes of `myfile.txt`
to my local Git repository?

1. ```shell
   $ git commit -m "my recent changes"
   ```

2. ```shell
   $ git init myfile.txt
   $ git commit -m "my recent changes"
   ```

3. ```shell
   $ git add myfile.txt
   $ git commit -m "my recent changes"
   ```

4. ```shell
   $ git commit -m myfile.txt "my recent changes"
   ```

> *Solution*
>
> 1. Would only create a commit if files have already been staged.  
> 2. Would try to create a new repository.  
> 3. Is correct: first add the file to the staging area, then commit.  
> 4. Would try to commit a file "my recent changes" with the message myfile.txt.  


**Committing Multiple Files**

The staging area can hold changes from any number of files
that you want to commit as a single snapshot.

1. Add some text to `mars.txt` noting your decision
to consider Venus as a base

2. Create a new file `venus.txt` with your initial thoughts
about Venus as a base for you and your friends

3. Add changes from both files to the staging area,
and commit those changes.

> *Solution*
>
> First we make our changes to the `mars.txt` and `venus.txt` files:  
> ```shell
> $ nano mars.txt
> $ cat mars.txt
> ```
>
> ```text
> Maybe I should start with a base on Venus.
> ```
>
> ```shell
> $ nano venus.txt
> $ cat venus.txt
> ```
>
> ```text
> Venus is a nice planet and I definitely should consider it as a base.
> ```
>
> Now you can add both files to the staging area. We can do that in one line:
>
> ```shell
> $ git add mars.txt venus.txt
> ```
> 
> Or with multiple commands:  
>
> ```shell
> $ git add mars.txt
> $ git add venus.txt
> ```
>
> Now the files are ready to commit. You can check that using `git status`. If you are ready to commit use:
>
> ```shell
> $ git commit -m "Write plans to start a base on Venus"
> ```
> ```text
> [master cc127c2]
> Write plans to start a base on Venus
> 2 files changed, 2 insertions(+)
> create mode 100644 venus.txt
> ```

**`bio` Repository**

* Create a new Git repository on your computer called `bio`.  
* Write a three-line biography for yourself in a file called `me.txt`, commit your changes. 
* Modify one line, add a fourth line. 
* Display the differences between its updated state and its original state.  

> *Solution*
>
> If needed, move out of the `planets` folder:
>
> ```shell
> $ cd ..
> ```
>
> Create a new folder called `bio` and 'move' into it:
>
> ```shell
> $ mkdir bio
> $ cd bio
> ```
>
> Initialise git:
>
> ```shell
> $ git init
> ```
>
> Create your biography file `me.txt` using `nano` or another text editor.
> Once in place, add and commit it to the repository:
>
> ```shell
> $ git add me.txt
> $ git commit -m'Adding biography file'
> ```
>
> Modify the file as described (modify one line, add a fourth line).
> To display the differences
> between its updated state and its original state, use `git diff`:
>
> ```shell
> $ git diff me.txt
> ```

### Ignoring things

**Ignoring Nested Files**

Given a directory structure that looks like:

```shell
results/data
results/plots
```

How would you ignore only `results/plots` and not `results/data`?

> *Solution*
>
> If you only want to ignore the contents of
> `results/plots`, you can change your `.gitignore` to ignore
> only the `/plots/` subfolder by adding the following line to
> your .gitignore:
> 
> ```text
> results/plots/
> ```
>
> This line will ensure only the contents of `results/plots` is ignored, and
> not the contents of `results/data`.
> 
> As with most programming issues, there
> are a few alternative ways that one may ensure this ignore rule is followed.
> The "Ignoring Nested Files: Variation" exercise has a slightly
> different directory structure
> that presents an alternative solution.
> Further, the discussion page has more detail on ignore rules.


**Including Specific Files**

How would you ignore all `.dat` files in your root directory except for `final.dat`?

Hint: Find out what `!` (the exclamation point operator) does.

> *Solution*
> 
> You would add the following two lines to your .gitignore:
> 
> ```text
> *.dat           # ignore all data files
> !final.dat      # except final.data
> ```
>
> The exclamation point operator will include a previously excluded entry.
> 
> Note also that because you've previously committed `.dat` files in this
> lesson they will not be ignored with this new rule. Only future additions
> of `.dat` files added to the root directory will be ignored.


### Remotes in GitHub

**GitHub GUI**

Browse to your `planets` repository on GitHub.
Under the Code tab, find and click on the text that says "XX commits" (where "XX" is some number).
Hover over, and click on, the three buttons to the right of each commit.
What information can you gather/explore from these buttons?
How would you get that same information in the shell?

> *Solution*
>
> The left-most button (with the picture of a clipboard) copies the full identifier of the commit 
> to the clipboard. In the shell, ```git log``` will show you the full commit identifier for each 
> commit.
> 
> When you click on the middle button, you'll see all of the changes that were made in that 
> particular commit. Green shaded lines indicate additions and red ones removals. In the shell we 
> can do the same thing with ```git diff```. In particular, ```git diff ID1..ID2``` where ID1 and 
> ID2 are commit identifiers (e.g. ```git diff a3bf1e5..041e637```) will show the differences 
> between those two commits.
> 
> The right-most button lets you view all of the files in the repository at the time of that 
> commit. To do this in the shell, we'd need to checkout the repository at that particular time. 
> We can do this with ```git checkout ID``` where ID is the identifier of the commit we want to 
> look at. If we do this, we need to remember to put the repository back to the right state 
> afterwards!


**GitHub Timestamp**

Create a remote repository on GitHub. Push the contents of your local
repository to the remote. Make changes to your local repository and push these
changes. Go to the repo you just created on GitHub and check the
timestamps of the files. How does GitHub record times, and why?

> *Solution*
>
> GitHub displays timestamps in a human readable relative format (i.e. "22 hours ago" or "three 
> weeks ago"). However, if you hover over the timestamp, you can see the exact time at which the 
> last change to the file occurred.


**Push vs. Commit**

In this lesson, we introduced the "git push" command.
How is "git push" different from "git commit"?

> *Solution*
>
> When we push changes, we're interacting with a remote repository to update it with the changes 
> we've made locally (often this corresponds to sharing the changes we've made with others). 
> Commit only updates your local repository.


**GitHub License and README files**

In this section we learned about creating a remote repository on GitHub, but when you initialized 
your GitHub repo, you didn't add a README.md or a license file. If you had, what do you think 
would have happened when you tried to link your local and remote repositories?

> *Solution*
>
> In this case, we'd see a merge conflict due to unrelated histories. When GitHub creates a 
> README.md file, it performs a commit in the remote repository. When you try to pull the remote 
> repository to your local repository, Git detects that they have histories that do not share a 
> common origin and refuses to merge.
>
> ```shell
> $ git pull origin master
> ```
>
> ```text
> warning: no common commits
> remote: Enumerating objects: 3, done.
> remote: Counting objects: 100% (3/3), done.
> remote: Total 3 (delta 0), reused 0 (delta 0), pack-reused 0
> Unpacking objects: 100% (3/3), done.
> From https://github.com/vlad/planets
>  * branch            master     -> FETCH_HEAD
>  * [new branch]      master     -> origin/master
> fatal: refusing to merge unrelated histories
> ```
>
> You can force git to merge the two repositories with the option `--allow-unrelated-histories`. 
> Be careful when you use this option and carefully examine the contents of local and remote 
> repositories before merging.
>
> ```shell
> $ git pull --allow-unrelated-histories origin master
> ```
> 
> ```text
> From https://github.com/vlad/planets
>  * branch            master     -> FETCH_HEAD
> Merge made by the 'recursive' strategy.
> README.md | 1 +
> 1 file changed, 1 insertion(+)
> create mode 100644 README.md
> ```


### Exploring history


**Recovering Older Versions of a File**

Jennifer has made changes to the Python script that she has been working on for weeks, and the
modifications she made this morning "broke" the script and it no longer runs. She has spent
~ 1hr trying to fix it, with no luck...

Luckily, she has been keeping track of her project's versions using Git! Which commands below will
let her recover the last committed version of her Python script called
`data_cruncher.py`?

1. `$ git checkout HEAD`

2. `$ git checkout HEAD data_cruncher.py`

3. `$ git checkout HEAD~1 data_cruncher.py`

4. `$ git checkout <unique ID of last commit> data_cruncher.py`

5. Both 2 and 4

> *Solution*
>
> The answer is (5)-Both 2 and 4. 
> 
> The `checkout` command restores files from the repository, overwriting the files in your working 
> directory. Answers 2 and 4 both restore the *latest* version *in the repository* of the file 
> `data_cruncher.py`. Answer 2 uses `HEAD` to indicate the *latest*, whereas answer 4 uses the 
> unique ID of the last commit, which is what `HEAD` means. 
> 
> Answer 3 gets the version of `data_cruncher.py` from the commit *before* `HEAD`, which is NOT 
> what we wanted.
> 
> Answer 1 can be dangerous! Without a filename, `git checkout` will restore **all files** 
> in the current directory (and all directories below it) to their state at the commit specified. 
> This command will restore `data_cruncher.py` to the latest commit version, but it will also 
> restore *any other files that are changed* to that version, erasing any changes you may 
> have made to those files!
> As discussed above, you are left in a *detached* `HEAD` state, and you don't want to be there.


**Reverting a Commit**

Jennifer is collaborating on her Python script with her colleagues and
realizes her last commit to the project's repository contained an error and
she wants to undo it.  `git revert [erroneous commit ID]` will create a new 
commit that reverses Jennifer's erroneous commit. Therefore `git revert` is
different to `git checkout [commit ID]` because `git checkout` returns the
files within the local repository to a previous state, whereas `git revert`
reverses changes committed to the local and project repositories.  
Below are the right steps and explanations for Jennifer to use `git revert`,
what is the missing command?

1. `________ # Look at the git history of the project to find the commit ID`

2. Copy the ID (the first few characters of the ID, e.g. 0b1d055).

3. `git revert [commit ID]`

4. Type in the new commit message.

5. Save and close


**Understanding Workflow and History**

What is the output of the last command in

```
$ cd planets
$ echo "Venus is beautiful and full of love" > venus.txt
$ git add venus.txt
$ echo "Venus is too hot to be suitable as a base" >> venus.txt
$ git commit -m "Comment on Venus as an unsuitable base"
$ git checkout HEAD venus.txt
$ cat venus.txt #this will print the contents of venus.txt to the screen
```

1. ```text
   Venus is too hot to be suitable as a base
   ```

2. ```
   Venus is beautiful and full of love
   ```

3. ```
   Venus is beautiful and full of love
   Venus is too hot to be suitable as a base
   ```

4. ```
   Error because you have changed venus.txt without committing the changes
   ```

> *Solution*
>
> The answer is 2. 
>  
> The command `git add venus.txt` places the current version of `venus.txt` into the staging area. 
> The changes to the file from the second `echo` command are only applied to the working copy, 
> not the version in the staging area.
>  
> So, when `git commit -m "Comment on Venus as an unsuitable base"` is executed, 
> the version of `venus.txt` committed to the repository is the one from the staging area and
> has only one line.
>   
> At this time, the working copy still has the second line (and 
> `git status` will show that the file is modified). However, `git checkout HEAD venus.txt` 
> replaces the working copy with the most recently committed version of `venus.txt`.
>   
>  So, `cat venus.txt` will output 
>  
>  ```text
>  Venus is beautiful and full of love.
>  ```


**Checking Understanding of `git diff`**

Consider this command: `git diff HEAD~9 mars.txt`. What do you predict this command
will do if you execute it? What happens when you do execute it? Why?

Try another command, `git diff [ID] mars.txt`, where [ID] is replaced with
the unique identifier for your most recent commit. What do you think will happen,
and what does happen?


**Getting Rid of Staged Changes**

`git checkout` can be used to restore a previous commit when unstaged changes have
been made, but will it also work for changes that have been staged but not committed?
Make a change to `mars.txt`, add that change, and use `git checkout` to see if
you can remove your change.


**Explore and Summarize Histories**

Exploring history is an important part of Git, often it is a challenge to find
the right commit ID, especially if the commit is from several months ago.

Imagine the `planets` project has more than 50 files.
You would like to find a commit with specific text in `mars.txt` is modified.
When you type `git log`, a very long list appeared,
How can you narrow down the search?

Recall that the `git diff` command allow us to explore one specific file,
e.g. `git diff mars.txt`. We can apply a similar idea here.

```shell
$ git log mars.txt
```

Unfortunately some of these commit messages are very ambiguous e.g. `update files`.
How can you search through these files?

Both `git diff` and `git log` are very useful and they summarize a different part of the history 
for you.
Is it possible to combine both? Let's try the following:

```shell
$ git log --patch mars.txt
```

You should get a long list of output, and you should be able to see both commit messages and 
the difference between each commit.

Question: What does the following command do?

```shell
$ git log --patch HEAD~9 *.txt
```


## Key Points {#rse-cmdline-git-keypoints}

```{r, child="keypoints/rse-cmdline-git.md"}
```

```{r, child="./links.md"}
```


