# Bash Shell {#bash}

## Questions {#bash-questions}

```{r, child="questions/bash.md"}
```

## Objectives {#bash-objectives}

```{r, child="objectives/bash.md"}
```

## Introduction {#bash-intro}

At a high level, computers do four things:

-   run programs
-   store data
-   communicate with each other, and
-   interact with us

They can do the last of these in many different ways,
including through a keyboard and mouse, touch screen interfaces, or using speech recognition systems.
While touch and voice interfaces are becoming more commonplace, most interaction is still
done using traditional screens, mice, touchpads and keyboards.

The **graphical user interface** (GUI) is the most widely used way to interact with
personal computers. We give instructions (to run a program, to copy a file, to create
a new folder/directory) with the convenience of a few mouse clicks. This way of interacting
with a computer is intuitive and very easy to learn. But this way of giving instructions to
a computer scales very poorly if we are to give a large stream of instructions even if they
are similar or identical. For example if we have to copy the third line of each of a thousand
text files stored in thousand different directories and paste it into a single file
line by line. Using the tradition GUI approach of clicks will take several hours to do this.

This is where we take advange of the shell - a **command-line interface**
to make such repetitive tasks automatic and fast. It can take a single instruction and
repeat it as is or with some modification as many times as we want. The task in
the example above can be accomplished in a few minutes at most.

The heart of a command-line interface is a **read-evaluate-print loop** (REPL). It is called
so because when you type a command and press <kbd>Return</kbd> (also known as <kbd>Enter</kbd>) the shell
reads your command,
evaluates (or "executes") it,
prints the output of your command,
loops back and waits for you to enter another command.

## The Shell

The Shell is a program which runs other programs rather than doing calculations itself.
Those programs can be as complicated as climate modeling software and as simple as a
program that creates a new directory. The simple programs which are used to perform
stand alone tasks are usually refered to as commands.
The most popular Unix shell is Bash, (the Bourne Again SHell --- so-called because
it's derived from a shell written by Stephen Bourne).
Bash is the default shell on most modern implementations of Unix
and in most packages that provide Unix-like tools for Windows.

When the shell is first opened, you are presented with a **prompt**,
indicating that the shell is waiting for input.

```shell
$
```

The shell typically uses `$ ` as the prompt, but may use a different symbol.
In the examples for this lesson, we'll show the prompt as `$ `.
Most importantly:
when typing commands, either from these lessons or from other sources,
*do not type the prompt*, only the commands that follow it.

So let's try our first command, which will list the contents of the current directory:

```shell
$ ls
```

```text
Desktop     Downloads   Movies      Pictures
Documents   Library     Music       Public
```

### Is it difficult?

It is a different model of interacting than a GUI, and that
will take some effort - and some time - to learn. A GUI
presents you with choices and you select one. With a **command line interface** (CLI) the choices are combinations
of commands and parameters, more like words in a language than buttons on a screen. They
are not presented to you so
you must learn a few, like learning some vocabulary in a new language. But a small
number of commands gets you a long way, and we'll cover those essential few today.

### Flexibility and automation

The grammar of a shell allows you to combine existing tools into powerful
pipelines and handle large volumes of data automatically. Sequences of
commands can be written into a *script*, improving the reproducibility of
workflows and allowing you to repeat them easily.

In addition, the command line is often the easiest way to interact with remote machines and supercomputers.
Familiarity with the shell is near essential to run a variety of specialized tools and resources
including high-performance computing systems.
As clusters and cloud computing systems become more popular for scientific data crunching,
being able to interact with the shell is becoming a necessary skill.
We can build on the command-line skills covered here
to tackle a wide range of scientific questions and computational challenges.


## Navigating Files and Directories

The part of the operating system responsible for managing files and directories
is called the **file system**.
It organizes our data into files,
which hold information,
and directories (also called "folders"),
which hold files or other directories.

Several commands are frequently used to create, inspect, rename, and delete files and directories.
To start exploring them, we'll go to our open shell window.

First let's find out where we are by running a command called `pwd`
(which stands for "print working directory"). Directories are like *places* - at any time
while we are using the shell we are in exactly one place, called
our **current working directory**. Commands mostly read and write files in the
current working directory, i.e. "here", so knowing where you are before running
a command is important. `pwd` shows you where you are:

```shell
$ pwd
```

```text
/Users/nelle
```




## Summary {#integrate-summary}

TODO: summary for bash. 

## Exercises {#bash-exercises}

TODO: exercises for bash.

## Key Points {#integrate-keypoints}

```{r, child="keypoints/bash.md"}
```

```{r, child="etc/links.md"}
```