# Advanced Git {#rse-advanced-git}

## Questions {#rse-advanced-git-questions}

```{r, child="questions/rse-advanced-git.md"}
```

## Introduction {#rse-advanced-git-intro}

Now that we're comfortable using Git at the command line,
we can look at two important concepts for keeping our code development organised
and for working with others: branching and pull requests. 


## Branches {#rse-advanced-git-branches}

*FIXME: This lesson is from [UW-Madison](https://uw-madison-datascience.github.io/git-novice-custom/07-branches/)
and therefore doesn't quite match the storyline of the other lessons.*

In the Git lessons so far we've been working in a straight timeline.
However, there are times when we might want to keep
our main work safe from experimental changes we are working on.
To do this we can use branches to work on separate tasks in parallel
without changing our current branch, `master`.

We didn't see it before but the first branch made is called `master`.
This is the default branch created when initializing a repository and
is often considered to be the "clean" or "working" version of a
repository's code.

We can see what branches exist in a repository by typing

```shell
$ git branch
```

```text
* master
```

The '*' indicates which branch we are currently on.

In this lesson, Dracula is trying to run an analysis
and doesn't know if it will be faster in bash or python.
To keep his master branch safe he will use separate branches
for both bash and python analysis.
Then he will merge the branch with the faster script
into his master branch.

First let's make the python branch.
We use the same `git branch` command but now add the 
name we want to give our new branch

```shell
$ git branch pythondev
```

We can now check our work with the `git branch` command.

```shell
$ git branch
```

```text
* master
  pythondev
```

We can see that we created the `pythondev` branch but we
are still in the master branch.

We can also see this in the output of the `git status` command.

```shell
$ git status
```

```text
On branch master
nothing to commit, working directory clean
```

To switch to our new branch we can use the `checkout` command
we learned earlier and check our work with `git branch`.

```shell
$ git checkout pythondev
$ git branch
```

```text
  master
* pythondev
```

Before we used the `checkout` command to checkout a file from a specific commit
using commit hashes or `HEAD` and the filename (`git checkout HEAD <file>`). The
`checkout` command can also be used to checkout an entire previous version of the
repository, updating all files in the repository to match the state of a desired commit.

Branches allow us to do this using a human-readable name rather than memorizing
a commit hash. This name also typically gives purpose to the set of changes in
that branch. When we use the command `git checkout <branch_name>`, we are using
a nickname to checkout a version of the repository that matches the most recent
commit in that branch (a.k.a. the HEAD of that branch).

Here you can use `git log` and `ls` to see that the history and 
files are the same as our `master` branch. This will be true until
some changes are committed to our new branch.

Now lets make our python script.  
For simplicity sake, we will `touch` the script making an empty file
but imagine we spent hours working on this python script for our analysis.

```shell
$ touch analysis.py
```

Now we can add and commit the script to our branch.

```shell
$ git add analysis.py
$ git commit -m "Wrote and tested python analysis script"
```

```text
[pythondev x792csa1] Wrote and tested python analysis script
 1 file changed, 1 insertion(+)
 create mode 100644 analysis.py
```

Lets check our work!

```shell
$ ls
$ git log --oneline
```

As expected, we see our commit in the log.

Now let's switch back to the `master` branch.

```shell
$ git checkout master
$ git branch
```

```text
* master
  pythondev
```

Let's explore the repository a bit.

Now that we've confirmed we are on the `master` branch again.
Let's confirm that `analysis.py` and our last commit aren't in `master`.

```shell
$ ls
$ git log --oneline
```

We no longer see the file `analysis.py` and our latest commit doesn't
appear in this branch's history. But do not fear! All of our hard work
remains in the `pythondev` branch. We can confirm this by moving back
to that branch.

```shell
$ git checkout pythondev
$ git branch
```

```text
  master
* pythondev
```

```shell
$ ls
$ git log --oneline
```

And we see that our `analysis.py` file and respective commit have been
preserved in the `pythondev` branch.

Now we can repeat the process for our bash script in a branch called
`bashdev`.

First we must checkout the `master` branch again. New branches will
include the entire history up to the current commit, and we'd like
to keep these two tasks separate.

```shell
$ git checkout master
$ git branch
```

```text
* master
  pythondev
```

This time let's create and switch two the `bashdev` branch
in one command.

We can do so by adding the `-b` flag to checkout.

```shell
$ git checkout -b bashdev
$ git branch
```

```text
* bashdev
  master
  pythonndev
```

We can use `ls` and `git log` to see that this branch is 
the same as our current `master` branch.

Now we can make `analysis.sh` and add and commit it.
Again imagine instead of `touch`ing the file we worked 
on it for many hours.

```shell
$ touch analysis.sh
$ git add analysis.sh
$ git commit –m “Wrote and tested bash analysis script”
```

```text
[bashdev 2n779ds] Wrote and tested bash analysis script
 1 file changed, 1 insertion(+)
 create mode 100644 analysis.sh
```

Lets check our work again before we switch back to the master branch.

```shell
$ ls
$ git log --oneline
```

So it turns out the python `analysis.py` is much faster than `analysis.sh`.

Let's merge this version into our `master` branch so we can use it for
our work going forward.

Merging brings the changes from a different branch into 
the current branch.

First we must switch to the branch we're merging changes into, `master`.

```shell
$ git checkout master
$ git branch
```

```text
  bashdev
* master
  pythonndev
```

Now we can `merge` the `pythondev` branch into our current branch
(`master`). In simple terms, this command could be stated as "`git`, please
`merge` the changes in the `pythondev` branch into the current branch
I'm in".

```shell
$ git merge pythondev
```

```text
Updating 12687f6..x792csa1
Fast-forward
 analysis.py | 0
 1 file changed, 0 insertions(+), 0 deletions(-)
 create mode 100644 analysis.py
```

Now that we've merged the `pythondev` into `master`, these changes
exist in both branches. This could be confusing in the future if we
stumble upon the `pythondev` branch again.

We can delete our old branches so as to avoid this confusion later.
We can do so by adding the `-d` flag to the `git branch` command.

```shell
git branch -d pythondev
```

```text
Deleted branch pythondev (was x792csa1).
```

And because we don't want to keep the changes in the `bashdev` branch,
we can delete the `bashdev` branch as well

```shell
$ git branch -d bashdev
```

```text
error: The branch 'bashdev' is not fully merged.
If you are sure you want to delete it, run 'git branch -D bashdev'.
```

Since we've never merged the changes from the `bashdev` branch,
Git warns us about deleting them and tells us to use the `-D` flag instead.

Since we really want to delete this branch we will go ahead and do so.

```shell
git branch -D bashdev
```

```text
Deleted branch bashdev (was 2n779ds).
```


### A branching workflow

Now that we're comfortable with creating and merging branches,
we need to consider when and how we might use branches in our code development workflows.

Experienced developers tend to use a "branch-per-feature" workflow,
even when they are working on their own.
Whenever they want to create a new feature,
they create a new branch from `master` and do the work there.
When it's done,
they merge that branch back into `master.

In the context of what we've learned about branches so far,
our branch-per-feature workflow looks like this:

1.  `git checkout master` to make sure you're working in the `master` branch.

2.  `git pull origin master` to make sure that
    your starting point is up to date
    (i.e., that the changes you committed while working on the cluster yesterday are now in your local copy,
    or that the fix your colleague did last week has been incorporated).

3.  `git checkout -b name-of-feature` to create a new branch.
    The branch's name should be as descriptive as the name of a function
    or of a file containing source code:
    `"stuff"` or `"fixes"` saves you a couple of seconds of typing when you create the branch,
    and will cost you much more time than that later on
    when you're juggling half a dozen branches
    and trying to figure out what's in each.

4.  Do the work you want to do.
    If some other task occurs to you while you're doing it---for example,
    if you're writing a new function and realize that the documentation for some other function should be updated---*don't*
    do that work in this branch just because you happen to be there.
    Instead,
    commit your changes,
    switch back to `master`,
    create a new branch for the other work,
    and carry on.

5.  Periodically,
    you may `git push origin name-of-feature` to save your work to GitHub.

6.  When the work is done,
    you `git pull origin master` to get any changes that anyone else has put in the master copy
    into your branch
    and merge any conflicts.
    This is an important step:
    you want to do the merge *and test that everything still works* in *your* branch,
    not in `master`.

7.  Finally,
    use `git checkout master` to switch back to the `master` branch on your machine
    and `git merge name-of-feature master` to merge your changes into `master`.
    Test it one more time,
    and then `git push origin master` to send your changes to the remote repository
    for everyone else to admire and use.


### What is a feature?

But what is a "feature," exactly?
What's large enough to merit creation of a new branch?
These rules make sense for small projects with or without collaborators:

1.  Anything cosmetic that is only one or two lines long can be done in `master` and committed right away.
    "Cosmetic" means changes to comments or documentation:
    nothing that affects how code runs, not even a simple variable renaming.

2.  A pure addition that doesn't change anything else is a feature and goes into a branch.
    For example,
    if you run a new analysis and save the results,
    that should be done on its own branch
    because it might take several tries to get the analysis to run,
    and you might interrupt yourself several times to fix things that you've discovered aren't working.

3.  Every change to code that someone might want to undo later in one step gets done as a feature.
    For example,
    if a parameter is added to a function,
    every call to the function has to be updated;
    since neither alteration makes sense without the other,
    it's considered a single feature and should be done in one branch.

The hardest thing about using a branch-per-feature workflow is actually doing it for small changes.
As the first point in the list above suggests,
most people are pragmatic about this on small projects;
on large ones,
where dozens of people might be committing,
even the smallest and most innocuous change needs to be in its own branch
so that it can be reviewed (which we discuss below).


## Pull requests  {#rse-advanced-git-pull-requests}

*FIXME: This lesson is from [UW-Madison](https://uw-madison-datascience.github.io/git-novice-custom/10-pull-requests/)
and therefore doesn't quite match the storyline of the other lessons.*

Pull requests are a great way to collaborate with others using GitHub.
Instead of making changes directly to a repository, you can suggest changes via a pull request.
This can be useful if you don't have permission to modify a repository directly or
you want someone else to review your changes.

To be more specific, a pull request is essentially a marker saying,
"Someone would like to merge branch A into branch X".
The pull request does *not* contain the changes:
instead,
it points at two branches,
so that if either branch is changed,
the differences displayed are always up to date.

FIXME: figure

Pull requests can be created between two branches in a single repository,
or between branches of different repositories.
The latter is common in projects with many contributors:
each contributor works in a branch in their own [fork](glossary.html#fork) of the repository,
and when they're done,
they create a pull request offering to merge their changes into the `master` branch
of the original repository.
In order to update their own repository,
they pull changes from the original:

FIXME: figure

A pull request can store more than the location of the source and destination branch.
In particular,
it can store comments that people have made about the proposed merge.
GitHub and other [forges](glossary.html#forge) allow users to add comments on the pull request as a whole,
or on particular lines,
and can mark old comments as out of date if the lines of code the comment is attached to are updated.


### Setting up contribution repositories

To see this all in action,
let's go ahead and work on the `countries` repository together.
Open the GitHub link for the `countries` repo provided by the instructor 
in your browser window.

*TODO: Figure - Upstream repo [../fig/github_screenshot_upstream_repo.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_upstream_repo.png)*

> **Repo owner differences**
>
> You may have noticed that the `countries` repo in this lesson's pictures 
> is owned by the 'McMahonLab' organization and this doesn't match
> the address you were given.
> This is to be expected because this will differ depending on 
> what organization your instructor used to setup the `countries` repo.
> 
> You will also see your username where the 'sstevens2' is in the pictures.
> 

Once at the `countries` repo, click the **Fork** button which can be found
in the upper right hand conner of the window. 
Forking the repository makes us each our own copy of the repo in our GitHub 
account which we can edit.

*TODO: Figure - Upstream Forking [../fig/github_screenshot_upstream_forking.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_upstream_forking.png)*

Next we need to get this repo on our local computer and
setup connections from our computer to both our forked version 
and the authoritative version we forked it from.

First we will **clone** the repo from our forked version.
The clone command does two things:

1. Copies the repo to your local computer
2. Sets up a remote called 'origin' between your computer and the GitHub repo

Copy the web address for your forked version of repo 
(from the web address line or click 'Clone and download' and copy that).

*TODO: Figure - Clone Origin [../fig/github_screenshot_cloneOrigin.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_cloneOrigin.png)*

At the command line, navigate to a folder you'd like to hold this repo,
we will place it on our `Desktop`.
Once there you can use the `clone` command with the link you copied as the first argument.

```shell
$ cd ~/Desktop
$ git clone https://github.com/USERNAME/countries.git
```

> **Why does the command above say 'USERNAME'?**
>
> So that we can't copy the command above and accidentally clone someone else's
> version of countries to our computer, the command above uses the placeholder
> 'USERNAME' where you should put your own username if your copied from above
> instead of copying the link from your browser and pasting it into the command.

```text
Cloning into 'countries'...
remote: Counting objects: 6, done.
remote: Compressing objects: 100% (4/4), done.
remote: Total 6 (delta 0), reused 6 (delta 0), pack-reused 0
Unpacking objects: 100% (6/6), done.
```

Next we will set up a connection or **remote** to the authoritative repository 
(the original GitHub repository given to you by your instructor).
In your browser, you can go back this repo by clicking on the link that says 'forked from'
in the upper left hand corner, under your username and repo name.

Copy the web address for this repo 
(from the web address line or click the 'clone and download' and copy that).

*TODO: Figure - Upstream Repo [../fig/github_screenshot_upstream_repo.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_upstream_repo.png)*

Then back in your terminal, navigate into the cloned repo and add the remote 
connection to this repository.
For this command we must give the remote a different nickname, 
where our original remote is 'origin'
this new remote will be called 'upstream'.
You could give it a different nickname but 'upstream' is a common nickname for
the authoritative repository.

```shell
$ cd countries
$ git remote add upstream https://github.com/INSTRUCTOR-GIVEN/countries.git
```

> **If you tried copying the command above...**
>
> You will have to replace 'INSTRUCTOR-GIVEN' with the site your instructor
> indicated at the beginning of this lesson. This will vary depending
> on how your instructor set up for this lesson.

At anytime you can see the remote connections your repo has using the following command:

```shell
$ git remote -v
```

```text
origin	https://github.com/USERNAME/countries.git (fetch)
origin	https://github.com/USERNAME/countries.git (push)
upstream	https://github.com/INSTRUCTOR-GIVEN/countries.git (fetch)
upstream	https://github.com/INSTRUCTOR-GIVEN/countries.git (push)
```


### Submitting a pull request

Now that we have this setup done we will be able to suggest 
changes to this repo using a pull request.
Each person will add a new file with info about a new country in it.

The instructor will now add a single file to the repository containing 
information about the the United States.

Next, we will update our local version of the repo to include the new file.
We use a command called `pull` to bring these changes to our local repository.
We must specify the remote and branch we want to pull from, in this case the 
`upstream` remote's `master` branch.

```shell
$ git pull upstream master
```

Now your local version of the repo is updated but our forked version of the 
repo is not yet up to date.
You can reload your fork in GitHub and see it does not contain the new 
`united_states.txt` file.
Now we need to update our forked version.
To do so we can `push` the changes in our local version to the master branch of our fork, 
called 'origin'.

```shell
$ git push origin master
```

Now let's each add a new country to the repository.
First let's make a new branch to work on.  This will keep our 'master' version
in sync with the authoritative version of the repository.
We can name our branch descriptively after the country we will be adding.
Mine will be `addFrance` since I'll be working with France.
Please pick a different country and shout it out (or add it to the etherpad) 
so no one else chooses the same one.
We will create the branch and switch into in one step 
as we learned earlier in the branching lesson.

```shell
$ git checkout -b addFrance
```

```text
Switched to a new branch 'addFrance'
```

Finally before we proceed to adding the new file, we will double 
check that we are on the right branch.

```text
$ git branch
```

```text
* addFrance
  master
```

Next we will copy `united_states.txt` and change the name to the name of our chosen country.
Then we can use nano to edit the contents to reflect the info of your chosen country.  
Hint: You may need to do some internet searching to fill in the information.

```shell
$ cp united_states.txt france.txt
$ nano france.txt
$ cat france.txt
```

```text
Population: 66,991,000
Capital: Paris
```

Next let's add and commit the changes to the repo.

```shell
$ git add france.txt
$ git commit -m "Added file on france"
```

```text
[addFrance 79a312a] Added file on france
 1 file changed, 2 insertions(+), 2 deletions(-)
```

In some cases we may not have permission to push changes directly to the 
upstream/authoritative repo or we might like our changes to be reviewed regardless 
of permissions, so we'll create a `pull request`. 
A `pull request` is a **request** for a member of the upstream repository to **pull** 
our changes into the upstream repository from a `fork`, allowing them to request further 
changes/improvements and make comments on the changes before doing so. 
In order to create a `pull request`, we must push our new branch containing the
 changes we'd like to submit to the remote linked to our fork, `origin`, on GitHub.

```shell
$ git push origin addFrance
```

```text
Counting objects: 4, done.
Delta compression using up to 4 threads.
Compressing objects: 100% (4/4), done.
Writing objects: 100% (4/4), 783 bytes | 0 bytes/s, done.
Total 4 (delta 3), reused 0 (delta 0)
remote: Resolving deltas: 100% (3/3), completed with 3 local objects.
To https://github.com/USERNAME/countries.git
   2037539..79a312a  addFrance -> addFrance
```

Next go to your forked GitHub version of the repo and reload the page.
You won't see the new file added in the list of files but you will see 
that you recently pushed a new branch to the repository.

*TODO: Figure - Add France [../fig/github_screenshot_origin_master_addFrance.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_origin_master_addFrance.png)*


If you wish to view your new branch you can click on the 'Branch' drop down menu
and select that branch.

*TODO: Figure - Switch Branch [../fig/github_screenshot_switch_github_branch.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_switch_github_branch.png)*

Then you should be able to view the files and commits in that branch.

*TODO: Figure - Add France [../fig/github_screenshot_origin_master_addFrance.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_origin_master_addFrance.png)*

GitHub already suspects that we are going to want to make a pull request so we can click
the 'Compare & pull request' button to start a new pull request.

*TODO: Figure - Pull Request (Part 1) [../fig/github_screenshot_makingPR1.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_makingPR1.png)*

The base fork should be the upstream/authoritative version's master branch and then 
the head fork should be the new branch of our fork.
You can add more information into the comment section if there is anything you'd like 
to add for the person who reviews your suggestion.
Then you can click the 'create pull request button' to submit the pull request.

*TODO: Figure - Pull Request (Part 2) [../fig/github_screenshot_makingPR2.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_makingPR2.png)*

Now someone with privileges to the upstream repo can review it, give comments and
suggestions, and merge it into the upstream version.
In our pull request they can see any messages we left or click and look at the commits that were made and see the files changed.

Our collaborator reviewing the pull request noticed that 
we forgot to add the largest city so let's add it and update our pull request.

```shell
$ nano france.txt
$ cat france.txt
```

```text
Population: 66,991,000
Capital: Paris
Largest City: Paris
```

Next we will add and commit these changes.
Then we can push them to our fork of the repo.

```shell
$ git add france.txt
$ git commit -m "Added largest city to france file"
$ git push origin addFrance
```

```text
Counting objects: 3, done.
Delta compression using up to 4 threads.
Compressing objects: 100% (3/3), done.
Writing objects: 100% (3/3), 387 bytes | 0 bytes/s, done.
Total 3 (delta 0), reused 0 (delta 0)
To https://github.com/USERNAME/countries.git
   31aa2e3..609acfe  addFrance -> addFrance
```

If we reload the pull request, we'll see that the new commit was added to the 
pull request and the changes have been automatically updated.
New commits pushed to the same branch are included in the previous pull request.
If you want to suggest changes separately you need to make separate branches but 
if you want the changes to be considered together you should put them in the same branch.

*TODO: Figure - New Commit [../fig/github_screenshot_after_new_commit.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_after_new_commit.png)*


### Dealing with Conflicting Changes

When working with others it's possible that two people will edit the same code,
creating a conflict when we try and merge changes.
Here we will look at how to resolve conflicts when working collaboratively.

We will continue to work in the `addFrance` branch from before and check we are
in that branch before we start.


```shell
$ git branch
```

```text
* addFrance
  master
```

Next we will each add our country to the existing `README.md` file in the repository in the line directly following the `Countries:` line.

```shell
$ nano README.md
$ cat README.md
```

```text
# countries
Sandbox for learning PR's in Software Carpentry workshop

Countries:
France
```

Next we need to add, commit, and push these requests to our existing pull request.

```shell
$ git add README.md
$ git commit -m "Added France to list of countries in README"
```

```text
[addFrance 66d7ebf] Added France to list of countries in README
 1 file changed, 1 insertions(+), 1 deletions(-)
```

```shell
$ git push origin addFrance
```

```text
Counting objects: 3, done.
Delta compression using up to 4 threads.
Compressing objects: 100% (3/3), done.
Writing objects: 100% (3/3), 376 bytes | 0 bytes/s, done.
Total 3 (delta 1), reused 0 (delta 0)
remote: Resolving deltas: 100% (1/1), completed with 1 local object.
To https://github.com/USERNAME/countries.git
   609acfe..66d7ebf  addFrance -> addFrance
```

Now if we reload the page we had a pull request we notice that our `addFrance`
branch is conflicting with upstream's `master` branch.
This is because someone else edited the same line of the `README.md` file by 
adding 'United States' where we added 'France'.

*TODO: Figure - Conflicting PR [../fig/github_screenshot_conflicting_PR.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_conflicting_PR.png)*

In this case, it is possible to resolve this conflict in GitHub by 
clicking the 'Resolve Conflicts' button.
However, we can also resolve this conflict locally.

First we need to pull down the changes from upstream's `master` branch into our
`addFrance` branch.

```shell
$ git pull upstream master
```

```text
From https://github.com/McMahonLab/countries
 * branch            master     -> FETCH_HEAD
Auto-merging README.md
CONFLICT (content): Merge conflict in README.md
Automatic merge failed; fix conflicts and then commit the result.
```

From the conflict error message we can see the conflict is in `README.md` or by running `git status` and seeing the 'both modified' status.

```shell
$ git status
```

```text
On branch addFrance
You have unmerged paths.
  (fix conflicts and run "git commit")

Unmerged paths:
  (use "git add <file>..." to mark resolution)

	both modified:   README.md

no changes added to commit (use "git add" and/or "git commit -a")
```

Now we will resolve the conflict by editing the `README.md` file to contain both 
'United States' and 'France' and none of the additional lines git added to
indicate conflict

```shell
$ nano README.md
$ cat README.md
```

```text
# countries
Sandbox for ComBEE GitHub workshop on PR's

Countries:
France
United States
```

Then we add and commit our resolved conflict.

```shell
$ git add README.md
$ git commit -m "Resolved conflict in readme w two countries"
```

```text
[addFrance 912317b] Resolved conflict in readme w two countries
```

Finally we can update the pull request by pushing these changes to our GitHub 
fork of the repository.

```shell
$ git push origin addFrance
```

```text
git push origin addFrance
Counting objects: 6, done.
Delta compression using up to 4 threads.
Compressing objects: 100% (6/6), done.
Writing objects: 100% (6/6), 732 bytes | 0 bytes/s, done.
Total 6 (delta 2), reused 0 (delta 0)
remote: Resolving deltas: 100% (2/2), completed with 1 local object.
To https://github.com/USERNAME/countries.git
   66d7ebf..912317b  addFrance -> addFrance
```

Now if we reload our browser we will see that the new commit is in the pull 
request and it has no conflicts with the base branch.

*TODO: Figure - Resolved Conflict [../fig/github_screenshot_resolved_PR_conflict.png](https://github.com/UW-Madison-DataScience/git-novice-custom/tree/gh-pages/fig/github_screenshot_resolved_PR_conflict.png)*

Now the owner/administrator/manager of the authoritative repo can review our pull requests and decide to incorporate them.



## Exercises {#rse-advanced-git-exercises}


### Branches

TODO.


### Pull requests

**Add new country file and make additional PR**

* Starting in the master branch make a new branch
* Copy other country file into a new country
* Edit the file to include info on the new country
* Add and commit this new file
* Push the new changes to GitHub

> *Solution*
> 
> ```shell
> $ git checkout master
> $ git checkout -b addItaly
> $ cp united_states.txt italy.txt
> $ nano italy.txt #Add the right info into the file
> $ git add italy.txt
> $ git commit -m "Added file on Italy"
> $ git push origin addItaly
> ```


## Key Points {#rse-advanced-git-keypoints}

```{r, child="keypoints/rse-advanced-git.md"}
```

```{r, child="./links.md"}
```


