# RSE Learning Objectives {#objectives}

This appendix lays out the learning objectives for each set of lessons,
and is intended to help instructors who want to use this curriculum.

## The Unix Shell
```{r, child="objectives/bash.md"}
```

## Automating Analyses
```{r, child="objectives/rse-automate.md"}
```

## Configuring Software
```{r, child="objectives/configure.md"}
```

## Unit Testing
```{r, child="objectives/unit.md"}
```

## Verification
```{r, child="objectives/verify.md"}
```

## A Branching Workflow
```{r, child="objectives/branches.md"}
```

## Managing Backlog
```{r, child="objectives/backlog.md"}
```

## Code Style and Review
```{r, child="objectives/style.md"}
```

## Development Process
```{r, child="objectives/process.md"}
```

## Continuous Integration
```{r, child="objectives/integrate.md"}
```

## Working Remotely
```{r, child="objectives/remote.md"}
```

## Other Tools
```{r, child="objectives/tools.md"}
```

## Documentation
```{r, child="objectives/docs.md"}
```

## Refactoring
```{r, child="objectives/refactor.md"}
```

## Project Structure
```{r, child="objectives/project.md"}
```

## Including Everyone
```{r, child="objectives/inclusive.md"}
```

## R Packaging
```{r, child="objectives/rse-package-r.md"}
```

## Python Packaging
```{r, child="objectives/rse-package-py.md"}
```

## Publishing
```{r, child="objectives/publish.md"}
```

## Teamwork
```{r, child="objectives/teamwork.md"}
```

## Pacing
```{r, child="objectives/pacing.md"}
```

## Finale
```{r, child="objectives/finale.md"}
```
