# The Unix Shell {#shell}

## Questions {#shell-questions}

```{r, child="questions/shell.md"}
```

## Objectives {#shell-objectives}

```{r, child="objectives/shell.md"}
```

## Introduction {#shell-intro}

**Look at [TLDR pages](https://tldr.sh/) (which has a command-line client `tldr`) for content.**

-   What is the difference between an operating system and a shell?
-   Why would I use the shell?

## Working With Files {#shell-files}

-   How do we see what's in a folder?
-   How do we move around in the shell?
-   How do we view the contents of a text file?
-   How do we copy a file?
-   How do we move a file?
-   How do we rename a file?
-   How do we delete a file?

## Editing Text Files {#shell-edit}

-   What's the difference between a text file and a binary file?
-   How do we run a text editor?
-   How do we edit text in a text editor?
-   How do we move around in a text editor?
-   How do we find and replace in a text editor?

## Operating on Text {#shell-text}

-   How do we find out how long a file is?
-   How do we select the top or bottom of a file?
-   How do we save the output of a command in a file?
-   How do we select columns of text?

## Shortcuts {#shell-shortcut}

-   How do we see what I've done?
-   How do we use tab completion to fill in filenames?
-   How do we repeat previous commands?

## Working with Multiple Files {#shell-glob}

-   How can we select many files at once ([globbing](glossary.html#globbing))?
-   How can we make selections more specific?

## Paths {#shell-path}

-   What is an absolute path?
-   What is a relative path?
-   What special paths does the shell understand?

## Pipes {#shell-pipe}

-   How do we combine two or more commands?

## Variables {#shell-var}

-   What is a variable?
-   How do we create variables?
-   How do we use variables?
-   What variables does the shell automatically create for us?

## Writing Shell Scripts {#shell-script}

-   How do we execute commands saved in a file?
-   How do we pass filenames into a script?

## Repeating Things in the Shell {#shell-loop}

-   What is a loop?
-   How do we write a loop in a shell script?
-   How do we write a loop interactively?

## Key Points {#shell-keypoints}

```{r, child="keypoints/shell.md"}
```

```{r, child="etc/links.md"}
```
