# Teamwork {#teamwork}

## Questions {#teamwork-questions}

-   FIXME

## Objectives {#teamwork-objectives}

```{r, child="objectives/teamwork.md"}
```

## Introduction {#teamwork-intro}

FIXME: introduction

## How can we run meetings more efficiently? {#teamwork-efficient}

Most people do meetings poorly:
they don't have an agenda going in,
they don't take minutes,
they waffle on or wander off into irrelevancies,
they repeat what others have said or recite banalities simply so that they'll have said something,
and they hold side conversations
(which pretty much guarantees that the meeting will be a waste of time).
Knowing how to run a meeting efficiently is a core skill for anyone who wants to get things done.
Knowing how to take part in someone else's meeting is just as important,
but gets far less attention:
everyone offers leadership training,
nobody offers followership training.

The rules for running meetings quickly and smoothly are well known but rarely followed:

Decide if there actually needs to be a meeting.
:   If the only purpose is to share information,
    have everyone send a brief email instead.
    Remember, people can read faster than they can speak:
    if someone has facts for the rest of the team to absorb,
    the most polite way to communicate them is to type them in.

Write an agenda.
:   If nobody cares enough about the meeting to prepare a point-form list of what's to be discussed,
    the meeting itself probably doesn't need to happen.

Include timings in the agenda.
:   Timings help you prevent early items stealing time from later ones.
    Your first estimates with any new group will be wildly optimistic,
    so revise them upward for subsequent meetings.
    However,
    you shouldn't plan a second or third meeting just because the first one ran over-time:
    instead, try to figure out why you're running over and fix the underlying problem.

Prioritize.
:   Tackle issues that will have high impact but take little time first,
    and things that will take more time but have less impact later.
    That way, if the first things run over time,
    the meeting will still have accomplished something.

Make one person responsible for keeping things moving.
:   One person should be made chair,
    and be responsible for keeping items to time,
    chiding people who are having side conversations or checking email,
    and asking people who are talking too much to get to the point.
    The chair should *not* do all the talking;
    in fact,
    whoever is in charge will talk less in a well-run meeting than most other participants.

Require politeness.
:   No one gets to be rude,
    no one gets to ramble,
    and if someone goes off topic,
    it's the chair's job to say,
    "Let's discuss that elsewhere."

No interruptions.
:   Participants should raise a finger,
    put up a sticky note,
    or make one of the other gestures people use at high-priced auctions
    when they want to speak.
    The chair should keep track of who wants to speak and give them time in turn.

No side conversations.
:   This makes meetings more efficient because
    Nobody can actually pay attention to two things at once.
    If distractions are tolerated,
    people will miss things or they'll have to be repeated.
    More importantly,
    not paying attention is insulting---chatting with a friend
    while someone explains what they did last week is a really effective way to say,
    "I don't think you or your work is important."

No technology.
:   Insist that everyone put their phones, tablets, and laptops into politeness mode
    (i.e., close them).
    If this is too stressful,
    let participants hang on to their electronic pacifiers but turn off the network
    so that they really *are* using them just to take notes or check the agenda.
    The one exception is accessibility needs:
    if someone needs their phone, their laptop, or some other aid
    in order to take part in the meeting,
    nobody has a right to tell them "no".

Take minutes.
:   Someone other than the chair should take point-form notes
    about the most important information that was shared,
    and about every decision that was made or every task that was assigned to someone.

Take notes.
:   While other people are talking,
    participants should take notes of questions they want to ask
    or points they want to make.
    (You'll be surprised how smart it makes you look when it's your turn to speak.)

End early.
:   If your meeting is scheduled for 10:00-11:00,
    aim to end at 10:55 to give people time to get where they need to go next.

## What should we do when the meeting is over? {#teamwork-after}

As soon as the meeting is over, circulate the minutes
(i.e., emailed them to everyone or post them to a wiki):

People who weren't at the meeting can keep track of what's going on.
:   You and your teammates all have to juggle tasks from other projects or courses,
    which means that sometimes you won't be able to make it to meetings.
    A wiki page,
    email message,
    or blog entry is a much more efficient way to catch up than asking a team mate,
    "Hey, what did I miss?"

Everyone can check what was actually said or promised.
:   More than once,
    one of us has looked over the minutes of a meeting and thought,
    "Did I say that?" or,
    "Wait a minute, I didn't promise to have it ready then!"
    Accidentally or not,
    people will often remember things differently;
    writing it down gives the team a chance to correct mis-recollection,
    mis-interpretation,
    or mis-representation,
    which can save a lot of anguish later on.

People can be held accountable at subsequent meetings.
:   There's no point making lists of questions and action items
    if you don't follow up on them later.
    If you're using a ticketing system (Chapter \@ref(backlog)),
    create a ticket for each new question or task right after the meeting
    and update those that are being carried forward.
    That way,
    your agenda for the next meeting can start by rattling through
    the list of open tickets.

Run all your meetings like this for a month,
with the goal of making each one a minute shorter than the one before,
and we promise that you'll build better software.

## How can we keep people from talking too much or too little? {#teamwork-fair}

Some people are so used to the sound of their own voice that
they will talk half the time no matter how many other people are in the room.
One way to combat this is to give everyone three sticky notes at the start of the meeting.
Every time they speak,
they have to give up one sticky note.
When they're out of stickies,
they aren't allowed to speak until everyone has used at least one,
at which point everyone gets all of their sticky notes back.
This ensures that nobody talks more than three times as often as
the quietest person in the meeting,
which completely changes group dynamics:
people who have given up trying to be heard because they always get trampled
suddenly have space to contribute,
and the overly-frequent speakers quickly realize just how unfair they have been.

Another useful technique is called interruption bingo.
Draw a grid and label the rows and columns with the participants' names.
Each time one person interrupts another,
add a tally mark to the appropriate cell;
halfway through the meeting,
take a moment to look at the results.
In most cases,
you will see that one or two people are doing all of the interrupting,
often without being aware of it.
After that, saying, "All right, I'm adding another tally to the bingo card,"
is often enough to get them to throttle back.
(Note that this technique is for managing interruptions,
not speaking time.
It may be completely appropriate for people with more knowledge of a subject
to speak about it more often in a meeting,
but it is never appropriate to repeatedly cut people off.)

## How should we run online meetings? {#teamwork-online}

Chelsea Troy's discussion of
[why online meetings are often frustrating and unproductive][troy-meetings]
makes an important point:
in most online meetings,
the first person to speak during a pause gets the floor.
As a result,
"If you have something you want to say,
you have to stop listening to the person currently speaking
and instead focus on when they're gonna pause or finish
so you can leap into that nanosecond of silence and be the first to utter something.
The format...encourages participants who want to contribute
to say more and listen less."

The solution is to run a text chat beside the video conference
where people can signal that they want to speak.
The chair can then select people from the waiting list.
This practice can be reinforced by having everyone mute themselves,
and only allowing the moderator to unmute people.

## What sort of people make teamwork hard? {#teamwork-people}

FIXME: introduction (without Tolstoy)

Anna
:   knows more about every subject than everyone else on the team put together---at least,
    she thinks she does.
    No matter what you say, she'll correct you;
    no matter what you know, she knows better.
    Anna is pretty easy to spot:
    if you keep track of how often people interrupt one another in team meetings,
    her score is usually higher than everyone else's put together.

Bao
:   is a contrarian:
    no matter what anyone says, he'll take the opposite side.
    This can be healthy in small doses,
    but if someone plays devil's advocate at every turn,
    they're just impeding progress.

Caitlin
:   has so little confidence in her own ability
    (despite her good grades)
    that she won't make any decision,
    no matter how small,
    until she has checked with someone else.
    Everything has to be spelled out in detail for her
    so that there's no possibility of her getting anything wrong.

Frank
:   believes that knowledge is power.
    He enjoys knowing things that other people don't---or to be more accurate,
    he enjoys it when people know he knows things they don't.
    Frank is good at making things work,
    but when asked how he did it,
    he'll grin and say, "Oh, I'm sure you can figure it out."

Gary
:   is a hitchhiker.
    He has discovered that most people would rather shoulder some extra work than snitch,
    and he takes advantage of it at every turn.
    The frustrating thing is that he's so damn *plausible*
    when someone finally does confront him.
    "There have been mistakes on all sides," he says,
    or, "Well, I think you're nit-picking."
    The only way to deal with Kenny is to stand up to him:
    remember, if he's not doing his share,
    *he's the bad guy*, not you.

Hediyeh
:   is quiet---very quiet.
    She never speaks up in meetings,
    even when she knows that what other people are saying is wrong.
    She might contribute to the mailing list,
    but she's very sensitive to criticism,
    and will always back down rather than defending her point of view.
    Hediyeh isn't a troublemaker,
    but rather a lost opportunity.

Melissa
:   would easily have made the varsity procrastination team
    if she'd bothered to show up to tryouts.
    She means well,
    and she really does feel bad about letting people down,
    but somehow something always comes up,
    and her tasks are never finished until the last possible moment.
    Of course,
    that means that everyone who is depending on her
    can't do their work until *after* the last possible moment...

Petra
:   has a favorite phrase: "why don't we".
    Why don't we write a GUI to help people edit the program's configuration files?
    Hey, why don't we invent our own little language for designing GUIs?
    Her energy and enthusiasm are hard to argue with,
    but argue you must.
    Otherwise,
    for every step you move forward,
    the project's goalposts will recede by two.
    This is called [feature creep](glossary.html#feature-creep),
    and has ruined many projects that might otherwise have delivered something small but useful.

Raj
:   is rude.
    "That's stupid," and a mixed bag of obscenities are his favorite phrases.
    "It's just the way I talk," he says,
    not knowing or not caring that for a lot of people,
    that kind of language has often been a prelude to bullying or worse.
    His only redeeming grace is that he can't dissemble as well as Gary,
    so he's easier to get rid of.

Sergei
:   is simply incompetent.
    He doesn't understand the problem,
    he hasn't bothered to master the tools and libraries he's supposed to be using,
    the code he checks in doesn't run,
    and his thirty-second bug fixes introduce more problems than they solve.
    If he means well,
    try to re-partition the work so that he'll do less damage.
    If he doesn't,
    he should be treated like any other hitchhiker.

## How should we handle conflict within the team? {#teamwork-conflict}

FIXME: this is not about harassment or abuse---see Chapter \@ref(inclusive) for that.

You just missed an important deadline,
and people are unhappy.
The sick feeling in the pit of your stomach has turned to anger:
you did *your* part,
but Marta didn't finish her stuff until the very last minute,
which meant that no one else had time to spot the two big mistakes she'd made.
As for Chul, well, he didn't deliver at all---again.
If something doesn't change,
this project is going to pull down your performance review so far
that you might have to start looking for a new job.

Situations like this come up all the time.
Broadly speaking, there are four ways you can deal with them:

1.  Cross your fingers and hope that things will get better on their own,
    even though the last three times you hoped they would,
    they didn't.

2.  Do extra to make up for others' shortcomings.
    This sometimes works in the short term,
    and saves you the mental anguish of confronting others,
    but the time for that "extra" has to come from somewhere;
    what usually ends up happening is that other parts of the project,
    or your personal life,
    suffer.

3.  Lose your temper and start shouting.
    Unfortunately,
    people often wind up displacing their anger into other parts of their life:
    I've seen developers yell at waitresses for bringing incorrect change
    when what they really needed to do was tell their boss,
    "No, I *won't* work through another holiday weekend
    to make up for your decision to short-staff the project."

4.  Take constructive steps to fix the underlying problem.

Most of us find number four hard because we don't like confrontation.
If you manage it properly,
though,
it is a lot less bruising,
which means that you don't have to be as afraid of initiating it.
Also,
if people believe that you will actually take steps
when they bully, lie, procrastinate, or do a half-assed job,
they will usually avoid making it necessary.
(A colleague once said that
she had never met anyone who wished they had waited longer
before putting their foot down.)

Here are the steps you should take when you feel that
a teammate isn't pulling their weight:

Make sure you're not guilty of the same sin.
:   You won't get very far complaining about someone else interrupting in meetings
    if you do it just as frequently.

Check expectations.
:   Are you sure the offender knows what standards they are supposed to be meeting?
    This is where things like job descriptions
    or up-front discussion of who's responsible for what
    come in handy.

Check the situation.
:   Is someone dealing with an ailing parent or immigration woes?
    Have they been put to work on three other projects that you don't know about?

Document the offense.
:   Write down what the offender has actually done and why it's not good enough.
    Doing this will help you clarify matters in your own mind.
    It's also absolutely necessary if you have to escalate.

Check with other team members.
:   Are you alone in feeling that the offender is letting the team down?
    If so, you aren't necessarily wrong,
    but it'll be a lot easier to fix things if you have the support of the rest of the team.
    Finding out who else on the team is unhappy can be the   hardest part of the whole process,
    since you can't even ask the question without letting on that you're upset,
    and word will almost certainly get back to whoever you're asking about,
    who might then turn around and accuse you of stirring up trouble.
    After a couple of unhappy experiences of this kind,
    I've learned that it's best to raise the issue for the first time in a group of three:
    you, the person you want to talk to,
    and a third person who can act as unofficial moderator.

Talk with the offender.
:   This should be a team effort:
    put it on the agenda at a team meeting,
    present your complaint,
    and make sure that the offender understands it.
    In most cases this is enough:
    human beings are herd animals,
    and if someone realizes that they're going to be called on their hitchhiking or bad manners,
    they will usually change their ways.

Escalate as soon as there's a second offense.
:   Hitchhikers and others who really don't have good intentions
    are counting on you giving them one last chance after another
    until the project is finished and they can go suck the life force out of their next victim.
    *Don't fall into this trap.*
    If someone stole your laptop, you'd report it right away.
    If someone steals your time,
    you're being pretty generous giving them more than one chance to mend their ways.

In the context of a research project,
"escalation" means "taking the issue to your supervisor".
(If you're reluctant to do this because you don't want to be a snitch,
go back and re-read the bit about people stealing your laptop.)
Of course,
your supervisor has probably had dozens of students complain to her over the years
about teammates not doing their share---it isn't uncommon to have both halves of a pair
tell the instructor that they're doing all the work.
(This is yet another reason to use version control:
it makes it easy to check who's actually written what.)
In order to get her to take you seriously and help you fix your problem,
you should send her an email, signed by several people describing the problem
and the steps you have already taken to resolve it.
Make sure the offender gets a copy as well,
and ask your instructor to arrange a meeting to resolve the issue.

This is where documentation is crucial.
Hitchhikers are usually very good at appearing reasonable;
they're very likely to nod as you present your case,
then say, "Well, yes, but..." and rhyme off a bunch of minor exceptions
or cases where others on the team have also fallen short of expectations.
If you can't back up your complaint,
your supervisor will likely be left with the impression that the whole team is dysfunctional,
and nothing will improve.

One technique your supervisor may ask you to use in a meeting like this
is [active listening](glossary.html#active-listening).
As soon as one person makes a point,
the person on the opposite side of the issue explains it back to them,
as in, "So what I think Igor is saying is..."
This guarantees that the second person has actually paid attention to what the first person said.
It can also defuse a lot of tension,
since explaining someone's position back to them
forces you to see the world through their eyes,
if only for a few moments.

## Summary {#teamwork-summary}

-   @Brow2007 has lots of good advice on running meetings.
-   @Broo2016 is a catalog of ways to get people talking productively.

## Key Points {#teamwork-keypoints}

```{r, child="keypoints/teamwork.md"}
```

