# Other Tools {#tools}

## Questions {#tools-questions}

-   What other tools can help me be more productive?

## Objectives {#tools-objectives}

```{r, child="objectives/tools.md"}
```

## Introduction {#tools-intro}

The previous chapters have explored several key tools in depth,
but there are many others that will improve your productivity.
You shouldn't try to pick them all up at once,
but each will help you do something in less time and with less pain.

## What should I look for in an editor? {#tools-editor}

The next most important tool after a version control system is your editor.
There are literally thousands to choose from;
if you want a plain text editor, your choices range from the very small
(such as Pico, which is included in most Linux installations)
to the very large
(like Emacs,
whose name doesn't actually stand for "eighty megabytes and constantly swapping"
and which isn't actually a Lisp-based operating system in disguise).
There are also [WYSIWYG](glossary.html#wysiwyg) tools like Microsoft Word, LibreOffice, and Google Docs,
though these usually can't be used for programming,
since they insert non-ASCII characters and formatting information in files
(even files that look unformatted).

You probably already have a favorite editor.
If you're like most programmers, you will change jobs, languages, operating systems, and nationality
before you'll switch to another,
because it's taken weeks or months for your hands to master the current one.
However, if your editor doesn't pass the following tests,
switching now will save you enough time and grief in the future
to make the temporary loss of productivity worthwhile.

FIXME: keep this personal?

Is it kind to your wrists?
:   If ergonomics was a standard part of the undergraduate curriculum,
    programming editors would probably be used as examples of bad design.
    I was an Emacs power user for many years,
    and have paid a heavy price for it:
    I can't type for more than half an hour now before my hands start to hurt
    from all those control-key combinations and long reaches for the escape key.

Is it portable?
:   I'm writing this on Windows;
    an hour from now I'll be using my Mac laptop,
    and on Monday I'll be back on a Linux machine at work.
    I don't know what I'll be using next year,
    but what I *do* know is that I won't want to have to retrain my hands to use it.
    You should therefore pick an editor that's portable across all three major operating systems.

Is it syntax-aware?
:   Human beings shouldn't have to indent code
    or count parentheses to make sure they have closed all the ones they opened.
    They equally shouldn't have to type every character in the name of every function or method they call:
    their editor should do these things for them.
    This won't just save wear and tear on your wrists:
    letting the machine do it will make the text you create more uniform,
    and hence more readable (Chapter \@ref(style)).

Is it programmable?
:   Just like spreadsheets, games, and other applications,
    editors are built out of functions.
    Good ones let you call those functions yourself so that you can write small programs
    (usually called [macros](glossary.html#macro)) to automate common operations.
    As a programmer,
    you have the skills to make your tools do more than their creators intended,
    so you should find tools that allow you to.

Does it handle Unicode characters?
:   Most programmers don't speak English as a first language,
    and spell their names using symbols that aren't in the old 7-bit ASCII character set.
    Programming languages haven't caught up yet---they still insist on using `<=`, for example---but at the very least,
    you have to be able to edit data and documentation that include Greek, Tamil, and the like.
    (See [Spolsky2003] for a developer-oriented introduction to character sets in general and Unicode in particular.)

One thing that *isn't* on the list above is [syntax highlighting](glossary.html#syntax-highlighting).
Most programming editors do it by default,
but the most recent empirical studies show that it has little or no benefit @Hann2018.
On the other hand,
many programmers claim to find it useful,
and there's no evidence it does any harm,
so we have turned it on for the code samples in these lessons.

## What are symbolic debuggers and how can they help me? {#tools-debugger}

A [symbolic debugger](glossary.html#symbolic-debugger) is a program
that allows you to control and inspect the execution of other programs.
You can step through the target program a line at a time,
display the values of variables or expressions,
look at the call stack,
or (my personal favorite) set [breakpoints](glossary.html#breakpoint) to say,
"Pause the program when it reaches this line".
Depending on the language you're using,
you may have to compile your program with certain options turned on to make it debuggable,
but that's a small price to pay for the hours or days a debugger can save you
when you're trying to track down a problem.

Some debuggers,
like [GDB][gdb],
are standalone programs,
while others are build into integrated development environments (discussed below).
Both are better than adding `print` statements to your program and re-running it because:

-   adding `print` statements takes longer than clicking on a line
    and setting a breakpoint;

-   adding `print` statements distorts the code you're debugging by moving things around in memory,
    altering the flow of control,
    and/or changing the timing of thread execution;
    and

-   it's all too easy to make a mistake in the `print` statement---few things are as frustrating
    as wasting an afternoon debugging a problem,
    only to realize that the `print` statement you copied and pasted
    isn't displaying the values you thought it was.

Debuggers and logging (Chapter \@ref(logging)) are complementary.
A debugger is what you use when you are trying to track down a specific problem;
logging is what you put into your code so that *if* there's a problem
you'll have the information you need to track it down more quickly.

Over the years,
I've been surprised by how strongly some programmers resist using a debugger.
The reason can't be the five or ten minutes it takes to learn how to use one---that pays for itself
almost immediately.
The only explanation I've been able to come up with is that some people *enjoy* being inefficient.
Typing in `print` statements and paging through screens of output lets them feel like they're being productive,
when in fact they're just being busy.
If your brain needs a break (which it sometimes will),
then take a break:
stretch your legs,
stare out a window,
practice your juggling,
or do whatever else you can to take your mind away from your problem for a few minutes.
Don't drag out the process of finding and fixing your bug by using sloppy technique
just to let your brain idle for a while.

## How can I tell why my program is running slowly? {#tools-profiler}

Section \@ref(unit-coverage) described how
a coverage tool can keep track of which lines of a program have and have not been executed
so that programmers can figure out which bits of code haven't been tested.
A [profiler](glossary.html#profiler) is a more sophisticated tool that records how much time was spent executing each line
so that programmers can figure out which parts are slowest.
As a very rough guide,
20% of the code accounts for 80% of execution time,
so finding and [tuning](glossary.html#tuning) a few [hot spots](glossary.html#hot-spot) is the fastest way to make programs faster.

There are two ways to do profiling.
[instrumenting profilers](glossary.html#instrumenting-profiler) insert extra statements into the application
to record the clock time,
while [sampling profilers](glossary.html#sampling-profiler) interrupt execution every few clock ticks
and see where the program is.
Both approaches distort the program's behavior slightly,
so there is an unavoidable tradeoff between how much information is collected
and how accurately that information represents the behavior of the unmodified program.

FIXME: figure showing two kinds of profilers

Profilers typically present two kinds of information:
how much time was spent in a statement,
and how much was spent in that statement and any functions it called.
They may also present time per statement and total time,
so that if a line of code executes quickly,
but is called millions of times,
it will show up as a hot spot.

Once you have found where your program is slow,
the next step is to speed it up.
That's beyond the scope of this tutorial,
but [the "Performance" section][advanced-r-performance] of @Wick2014b
or @Gore2014,@Lana2017 are good places to start.

## How can I get all of these tools in one go? {#tools-ide}

A build system,
a smart editor,
and a debugger all talking to one another:
that's a pretty good description of an [integrated development environment](glossary.html#ide) or IDE.
These were invented in the 1970s,
but didn't really catch on until Borland released Turbo Pascal in the 1980s.
Along with the tools described above,
modern IDEs usually include:

-   a [code browser](glossary.html#code-browser)
    that displays an overview of the packages, classes, methods, and data in your program;

-   an interactive prompt so that you can type in expressions or call functions
    and see the results without having to start (or restart) your program;

-   a [linter](glossary.html#linter) that can warn you when your code doesn't meet naming and indentation conventions
    (discussed in Chapter \@ref(style));

-   some refactoring tools to help you reorganize your code
    (discussed in Chapter \@ref(refactor));
    and

-   a [test runner](glossary.html#test-runner) to display the results of tests
    and let you jump directly to ones that have failed
    (Chapter \@ref(unit)).

In short,
an IDE is to programming what a well-equipped workbench is to a carpenter.
Microsoft's [Visual Studio Code][vs-code],
Apple's [XCode][xcode],
and [IntelliJ IDEA][intellij-idea] from JetBrains are three popular commercial IDEs;
[WingIDE][wing-ide] is a good place to start for Python programmers,
and [RStudio][rstudio-ide] is a full-featured IDE for R
that can also be used [online][rstudio-cloud].

There are dozens of others,
any of which will make you more productive than their disconnected counterparts.
Since most of these store project data (including build instructions) in a proprietary format,
your team will do much better if you all adopt the same IDE.
This will also let you help one another solve problems and share plugins.
(Having to agree on which IDE to use may be another reason why some programmers resist adopting any IDE at all,
since they require even more investment to master than editors.)

## How can I run tasks at regular times? {#tools-cron}

One other kind of tool that many programmers rely on is one that runs tasks at pre-defined times.
The best known of these is the venerable Unix utility [`cron`][cron],
and a related tool is [`at`][at],
which runs a single job at some time in the future.
`cron`'s configuration files are famously cryptic,
and it's more common these days to run jobs when something changes in a version control repository
(Chapter \@ref(integrate))
than at a specific time,
but you may find yourself in need.

## Summary {#tools-summary}

FIXME: summary of tools

## Exercises {#tools-exercises}

FIXME: exercises for tools

## Key Points {#tools-keypoints}

```{r, child="keypoints/tools.md"}
```

```{r, child="etc/links.md"}
```
