# Key Points {#keypoints}

<!-- Used `bin/list-keypoints-objectives.R` to create this list -->

## Getting Started

```{r, child='keypoints/getting-started.md'}
```

## The Basics of the Unix Shell

```{r, child='keypoints/bash-basics.md'}
```

## Building Tools with the Unix Shell

```{r, child='keypoints/bash-tools.md'}
```

## Going Further with the Unix Shell

```{r, child='keypoints/bash-advanced.md'}
```

## Building Command-Line Programs in Python

```{r, child='keypoints/scripting.md'}
```

## Using Git at the Command Line

```{r, child='keypoints/git-cmdline.md'}
```

## Going Further with Git

```{r, child='keypoints/git-advanced.md'}
```

## Working in Teams

```{r, child='keypoints/teams.md'}
```

## Automating Analyses with Make

```{r, child='keypoints/automate.md'}
```

## Configuring Programs

```{r, child='keypoints/config.md'}
```

## Handling Errors

```{r, child='keypoints/errors.md'}
```

## Testing Software

```{r, child='keypoints/testing.md'}
```

## Tracking Provenance

```{r, child='keypoints/provenance.md'}
```

## Creating Packages with Python

```{r, child='keypoints/packaging.md'}
```
