# Learning Objectives {#objectives}

This appendix lays out the learning objectives for each set of lessons,
and is intended to help instructors who want to use this curriculum.

<!-- Used `bin/list-keypoints-objectives.R` to create this list -->

## Getting Started

```{r, child='objectives/getting-started.md'}
```

## The Basics of the Unix Shell

```{r, child='objectives/bash-basics.md'}
```

## Building Tools with the Unix Shell

```{r, child='objectives/bash-tools.md'}
```

## Going Further with the Unix Shell

```{r, child='objectives/bash-advanced.md'}
```

## Building Command-Line Tools with Python

```{r, child='objectives/scripting.md'}
```

## Using Git at the Command Line

```{r, child='objectives/git-cmdline.md'}
```

## Going Further with Git

```{r, child='objectives/git-advanced.md'}
```

## Working in Teams

```{r, child='objectives/teams.md'}
```

## Automating Analyses with Make

```{r, child='objectives/automate.md'}
```

## Configuring Programs

```{r, child='objectives/config.md'}
```

## Testing Software

```{r, child='objectives/testing.md'}
```

## Handling Errors

```{r, child='objectives/errors.md'}
```

## Tracking Provenance

```{r, child='objectives/provenance.md'}
```

## Creating Packages with Python

```{r, child='objectives/packaging.md'}
```
