# Getting Started with R {#r-getting-started}

```{r r-getting-started-setup, include=FALSE}
source(here::here("_common.R"))
```

## Questions {#r-getting-started-questions}

```{r, child="questions/r-getting-started.md"}
```

As stated in the introduction, our overall goal is to work with people, programs
and data. In this section, we will focus on programs and data as we learn
how to run R code, as well as
how data is stored and accessed on a computer.

## Introduction to RStudio {#r-getting-started-rstudio}

Throughout this book, we'll be writing
programs (or, in verb form, *programming*) in order to accomplish our goals
of working with data on the computer.
Programming is one way to make a computer do something for us.
Instead of clicking, we'll mostly be typing; instead of doing
what someone else has pre-defined, we'll have a lot of flexibility to
do what we want.

Just like using a web browser to access websites, and a program like
MS Word to write documents, it's helpful to have a program on your computer
that is designed to make it easy to write and run code. This kind of program
is called an "IDE" or Integrated Development Environment.

The one we'll be using in this book for writing R code is called RStudio.
RStudio (like many IDEs) has many panes (or panels or boxes),
each of which has a different purpose.

FIXME: Screenshot or schematic.

We'll take it slowly and introduce the purpose of each pane one at a time.
We'll start with the pane occupying the lefthand side of your screen.

## Running Code in the Console {#r-getting-started-console}

Now that we have our space for writing and running code (our **environment**)
open, it's time to
actually run some code.

The first time you open a new installation of RStudio,
the pane occupying the entire lefthand side of the screen is called the console.
The console is a program that is constantly ready and waiting to accept and run
code. The console you see in RStudio is expecting to see R code.
Here's an example you can type or copy in to see how the console works:

```{r}
print("It was the best of times, it was the worst of times.")
```

A function is a set of code that can repeatedly perform a specific task.
As you might guess from the name, the R `print()` function
takes in text (indicated by the quotes) and then prints the text back out to
the console.
R has many built-in functions, each of which accomplishes different things:

```{r}
round(3.1415)
Sys.Date()
length("hippo")
```

Part of learning to program is learning some of the base R functions and
what they do; in upcoming chapters, we'll focus on functions that allow you to
read in and manipulate data. At some point it may be helpful to write your
*own* functions, which we'll also cover in another chapter.

## Running Code via Scripts {#r-getting-started-script}

Can you imagine a situation where continuing to type code into the console
could become tedious or challenging?

Here are some examples:

* Running the same code many times (you
*can* use the up-arrow to see previous commands - try it! - but if you
run many commands, you might end up scrolling a long time.
* Organizing long sections of code in a meaningful way, and working on
  subsections separately.
* Saving the code as a text file that can be stored on the computer's
  hard drive and shared with collaborators.

Clearly, we need an approach that will allow us to write and run code
while keeping a record of our work and allowing us to run (and re-run) parts.

The answer to these challenges is to
add our R code to a text file called a script.
You can create a script in RStudio by using the "File"
menu option to select "New file" and then "R Script."
Your R script (a blank text file) should appear on the left side of the screen
above the console.
Try copying in some
of the commands we've already run:

```{r}
round(3.1415)
Sys.Date()
```

FIXME: Screenshot or animated gif.

Now these commands are written in the script, but nothing has happened yet.
In order to run the commands from the script (just like we did in the console),
we can choose from a variety of run commands. RStudio includes a "Run" button
in the
script pane. Clicking this button runs the current line of code.
You should see both the code and output appear in the console pane on the
bottom left. Your cursor will also move to the next line of code.
Alternatively, you can use a keyboard shortcut to run code.
Hovering your mouse over the run button will show you the shortcut for
running code on your computer; this shortcut is typically `Ctrl+Enter`.
You can also try selecting multiple lines of code with your cursor to execute
multiple lines simultaneously.

Sometimes, when you want to experiment or check something, it can make sense
to write and run R code in the console. However, most of the time, you'll
want to write and run your code from a script.  Using
scripts has the benefit of saving your work while also being able to run
the code just like in the console. You and your collaborators can then also
easily share and run the code outside of the IDE.

We have now told the computer what to do by using R code, and we have run that
code in two different ways within our "workbench" - the RSudio IDE. Let's
see what else we can use in this environment to help us.

## Assigning and Recalling Objects {#r-getting-started-objects}

Besides running functions that *do* something (as above), we'll also want
to use R to keep track of information that we're using throughout
our analysis. We save that information by creating an
[object][object])
using a name we select, special set of symbols known as the
assignment operator (`<-`), and then the information to save.
For example:

```{r}
message <- "It is a far, far better thing that I do, than I have ever done"
name_length <- length("Sydney Carton")
```

On the top right pane, there's a
tab that says "Environment". If you click on that, you'll see the
objects you just created listed under "Values." Objects can represent data of
various *types* - character (collection of letters and numbers encased in
quotation marks), numeric (numbers, including decimals) and more.

Referencing the name of an object by itself will print the data you assigned
to the console. You can also use objects as input to a function:

```{r}
message
class(message)
```

These examples are objects representing small, fairly simple data.
Objects can also be much more complex, representing large datasets.
Hopefully you can imagine how assigning objects can allow you to
work more easily with complex data, which you'll have a chance to do
in the next section.

## Viewing installed packages {#r-viewing-installed-packages}

Functions in R are grouped together into collections of related code called
packages. There is a tab in the lower right pane in RStudio called "Packages"
that lists all of the packages installed in your version of R.
Scrolling down the list, you'll note that some packages have the box on the
left checked. This means the package has been loaded and is available for use.

In addition to the built-in packages that come with every installation of R,
we have the ability to use packages written by other programmers.
One of the most common places to find such code is the
[Comprehensive R Archive Network](https://cran.r-project.org), or CRAN.
RStudio allows a straightforward way to download and install packages
through CRAN. If you click on the "Install" button in the Packages tab,
a window will appear allowing you to install from "Repository (CRAN)."
In the space below "Packages," type "tidyverse." Making sure the box next to
"Install dependencies" is checked, click "Install" to download and install
this collection of code (which we'll be using in later sections).

If you'd prefer to use code to install packages, rather than the RStudio pane,
you could execute the following line of code in your console:

```{r eval=FALSE}
install.packages("tidyverse")
```

This accomplishes the same task as searching and installing using the Packages pane.
We don't recommend including this code in your script, since you won't need to
install the package every time you run the script.

Once a package is installed, you'll need to ensure it is loaded and all functions
within it are recognized by R for use. You can load a package by locating it among
installed packages in the Packages pane, and then checking the box to the left of its
name. Alternatively, you can perform the same task using code:

```{r eval=FALSE}
library("tidyverse")
```

It is appropriate to include this code in your script, so any functions from loaded
packages are recognized by R and run as expected.

## Getting Help in RStudio {#r-getting-started-rstudio-help}

There are a number of additional features in RStudio that may be of
use, but we'll discuss them in later sections as the need arises.
The remaining tab that will be useful to us now is "Help," located in the lower
right pane.

You can search for more information on functions in R by typing the name
of the function, such as "round", into the search box in the Help pane.
The text that appears in the window below can sometimes be extensive.
You can search within this particular help page
using the second search box that reads "Find in Topic."
In this case, type "decimal" to find information about how the function
determines the number of decimal points to print in the output.

Luckily, R help documentation tends to be formatted very consistently. At the
very top, you'll see the name of the function ("Round", which in this case
represents a collection of related functions) followed by the name of its
package inside curly brackets ("{base}", this is important because sometimes
different packages have functions with the same names!). Below that, a short title
indicates the purpose of the function. "Description" is a more extensive
explanation that should help you figure out if the function is appropriate for
your use. "Usage" provides information on how the function should be applied
through code, and "Arguments" is like a legend for the code described in Usage.
The last subheadings, "Details" through "References" and "Examples", should be
self-explanatory.

When you're writing code in your script, you can also access help documentation
by prefacing the name of the function with a question mark.
Similar to installing packages, searching help documentation is useful
when you're working on writing code, but isn't particularly helpful to include in
your script, so try typing the following code into the console:

```{r eval=FALSE}
?sum
```

After executing that code, you should see the help page for that function appear
in the lower right pane. This search method works on functions that are currently
loaded. If you would like to perform a global search (e.g., search among all R
packages installed on your computer), you can use two question marks instead
(remember to enter this in the console):

```{r eval=FALSE}
??filter
```

RStudio also provides helpful pop-up windows that attempt to predict the
function name you're typing. You may see windows that appear as you're coding,
which include short versions of the help documentation.

## Getting Help Online {#r-getting-started-web-help}

Sometimes you may not be able to find the answers you seek in the
documentation available through RStudio.
This often happens when you need to do something
completely new and you don't know where to start. In that scenario, there are
multiple internet resources that are helpful:

* blogs
* Twitter
* tutorials
* question pages on Stack Overflow, an open forum that allows community members
to ask and answer questions

It can be challenging to find the right words to search when you're
troubleshooting something new. Don't be discouraged, and think about different
ways you can apply the terminology you're learning to make your search terms
clearer and more specific.

Even when you find information that seems useful,
these resources can be overwhelming and confusing.
Some things to look for in a Stack Overflow discussion include:

* Is the question clearly stated and is there example code? If you're
debugging, does the example code look like yours?
* Answers have upvotes and downvotes. Is there one clear answer that has a lot
of upvotes in response to a question?
* How complex is the answer? While some questions will necessarily have a
complicated answer, for many common programming tasks, there should be a
solution that only takes a few lines of code.
* Do you recognize any terms in the solution? If you don't, are there other
terms for which you could search?


> **Conventions**:
>
> One tool in using help pages (and in reading the rest of the book) will
> be understanding the conventions that writers use when describing
> many of the terms introduced in this chapter. For example
>
> ```
> min()
> ```
>
> indicates that something is a function, with parentheses to indicate
> that it is an action with (potentially) input and output.
>
> Some other conventions used in this book are:
> * `folder_name/` for a folder
> * `variable_name` for variables
> * `column_name` for columns

## Exercises

1. Run each of the functions above. Can you explain what each function
expects as input, and what kind of output it produces?

1. What would be the pros and cons of using the console versus a script in each situation?
    - Writing a data analysis with multiple steps.
    - Opening a new data set and exploring its dimensions.
    - Checking the value of a variable.

1. After googling "import csv file r", the two following pages appear in the
search results:
  https://stackoverflow.com/questions/3391880/how-to-get-a-csv-file-into-r
  http://www.sthda.com/english/wiki/reading-data-from-txt-csv-files-r-base-functions
    - What is useful about these two pages?
    - Do you understand all of the code in the second page?
    - The second page provides many possible options for code to import a
    csv file. Where would you start in attempting to perform this task?

1. Try reading a csv file:

    ```{r eval=FALSE}
    data <- read_csv("measurements.csv")
    ```

    Which pane in RStudio will show our new `data` object?
    What happens when you click on it?


## Key Points {#r-getting-started-keypoints}

```{r, child="keypoints/py-getting-started.md"}
```
