# Code Style, Review, and Refactoring {#r-rse-style}

```{r style-setup, include=FALSE}
source(here::here("_common.R"))
```

To paraphrase @Dobz1973,
nothing in software development makes sense except in light of human psychology.
This is particularly true when we look at programming style.
Computers don't need to understand programs in order to execute them,
but people do in order to create them, maintain them, and fix them.
The more clearly those programs are laid out,
the easier it is to find things and make sense of them.

The biggest benefit of having a second person work on a programming project
is therefore not getting twice as much code written,
but having code reviewed.
Study after study over more than 40 years has shown that code review is the most effective way to find bugs in software
@Faga1976,@Faga1986,@Cohe2010@Bacc2013.
Despite this,
it still isn't common in research software development,
in part because it isn't part of the culture @Sega2005,
but also because code review is mostly useful
when the reviewers understand the problem domain well enough to comment on algorithms and design choices
rather than indentation and variable naming,
and the number of people who can do that for a research project is typically very small—sometimes
as small as one @Petr2014.

This lesson will look at the mechanics of code review
and present some short examples of the kinds of things reviewers should look for.

## How should I review someone else's code? {#r-rse-style-review}

How you review is just as important as what you look for:
being dismissive or combative are good ways to ensure that people don't pay attention to your reviews,
or avoid having you review their work.
Equally,
being defensive when someone offers suggestions politely and sincerely is very human,
but can stunt your development as a programmer.

Lots of people have written guidelines for doing reviews,
which are also useful when reviewing written work @Quen2018,@Sank2018.
A few key points are:

Work in small increments.
:   As @Cohe2010 and others have found,
    code review is most effective when done in short bursts.
    That means that change requests should also be short:
    anything that's more than a couple of screens long
    should be broken into smaller pieces.

Look for algorithmic problems first.
:   Code review isn't just (or even primarily) about style:
    its real purpose is to find bugs before they can affect anyone.
    The first pass over any change should therefore look for algorithmic problems.
    Are the calculations right?
    Are any rare cases going to be missed?
    Are errors being caught and handled Chapter \@ref(logging)?

Use a rubric.
:   Linters are great,
    but can't decide when someone should have used a lookup table instead of conditionals.
    A list of things to check for can make review faster and more comprehensible,
    especially when you can copy-and-paste or drag-and-drop specific comments
    onto specific lines
    (something that GitHub unfortunately doesn't yet support).

Ask for clarification.
:   If you don't understand something,
    or don't understand why the author did it,
    ask.
    (And when the author explains it,
    think about suggesting that the explanation ought to be documented somewhere.)

Offer alternatives.
:   Telling authors that something is wrong is helpful;
    telling them what they might do instead is more so.

Don't be sarcastic or disparaging.
:   "Did you maybe think about *testing* this garbage?"
    is a Code of Conduct violation in any well-run project.

Don't present opinions as facts.
:   "Nobody uses X any more" might be true.
    If it is,
    the person making the claim ought to be able to point at download statistics
    or a Google Trends search;
    if they can't,
    they should say,
    "I don't think anybody uses X any more" and explain why they think that.

Don't feign surprise or pass judgment.
:   "Gosh, didn't you know [some obscure fact]?" isn't helpful;
    neither is, "Geez, why don't you [some clever trick] here?"

Don't overwhelm people with details.
:   If someone has used the letter `x` as a variable name in several places,
    and they shouldn't have,
    comment on the first two or three and simply put a check beside the others—the reader
    won't need the comment repeated.

Don't ask people to do extra work.
:   Nobody enjoys fixing bugs and style violations.
    Asking them to add a few features while they're at it is rude.

Don't let people break these rules just because they're frequent contributors or in positions of power.
:   The culture of any organization is shaped by the worst behavior it is willing to tolerate @Grue2015.
    If you let people be rude to one another,
    *that* is your culture.

Be specific in replies to reviewers.
:   If someone has suggested a better variable name,
    you can probably simply fix it.
    If someone has suggested a major overhaul to an algorithm,
    you should reply to their comment to point at the commit that includes the fix.

Thank your reviewers.
:   If someone has taken the time to read your code carefully,
    thank them for doing it.

## How can I check that code follows style guidelines? {#r-rse-style-check}

Checking that code conforms to guidelines like the ones above can be time consuming,
but luckily it doesn't have to be done by hand.
Most languages have tools that will check code style rules and report violations.
These are often called [linters][linter],
after an early tool called `[lint][lint]` that found lint (or fluff) in C code.

FIXME: Include text from r-reproducibility about style here.

## How should I collaborate with people when doing code reviews? {#r-rse-style-how}

FIXME: how to review on GitHub.

## In what order should functions be defined? {#r-rse-style-order}

When encountering code for the first time,
most people scan it from top to bottom.
If that code is a program or script,
rather than a library,
its main function should be put first,
and should probably be called `main`.

After reading that function,
someone should have a good idea of what the program does in what order.
Three common patterns that people might match against are:

1.  Figure out what the user has asked it to do (Chapter \@ref(configure)).
2.  Read all input data.
3.  Process it.
4.  Write output.

or:

1.  Figure out what the user has asked for.
2.  For each input file:
    1.  Read.
    2.  Process.
    3.  Write file-specific output (if any).
3.  Write summary output (if any).

or:

1.  Figure out what the user has asked for.
2.  Repeatedly:
    1.  Wait for user input.
    2.  Do what the user has asked.
3.  Exit when a "stop" command of some sort is received.

Each step in each of the outlines above usually becomes a function.
Those functions depend on others,
some of which are written to break code into comprehensible chunks and are then called just once,
others of which are utilities that may be called many times from many different places.

FIXME: figure

Different people order these differently;
our preferred order is:

1.  Put all of the single-use functions in the first half of the file
    in the order in which they are likely to be called.
2.  Put all of the multi-use utility functions in the bottom of the file in alphabetical order.

If any of those utility functions are used by other scripts or programs,
they should go in a file of their own.
In fact,
this is a good practice even if they're only used by one program,
since it signals even more clearly which functions are in the "structural" layer
and which are in the "utility" layer.

## What should I look for when reviewing code? {#r-rse-style-refactor}

The best way to answer the question in this section's title is
to describe some of the things people should change in their software.
[Refactoring][refactoring] means changing the structure of code without changing what it does,
like refactoring an equation to simplify it.
It is just as much a part of programming as writing code in the first place:
nobody gets things right the first time @Bran1995,
and needs or insights can change over time.

Most discussions of refactoring focus on ...
TODO: Complete for R
This lesson describes and motivates some of the most useful patterns;
These rules are examples of [design patterns][design-pattern]:
general solutions to commonly occurring problems in software design.
Knowing them and their names will help you create better software,
and also make it easier for you to communicate with your peers.

### Do not repeat values

Our first and simplest refactoring is called "replace value with name".
It tells us to replace magic numbers with names,
i.e., to define constants.
This can seem ridiculous in simple cases
(why define and use `inches_per_foot` instead of just writing 12?).
However,
what may be obvious to you when you're writing code won't be obvious to the next person,
particularly if they're working in a different context
(most of the world uses the metric system and doesn't know how many inches are in a foot).
It's also a matter of habit:
if you write numbers without explanation in your code for simple cases,
you're more likely to do so for complex cases,
and more likely to regret it afterward.

Using names instead of raw values also makes it easier to understand code when you read it aloud,
which is always a good test of its style.
Finally,
a single value defined in one place is much easier to change
than a bunch of numbers scattered throughout your program.
You may not think you will have to change it,
but then people want to use your software on Mars and you discover that constants aren't @Mak2006.

TODO: Include R example here.

### Do not repeat calculations in loops

It's inefficient to calculate the same value over and over again.
It also makes code less readable:
if a calculation is inside a loop or a function,
readers will assume that it might change each time the code is executed.

Our second refactoring,
"hoist repeated calculation out of loop",
tells us to move the repeated calculation out of the loop or function.
Doing this signals that its value is always the same.
And by naming that common value,
you help readers understand what its purpose is.

TODO: R example

### Replace tests with flags to clarify repeated tests

Novice programmers frequently write conditional tests like this:

```r
if ((a > b) == TRUE) {
    # ...do something...
}
```

The comparison to `TRUE` is unnecessary because `a > b` is a Boolean value
that is itself either `TRUE` or `FALSE`.
Like any other value,
Booleans can be assigned to variables,
and those variables can then be used directly in tests:

```r
was_greater = estimate > 0.0
# ...other code that might change estimate...
if (was_greater) {
    # ...do something...
}
```

This refactoring is called "replace repeated test with flag".
When it is used,
there is no need to write `if (was_greater == TRUE)`:
that always produces the same result as `if was_greater`.
Similarly, the equality tests in `if was_greater == False` is redundant:
the expression can simply be written `if not was_greater`.
Creating and using a [flag][flag-variable] instead of repeating the test
is therefore like moving a calculation out of a loop:
even if that value is only used once,
it makes our intention clearer—these really are the same test.

TODO: R example

If it takes many lines of code to process data and create a score,
and the test then needs to change from `>` to `>=`,
we're more likely to get the refactored version right the first time,
since the test only appears in one place and its result is given a name.

### Handle special cases first

A [short circuit test][short-circuit-test] is a quick check to handle a special case,
such as checking the length of a list of values
and returning `math.nan` for the average if the list is empty.
"Place short circuits early" tells us to put short-circuit tests near the start of functions
so that readers can mentally remove special cases from their thinking
while reading the code that handles the usual case.

TODO: R example

A related refactoring pattern is called "default and override".
To use it,
find cases where a value is set conditionally;
assign the default or most common value unconditionally,
and then override it in a special case.
The result is fewer lines of code and clearer control flow;
however,
it does mean executing two assignments instead of one,
so it shouldn't be used if the common case is expensive
(e.g., involves a database lookup or a web request).

TODO: R example

### Use functions to make code more comprehensible

Functions were created so that programmers could write common operations and re-use them
in order to reduce the amount of code that needed to be compiled.
Moving complex operations into functions also reduces [cognitive load][cognitive-load]
by reducing the number of things that have to be understood simultaneously.

TODO: R example

You should always extract functions when code can be used in other contests.
Even if it can't,
you should extract functions whenever it makes the function clearer
when it is read aloud.
Multi-part conditionals,
parts of long equations,
and the bodies of loops are good candidates for extraction;
if you can't think of a plausible name,
or if a lot of data has to be passed into the function after it's extracted,
the code should probably be left where it is.
Finally,
it's often helpful to keep using the original variable names as parameter names during refactoring
to reduce typing.

### Combine operations in functions

"Combine functions" is the opposite of "extract function".
If operations are always done together,
it can sometimes be be more efficient to do them together,
and *might* be easier to understand.
However,
combining functions often reduces their reusability and readability;
one sign that functions shouldn't have been combined is
how often people use the combination and throw some results away.

The fragment below shows how two functions can be combined:

TODO: R example

One thing you may not notice about the combination is that
it assumes characters are either vowels or consonants,
which means it might work differently than separate calls to the two original functions.
Issues like this are why experienced developers write unit tests (Chapter \@ref(r-rse-correct))
*before* starting to refactor.

### Replace code with data

It is sometimes easier to understand and maintain lookup tables than complicated conditionals,
so the "create lookup table" refactoring tells us to turn the latter into the former:

The more distinct cases there are,
the greater the advantage lookup tables have over multi-branch conditionals.
Those advantages multiply when items can belong to more than one category,
in which case the table is often best written as a dictionary with items as keys
and sets of categories as values:

The expressions used to update `num_vowels` and `num_consonants` make use of the fact that
`in` produces either `True` or `False`,
which the function `int` converts to either 1 or 0.
We will explore ways of making this code more readable in the exercises.

## Summary {#r-rse-style-summary}

There will always be cases where your code will be easier to understand
if you *don't* do the things described in this lesson,
but there are probably fewer of them than you think.

FIXME: create concept map for style and review.

## Exercises {#r-rse-style-exercises}

FIXME: create exercises for review.

## Key Points {#r-rse-style-keypoints}

```{r, child="keypoints/rse-style.md"}
```
