# Program Configuration {#py-rse-configuration}

Command-line options are only one level of configuration.
Depending on how complex a program is,
it may also provide:

1.  A system-wide configuration file for general settings.
2.  A user-specific configuration file for personal preferences.
3.  A job-specific file with settings for a specific run.
4.  Command-line options to change things that commonly change.

This is sometimes called [overlay configuration][overlay-configuration]
because each level overrides the ones above it:
the user's configuration file overrides the system settings,
the job configuration overrides the user's defaults,
and the command-line options overrides that.
While this is far more complexity than most research software needs (at least initially),
being able to read a complete set of options from a file is a big boost to reproducibility.

Programmers have invented many formats for configuration files,
so please do not create your own.
One possibility is to write the configuration as a Python data structure
and then load it as if it was a library.
This is clever,
but it's hard for tools in other languages to process.
A second option is the [Windows INI format][ini-format],
which is laid out like this:

```text
[section_1]
key_1=value_1
key_2=value_2

[section_2]
key_3=value_3
key_4=value_4
```

INI files are simple to read and write,
but the format is slowly falling out of use in favor of [YAML][yaml]
(Appendix \@ref(py-rse-yaml)).
Here's a sample YAML configuration file:

```yaml
# Standard settings for thesis.
logfile: "/tmp/log.txt"
quiet: false
overwrite: false
fonts:
- Verdana
- Serif
```

And here's an extension to our script template
that uses `argparse` and provide
a default configuration that the user can override:

```python
import argparse
import yaml

DEFAULT_CONFIG = {
    'source': 'DEFAULT',
    'logfile': '/tmp/log.txt',
    'quiet': False,
    'overwrite': False,
    'fonts': ['Verdana', 'Serif']
}

def main(args):
    '''Run the program.'''

    if args.configfile:
        with open(args.configfile, 'r') as reader:
            config = yaml.load(reader)
            config['source'] = args.configfile
    else:
        config = DEFAULT_CONFIG

    if args.infile:
        config['infile'] = args.infile

    if args.outfile:
        config['outfile'] = args.outfile

    print('Configuration:', config)


if __name__ == '__main__':

    description = 'Print options to the screen.'
    parser = argparse.ArgumentParser(description=description)

    parser.add_argument('--configfile', type=str, help='Configuration file name')
    parser.add_argument('--infile', type=str, help='Input file name')
    parser.add_argument('--outfile', type=str, help='Output file name')

    args = parser.parse_args()
    main(args)
```

```shell
$ python config_template.py --infile chapter2.txt --configfile thesis.yml 
```

```text
Configuration: {'logfile': '/tmp/log.txt',
                'quiet': False,
                'overwrite': False,
                'fonts': ['Verdana', 'Serif'],
                'source': 'thesis.yml',
                'infile': 'chapter2.txt'}
```

YAML allows nested keys and lists,
but if you need them,
you're probably doing something wrong [@Xu2015]:
most users never use most configuration options and find their presence confusing.

## Where should configuration files go? {#py-rse-configuration-loc}

System-wide settings for an application called `app` are often stored in `/etc/app.yml`.
(Older programs sometimes use a name like `app.rc` for their configuration file,
where `rc` stands for "resource control".)
Alternatively,
some programs will set an environment variable (Section \@ref(py-rse-bash-advanced-vars))
to the name of the installation directory,
and then read the system configuration file from there.
For example,
if the application's environment variable is `THESIS`,
we can read configuration from:

```python
install_dir = os.getenv('THESIS')
config_file = os.path.join(install_dir, 'config.yml')
if os.path.isfile(config_file):
    with open(config_file, 'r') as reader:
        config = yaml.load(reader)
```

Similarly,
we can get personal settings from `$HOME/.thesis.yml`;
the leading '.' hides the configuration file from `ls`.
Finally,
per-job settings can come from `thesis.yml` in the current directory,
where again "app" is replaced with the name of the program.

## How can I record the configuration that produced particular results? {#py-rse-configuration-log}

We don't need to write any of this when we first create a new tool.
However,
experience teaches that little tools grow into big ones.
As soon as we have more than a couple of configuration parameters,
or as soon as some of them are usually set to the same value,
we should start to think about saving our settings in files.

This is particularly important if we want others (including our future self) to be able to reproduce our work.
Careful record keeping is essential to reproducible science,
and if we add just a few lines to our program,
the computer can save the entire configuration for a particular run of a program.
For example,
if we have stored our settings in a Python dictionary called `settings` like this:

```python
settings = {
    'logfile': '/tmp/thesis.log',
    'quiet': False,
    'overwrite': True,
    'fonts': ['Comic Sans']
}
```

we can save that as YAML using this short function:

```python
def save_config(filename, settings):
    '''Save configuration settings to a file.'''

    with open(filename, 'w') as writer:
        yaml.dump(settings, writer)

save_config('thesis.yml', settings)
```

```shell
$ cat thesis.yml
```

```text
logfile: "/tmp/thesis.log"
quiet: false
overwrite: true
fonts:
- Comic Sans
```

This lets us re-create configuration on another machine
even if it has different default settings.
The test is whether our program can load a dumped configuration,
then dump it again and get the same result.

> **Version Everything**
>
> We should always include a version number as a field in the dumped configuration
> (Section \@ref(py-rse-git-advanced-tag)),
> and our program should print this when given a `--version` option.
> We need this because how we interpret options will change over time,
> and if you don't know what the version of the program was,
> we'll have to guess what options mean.


## Exercises {#py-rse-configuration-exercises}

TODO

## Key Points {#py-rse-configuration-keypoints}

```{r, child="keypoints/py-rse-configuration.md"}
```

```{r, child="./links.md"}
```
