# Learning Objectives {#py-rse-objectives}

This appendix lays out the learning objectives for each set of lessons,
and is intended to help instructors who want to use this curriculum.

<!-- Used `bin/list-keypoints-objectives.R` to create this list -->

## The Basics of the Unix Shell

```{r, child='objectives/py-rse/bash-basics.md'}
```

## Going Further with the Unix Shell

```{r, child='objectives/py-rse/bash-advanced.md'}
```

## Command Line Programs in Python

```{r, child='objectives/py-rse/scripting.md'}
```

## Git at the Command Line

```{r, child='objectives/py-rse/git-cmdline.md'}
```

## Advanced Git

```{r, child='objectives/py-rse/git-advanced.md'}
```

## Working in Teams

```{r, child='objectives/shared-rse/teams.md'}
```

## Automating Analyses

```{r, child='objectives/py-rse/automate.md'}
```

## Program Configuration

```{r, child='objectives/py-rse/config.md'}
```

## Error Handling

```{r, child='objectives/py-rse/errors.md'}
```

## Testing

```{r, child='objectives/py-rse/testing.md'}
```

## Provenance

```{r, child='objectives/shared-rse/provenance.md'}
```

## Python Packaging

```{r, child='objectives/py-rse/packaging.md'}
```
