# Learning Objectives {#py-rse-objectives}

```{r py-rse-objectives-setup, include=FALSE}
source(here::here("_common.R"))
```

This appendix lays out the learning objectives for each set of lessons,
and is intended to help instructors who want to use this curriculum.

## The Basics of the Unix Shell
```{r, child="objectives/py-rse/bash-basics.md"}
```

## Going Further with the Unix Shell
```{r, child="objectives/py-rse/bash-advanced.md"}
```

## Git at the Command Line
```{r, child="objectives/py-rse/git-cmdline.md"}
```

## Advanced Git
```{r, child="objectives/py-rse/git-advanced.md"}
```

## Code Style, Review, and Refactoring
```{r, child="objectives/py-rse/style.md"}
```

## Automating Analyses
```{r, child="objectives/py-rse/automate.md"}
```

## Working in Teams
```{r, child="objectives/shared-rse/teams.md"}
```

## Project Structure
```{r, child="objectives/py-rse/project.md"}
```

## Python Packaging
```{r, child="objectives/py-rse/packaging.md"}
```

## Correctness
```{r, child="objectives/py-rse/correct.md"}
```

## Publishing
```{r, child="objectives/shared-rse/publish.md"}
```
