# Command Line Programs in Python {#py-rse-py-scripting}

The Jupyter Notebook, PyCharm and other graphical interfaces
are great for prototyping code and exploring data,
but in many cases we ultimately need to apply our code to thousands of data files,
run it with many different parameters,
or combine it with other programs in a data analysis pipeline.
The easiest way to do this effectively is often
to turn our code into a standalone program that can be run in the Unix shell
just like other command-line tools [@Tasc2017].

In this chapter we will develop a command-line Python program
that can be controlled by several option flags,
handles input and output in ways that other command-line tools expect,
and provides useful information when things go wrong.
The result will have more scaffolding than useful application code,
but that scaffolding stays more or less the same as programs get larger.

## How can I tell if my code is a program or a module? {#py-rse-py-scripting-main}

If we are going to run a Python program from the command line,
the first thing we should do is add this to the bottom of the file:

```python
if __name__ == '__main__':
    main()
```

These two lines of code differentiate between
running a Python file as a standalone program
and importing it as a module.
When we import a Python file as a module in another program,
the `__name__` variable is automatically set to the name of the file:
for example,
if we run:

```python
import utilities
```

then for the code inside `utilities.py`,
`__name__` is `utilities`.
When we run a Python file as a standalone program,
on the other hand,
`__name__` is always set to the special string `"__main__"`.
This means that we can separate the two cases above
by checking the value of the variable `__name__`.
If that tells us that the file is running as a standalone program,
we can then handle command-line options, print help, or whatever else is appropriate.
Conventionally,
we define this functionality inside a function called `main`,
but we can call it whatever we want.
(We can also put this code directly under the `if` statement,
but that's generally considered bad practice,
since it makes testing harder.)

## How can I handle command-line options? {#py-rse-py-scripting-options}

The next thing we need is
a library to parse any options given to the program on the command line.
The most commonly used library in Python is [`argparse`][argparse],
which can handle options with or without arguments,
convert those arguments from strings to numbers or other types,
display help,
and many other things.

The simplest way to explain how `argparse` works is by example,
so let's create a short Python script called `script_template.py`:

```python
import argparse


def main(args):
    '''Run the program.'''

    print('Input file:', args.infile)
    print('Output file:', args.outfile)


if __name__ == '__main__':

    description = 'Print options to the screen.'
    parser = argparse.ArgumentParser(description=description)

    parser.add_argument('infile', type=str, help='Input file name')
    parser.add_argument('outfile', type=str, help='Output file name')

    args = parser.parse_args()
    main(args)
```

When we run `script_template.py` at the command line like this:

```shell
$ python script_template.py in.csv out.png
```

we see that `argparse` handles all the input arguments:

```text
Input file: in.csv
Output file: out.png
```

It also issues errors when users give the program invalid arguments:

```shell
$ python script_template.py in.csv
```

```text
usage: script_template.py [-h] infile outfile
script_template.py: error: the following arguments are required: outfile
```

and automatically generates help to tell us what we did wrong:

```shell
$ python script_template.py -h
```

```text
usage: script_template.py [-h] infile outfile

Print options to the screen.

positional arguments:
  infile      Input file name
  outfile     Output file name

optional arguments:
  -h, --help  show this help message and exit
```

As the `--help` hint shows,
`argparse` can handle [long options][long-option-shell] as well as traditional single-letter options;
the [online tutorial][argpapy-rse-tutorial] has examples of this and much more.

## Summary {#py-rse-py-scripting-summary}

In the novice lessons,
we learned how to reuse (rather than cut and paste)
code by defining functions.
In order to use those functions in other python notebooks/scripts,
we saw that we can save them in a file (called a module) that can be imported.
In this chapter we have seen that we can go one step further
and run our Python code outside of a Python environment,
by writing Python scripts that can be executed at the command line.


## Exercises {#py-rse-py-scripting-exercises}

TODO

## Key Points {#py-rse-py-scripting-keypoints}

```{r, child="keypoints/py-rse-scripting.md"}
```

```{r, child="./links.md"}
```
